### 2019-04-13 msnbase2.tex
initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jacsat,manuscript=article]{achemso}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{chemformula} % Formula subscripts using \ch{}
\usepackage[T1]{fontenc} % Use modern font encodings

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[EURAC]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[MSnbase v2]
  {\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{IR,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{tocentry}

Some journals require a graphical entry for the Table of Contents.
This should be laid out ``print ready'' so that the sizing of the
text is correct.

Inside the \texttt{tocentry} environment, the font used is Helvetica
8\,pt, as required by \emph{Journal of the American Chemical
Society}.

The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
permitted for  \emph{Journal of the American Chemical Society}
graphical table of content entries. The box will not resize if the
content is too big: instead it will overflow the edge of the box.

This box and the associated title will always be printed on a
separate page at the end of the document.

\end{tocentry}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  This is an example document for the \textsf{achemso} document
  class, intended for submissions to the American Chemical Society
  for publication. The class is based on the standard \LaTeXe\
  \textsf{report} file, and does not seek to reproduce the appearance
  of a published paper.

  This is an abstract for the \textsf{achemso} document class
  demonstration document.  An abstract is only allowed for certain
  manuscript types.  The selection of \texttt{journal} and
  \texttt{manuscript} will determine if an abstract is valid.  If
  not, the class will issue an appropriate error.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
This is a paragraph of text to fill the introduction of the
demonstration file.  The demonstration file attempts to show the
modifications of the standard \LaTeX\ macros that are implemented by
the \textsf{achemso} class.  These are mainly concerned with content,
as opposed to appearance.

\section{Results and discussion}

\subsection{Outline}

The document layout should follow the style of the journal concerned.
Where appropriate, sections and subsections should be added in the
normal way. If the class options are set correctly, warnings will be
given if these should not be present.

\subsection{References}

The class makes various changes to the way that references are
handled.  The class loads \textsf{natbib}, and also the
appropriate bibliography style.  References can be made using
the normal method; the citation should be placed before any
punctuation, as the class will move it if using a superscript
citation style
\cite{Mena2000,Abernethy2003,Friedman-Hill2003,EuropeanCommission2008}.
The use of \textsf{natbib} allows the use of the various citation
commands of that package: \citeauthor{Abernethy2003} have shown
something, in \citeyear{Cotton1999}, or as given by
Ref.~\citenum{Mena2000}.  Long lists of authors will be
automatically truncated in most article formats, but not in
supplementary information or reviews \cite{Pople2003}. If you
encounter problems with the citation macros, please check that
your copy of \textsf{natbib} is up to date. The demonstration
database file \texttt{achemso-demo.bib} shows how to complete
entries correctly. Notice that ``\latin{et al.}'' is auto-formatted
using the \texttt{\textbackslash latin} command.

Multiple citations to be combined into a list can be given as
a single citation.  This uses the \textsf{mciteplus} package
\cite{Johnson1972,*Arduengo1992,*Eisenstein2005,*Arduengo1994}.
Citations other than the first of the list should be indicated
with a star. If the \textsf{mciteplus} package is not installed,
the standard bibliography tools will still work but starred
references will be ignored. Individual references can be referred
to using \texttt{\textbackslash mciteSubRef}:
``ref.~\mciteSubRef{Eisenstein2005}''.

The class also handles notes to be added to the bibliography.  These
should be given in place in the document \bibnote{This is a note.
The text will be moved the the references section.  The title of the
section will change to ``Notes and References''.}.  As with
citations, the text should be placed before punctuation.  A note is
also generated if a citation has an optional note.  This assumes that
the whole work has already been cited: odd numbering will result if
this is not the case \cite[p.~1]{Cotton1999}.

\subsection{Floats}

New float types are automatically set up by the class file.  The
means graphics are included as follows (Scheme~\ref{sch:example}).  As
illustrated, the float is ``here'' if possible.
\begin{scheme}
  Your scheme graphic would go here: \texttt{.eps} format\\
  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
  these can be scaled without loss of quality, and can be\\
  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
  %\includegraphics{graphic}
  \caption{An example scheme}
  \label{sch:example}
\end{scheme}

\begin{figure}
  As well as the standard float types \texttt{table}\\
  and \texttt{figure}, the class also recognises\\
  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
  \caption{An example figure}
  \label{fgr:example}
\end{figure}

Charts, figures and schemes do not necessarily have to be labelled or
captioned.  However, tables should always have a title. It is
possible to include a number and label for a graphic without any
title, using an empty argument to the \texttt{\textbackslash caption}
macro.

The use of the different floating environments is not required, but
it is intended to make document preparation easier for authors. In
general, you should place your graphics where they make logical
sense; the production process will move them if needed.

\subsection{Math(s)}

The \textsf{achemso} class does not load any particular additional
support for mathematics.  If packages such as \textsf{amsmath} are
required, they should be loaded in the preamble.  However,
the basic \LaTeX\ math(s) input should work correctly without
this.  Some inline material \( y = mx + c \) or $ 1 + 1 = 2 $
followed by some display. \[ A = \pi r^2 \]

It is possible to label equations in the usual way (Eq.~\ref{eqn:example}).
\begin{equation}
  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
\end{equation}
This can also be used to have equations containing graphical
content. To align the equation number with the middle of the graphic,
rather than the bottom, a minipage may be used.
\begin{equation}
  \begin{minipage}[c]{0.80\linewidth}
    \centering
    As illustrated here, the width of \\
    the minipage needs to allow some  \\
    space for the number to fit in to.
    %\includegraphics{graphic}
  \end{minipage}
  \label{eqn:graphic}
\end{equation}

\section{Experimental}

The usual experimental details should appear here.  This could
include a table, which can be referenced as Table~\ref{tbl:example}.
Notice that the caption is positioned at the top of the table.
\begin{table}
  \caption{An example table}
  \label{tbl:example}
  \begin{tabular}{ll}
    \hline
    Header one  & Header two  \\
    \hline
    Entry one   & Entry two   \\
    Entry three & Entry four  \\
    Entry five  & Entry five  \\
    Entry seven & Entry eight \\
    \hline
  \end{tabular}
\end{table}

Adding notes to tables can be complicated.  Perhaps the easiest
method is to generate these using the basic
\texttt{\textbackslash textsuperscript} and
\texttt{\textbackslash emph} macros, as illustrated (Table~\ref{tbl:notes}).
\begin{table}
  \caption{A table with notes}
  \label{tbl:notes}
  \begin{tabular}{ll}
    \hline
    Header one                            & Header two \\
    \hline
    Entry one\textsuperscript{\emph{a}}   & Entry two  \\
    Entry three\textsuperscript{\emph{b}} & Entry four \\
    \hline
  \end{tabular}

  \textsuperscript{\emph{a}} Some text;
  \textsuperscript{\emph{b}} Some more text.
\end{table}

The example file also loads the optional \textsf{mhchem} package, so
that formulas are easy to input: \texttt{\textbackslash ce\{H2SO4\}}
gives \ch{H2SO4}.  See the use in the bibliography file (when using
titles in the references section).

The use of new commands should be limited to simple things which will
not interfere with the production process.  For example,
\texttt{\textbackslash mycommand} has been defined in this example,
to give italic, mono-spaced text: \mycommand{some text}.

\section{Extra information when writing JACS Communications}

When producing communications for \emph{J.~Am.\ Chem.\ Soc.}, the
class will automatically lay the text out in the style of the
journal. This gives a guide to the length of text that can be
accommodated in such a publication. There are some points to bear in
mind when preparing a JACS Communication in this way.  The layout
produced here is a \emph{model} for the published result, and the
outcome should be taken as a \emph{guide} to the final length. The
spacing and sizing of graphical content is an area where there is
some flexibility in the process.  You should not worry about the
space before and after graphics, which is set to give a guide to the
published size. This is very dependant on the final published layout.

You should be able to use the same source to produce a JACS
Communication and a normal article.  For example, this demonstration
file will work with both \texttt{type=article} and
\texttt{type=communication}. Sections and any abstract are
automatically ignored, although you will get warnings to this effect.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "Acknowledgement" section can be given in all manuscript
%% classes.  This should be given within the "acknowledgement"
%% environment, which will make the correct section or running title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgement}

Please use ``The authors thank \ldots'' rather than ``The
authors would like to thank \ldots''.

The author thanks Mats Dahlgren for version one of \textsf{achemso},
and Donald Arseneau for the code taken from \textsf{cite} to move
citations after punctuation. Many users have provided feedback on the
class, which is reflected in all of the different demonstrations
shown in this document.

\end{acknowledgement}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The same is true for Supporting Information, which should use the
%% suppinfo environment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{suppinfo}

A listing of the contents of each file supplied as Supporting Information
should be included. For instructions on what should be included in the
Supporting Information as well as how to prepare this material for
publications, refer to the journal's Instructions for Authors.

The following files are available free of charge.
\begin{itemize}
  \item Filename: brief description
  \item Filename: brief description
\end{itemize}

\end{suppinfo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The appropriate \bibliography command should be placed here.
%% Notice that the class file automatically sets \bibliographystyle
%% and also names the section correctly.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{refs}

\end{document}

### 2019-04-14 msnbase2.tex
initial plan for paper

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% This is a (brief) model paper using the achemso class
-%% The document class accepts keyval options, which should include
-%% the target journal and optionally the manuscript type.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \documentclass[journal=jacsat,manuscript=article]{achemso}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Place any additional packages needed here.  Only include packages
-%% which are essential, to avoid problems later.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage{chemformula} % Formula subscripts using \ch{}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Place any additional macros here.  Please use \newcommand* where
-%% possible, and avoid layout-changing macros (which are not used
-%% when typesetting).
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Meta-data block
-%% ---------------
-%% Each author should be given as a separate \author command.
-%%
-%% Corresponding authors should have an e-mail given after the author
-%% name as an \email command. Phone and fax numbers can be given
-%% using \phone and \fax, respectively; this information is optional.
-%%
-%% The affiliation of authors is given after the authors; each
-%% \affiliation command applies to all preceding authors not already
-%% assigned an affiliation.
-%%
-%% The affiliation takes an option argument for the short name.  This
-%% will typically be something like "University of Somewhere".
-%%
-%% The \altaffiliation macro should be used for new address, etc.
-%% On the other hand, \alsoaffiliation is used on a per author basis
-%% when authors are associated with multiple institutions.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \author{Laurent Gatto}
@@
 \author{Sebastian Gibb}
-\affiliation[Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
+\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
 \author{Johannes Rainer}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The document title should be given as usual. Some journals require
-%% a running title from the author: this should be supplied as an
-%% optional argument to \title.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[MSnbase v2]
+\title[MSnbase version 2]
   {\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Some journals require a list of abbreviations or keywords to be
-%% supplied. These should be set up here, and will be printed after
-%% the title and author information, if needed.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\abbreviations{IR,NMR,UV}
-\keywords{American Chemical Society, \LaTeX}
+\abbreviations{}
+\keywords{Bioconductor, mass spectrometry, software, reproducible research} %% up to 10 keywords
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The manuscript does not need to include \maketitle, which is
-%% executed automatically.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The "tocentry" environment can be used to create an entry for the
-%% graphical table of contents. It is given here as some journals
-%% require that it is printed as part of the abstract page. It will
-%% be automatically moved as appropriate.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{tocentry}
+%% \begin{tocentry}
+%% See achemso-demo.tex
+%% \end{tocentry}
 
-Some journals require a graphical entry for the Table of Contents.
-This should be laid out ``print ready'' so that the sizing of the
-text is correct.
-
-Inside the \texttt{tocentry} environment, the font used is Helvetica
-8\,pt, as required by \emph{Journal of the American Chemical
-Society}.
-
-The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
-permitted for  \emph{Journal of the American Chemical Society}
-graphical table of content entries. The box will not resize if the
-content is too big: instead it will overflow the edge of the box.
-
-This box and the associated title will always be printed on a
-separate page at the end of the document.
-
-\end{tocentry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The abstract environment will automatically gobble the contents
-%% if an abstract is not used by the target journal.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract}
-  This is an example document for the \textsf{achemso} document
-  class, intended for submissions to the American Chemical Society
-  for publication. The class is based on the standard \LaTeXe\
-  \textsf{report} file, and does not seek to reproduce the appearance
-  of a published paper.
-
-  This is an abstract for the \textsf{achemso} document class
-  demonstration document.  An abstract is only allowed for certain
-  manuscript types.  The selection of \texttt{journal} and
-  \texttt{manuscript} will determine if an abstract is valid.  If
-  not, the class will issue an appropriate error.
+\begin{abstract} %% 200 words max
+  We present version 2 of the \texttt{MSnbase} R/Bioconductor
+  package. \texttt{MSnbase} provides infrastructure for the
+  manipulation, processing and visualisation of mass spectrometry
+  data. Here we present how the new \textit{on disk} infrastructure
+  allows the handling of hundreds on commodity hardware and present
+  some application of the package.
 \end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Start the main part of the manuscript here.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-This is a paragraph of text to fill the introduction of the
-demonstration file.  The demonstration file attempts to show the
-modifications of the standard \LaTeX\ macros that are implemented by
-the \textsf{achemso} class.  These are mainly concerned with content,
-as opposed to appearance.
 
-\section{Results and discussion}
+What is \texttt{MSnbase}: \texttt{MSnbase} \cite{Gatto:2012} offers a
+solution between low level liraries such as jmzml, exclusively for the
+development of new applications and GUI software, that provides a
+limited and fixed set of functionality.
 
-\subsection{Outline}
+\section{Software functionality}
 
-The document layout should follow the style of the journal concerned.
-Where appropriate, sections and subsections should be added in the
-normal way. If the class options are set correctly, warnings will be
-given if these should not be present.
+\subsection{On disk backend}
 
-\subsection{References}
+Efficient low level access: in memory vs on disk mode (using mzR),
+benchmarking, used in ms-based proteomics and metabolomics.
 
-The class makes various changes to the way that references are
-handled.  The class loads \textsf{natbib}, and also the
-appropriate bibliography style.  References can be made using
-the normal method; the citation should be placed before any
-punctuation, as the class will move it if using a superscript
-citation style
-\cite{Mena2000,Abernethy2003,Friedman-Hill2003,EuropeanCommission2008}.
-The use of \textsf{natbib} allows the use of the various citation
-commands of that package: \citeauthor{Abernethy2003} have shown
-something, in \citeyear{Cotton1999}, or as given by
-Ref.~\citenum{Mena2000}.  Long lists of authors will be
-automatically truncated in most article formats, but not in
-supplementary information or reviews \cite{Pople2003}. If you
-encounter problems with the citation macros, please check that
-your copy of \textsf{natbib} is up to date. The demonstration
-database file \texttt{achemso-demo.bib} shows how to complete
-entries correctly. Notice that ``\latin{et al.}'' is auto-formatted
-using the \texttt{\textbackslash latin} command.
+\subsection{Use cases}
 
-Multiple citations to be combined into a list can be given as
-a single citation.  This uses the \textsf{mciteplus} package
-\cite{Johnson1972,*Arduengo1992,*Eisenstein2005,*Arduengo1994}.
-Citations other than the first of the list should be indicated
-with a star. If the \textsf{mciteplus} package is not installed,
-the standard bibliography tools will still work but starred
-references will be ignored. Individual references can be referred
-to using \texttt{\textbackslash mciteSubRef}:
-``ref.~\mciteSubRef{Eisenstein2005}''.
+Example applications: visualisation, ...
 
-The class also handles notes to be added to the bibliography.  These
-should be given in place in the document \bibnote{This is a note.
-The text will be moved the the references section.  The title of the
-section will change to ``Notes and References''.}.  As with
-citations, the text should be placed before punctuation.  A note is
-also generated if a citation has an optional note.  This assumes that
-the whole work has already been cited: odd numbering will result if
-this is not the case \cite[p.~1]{Cotton1999}.
+\section{Discussion}
 
-\subsection{Floats}
+To address (from guidelines):
 
-New float types are automatically set up by the class file.  The
-means graphics are included as follows (Scheme~\ref{sch:example}).  As
-illustrated, the float is ``here'' if possible.
-\begin{scheme}
-  Your scheme graphic would go here: \texttt{.eps} format\\
-  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
-  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
-  these can be scaled without loss of quality, and can be\\
-  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
-  %\includegraphics{graphic}
-  \caption{An example scheme}
-  \label{sch:example}
-\end{scheme}
+\begin{itemize}
 
-\begin{figure}
-  As well as the standard float types \texttt{table}\\
-  and \texttt{figure}, the class also recognises\\
-  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
-  \caption{An example figure}
-  \label{fgr:example}
-\end{figure}
+\item{potential for reuse}
+\item{general limitations}
+\item{system limitations}
+\item{end-user documentation}
+\item{developer documentation}
+\item{sample data}
+\item{benchmark data set}
+\item{availability}
+\item{license information}
+\item{system requirements}
 
-Charts, figures and schemes do not necessarily have to be labelled or
-captioned.  However, tables should always have a title. It is
-possible to include a number and label for a graphic without any
-title, using an empty argument to the \texttt{\textbackslash caption}
-macro.
+\end{itemize}
 
-The use of the different floating environments is not required, but
-it is intended to make document preparation easier for authors. In
-general, you should place your graphics where they make logical
-sense; the production process will move them if needed.
+Collaborative development, 11 contributors since creation (see blog
+post).
 
-\subsection{Math(s)}
+Count packages depending on MSnbase.
 
-The \textsf{achemso} class does not load any particular additional
-support for mathematics.  If packages such as \textsf{amsmath} are
-required, they should be loaded in the preamble.  However,
-the basic \LaTeX\ math(s) input should work correctly without
-this.  Some inline material \( y = mx + c \) or $ 1 + 1 = 2 $
-followed by some display. \[ A = \pi r^2 \]
+Future developments.
 
-It is possible to label equations in the usual way (Eq.~\ref{eqn:example}).
-\begin{equation}
-  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
-\end{equation}
-This can also be used to have equations containing graphical
-content. To align the equation number with the middle of the graphic,
-rather than the bottom, a minipage may be used.
-\begin{equation}
-  \begin{minipage}[c]{0.80\linewidth}
-    \centering
-    As illustrated here, the width of \\
-    the minipage needs to allow some  \\
-    space for the number to fit in to.
-    %\includegraphics{graphic}
-  \end{minipage}
-  \label{eqn:graphic}
-\end{equation}
-
-\section{Experimental}
-
-The usual experimental details should appear here.  This could
-include a table, which can be referenced as Table~\ref{tbl:example}.
-Notice that the caption is positioned at the top of the table.
-\begin{table}
-  \caption{An example table}
-  \label{tbl:example}
-  \begin{tabular}{ll}
-    \hline
-    Header one  & Header two  \\
-    \hline
-    Entry one   & Entry two   \\
-    Entry three & Entry four  \\
-    Entry five  & Entry five  \\
-    Entry seven & Entry eight \\
-    \hline
-  \end{tabular}
-\end{table}
-
-Adding notes to tables can be complicated.  Perhaps the easiest
-method is to generate these using the basic
-\texttt{\textbackslash textsuperscript} and
-\texttt{\textbackslash emph} macros, as illustrated (Table~\ref{tbl:notes}).
-\begin{table}
-  \caption{A table with notes}
-  \label{tbl:notes}
-  \begin{tabular}{ll}
-    \hline
-    Header one                            & Header two \\
-    \hline
-    Entry one\textsuperscript{\emph{a}}   & Entry two  \\
-    Entry three\textsuperscript{\emph{b}} & Entry four \\
-    \hline
-  \end{tabular}
-
-  \textsuperscript{\emph{a}} Some text;
-  \textsuperscript{\emph{b}} Some more text.
-\end{table}
-
-The example file also loads the optional \textsf{mhchem} package, so
-that formulas are easy to input: \texttt{\textbackslash ce\{H2SO4\}}
-gives \ch{H2SO4}.  See the use in the bibliography file (when using
-titles in the references section).
-
-The use of new commands should be limited to simple things which will
-not interfere with the production process.  For example,
-\texttt{\textbackslash mycommand} has been defined in this example,
-to give italic, mono-spaced text: \mycommand{some text}.
-
-\section{Extra information when writing JACS Communications}
-
-When producing communications for \emph{J.~Am.\ Chem.\ Soc.}, the
-class will automatically lay the text out in the style of the
-journal. This gives a guide to the length of text that can be
-accommodated in such a publication. There are some points to bear in
-mind when preparing a JACS Communication in this way.  The layout
-produced here is a \emph{model} for the published result, and the
-outcome should be taken as a \emph{guide} to the final length. The
-spacing and sizing of graphical content is an area where there is
-some flexibility in the process.  You should not worry about the
-space before and after graphics, which is set to give a guide to the
-published size. This is very dependant on the final published layout.
-
-You should be able to use the same source to produce a JACS
-Communication and a normal article.  For example, this demonstration
-file will work with both \texttt{type=article} and
-\texttt{type=communication}. Sections and any abstract are
-automatically ignored, although you will get warnings to this effect.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The "Acknowledgement" section can be given in all manuscript
-%% classes.  This should be given within the "acknowledgement"
-%% environment, which will make the correct section or running title.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{acknowledgement}
 
-Please use ``The authors thank \ldots'' rather than ``The
-authors would like to thank \ldots''.
-
-The author thanks Mats Dahlgren for version one of \textsf{achemso},
-and Donald Arseneau for the code taken from \textsf{cite} to move
-citations after punctuation. Many users have provided feedback on the
-class, which is reflected in all of the different demonstrations
-shown in this document.
+The authors thank the various contributors and users who have provided
+constructuve input and feedback that have helped, over the years, the
+improvement of the package. The authors declare no conflict of
+interest.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The same is true for Supporting Information, which should use the
-%% suppinfo environment.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{suppinfo}
+%% \begin{suppinfo} %% Supporting Information
+%%
+%% A listing of the contents of each file supplied as Supporting Information
+%% should be included. For instructions on what should be included in the
+%% Supporting Information as well as how to prepare this material for
+%% publications, refer to the journal's Instructions for Authors.
+%%
+%% The following files are available free of charge.
+%% \begin{itemize}
+%%   \item Filename: brief description
+%%   \item Filename: brief description
+%% \end{itemize}
+%% \end{suppinfo}
 
-A listing of the contents of each file supplied as Supporting Information
-should be included. For instructions on what should be included in the
-Supporting Information as well as how to prepare this material for
-publications, refer to the journal's Instructions for Authors.
-
-The following files are available free of charge.
-\begin{itemize}
-  \item Filename: brief description
-  \item Filename: brief description
-\end{itemize}
-
-\end{suppinfo}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The appropriate \bibliography command should be placed here.
-%% Notice that the class file automatically sets \bibliographystyle
-%% and also names the section correctly.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{refs}

### 2019-04-15 msnbase2.tex
potential for re-use examples

@@
-\documentclass[journal=jacsat,manuscript=article]{achemso}
+\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \keywords{Bioconductor, mass spectrometry, software, reproducible research} %% up to 10 keywords
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-\item{potential for reuse}
+\item{potential for reuse}: see
+  \cite{Wieczorek:2017,Griss:2019,Smith:2006} for examples.
 \item{general limitations}
@@
 
-Count packages depending on MSnbase.
+Count packages depending on \texttt{MSnbase}.
 
 Future developments.
+
+The version of \texttt{MSnbase} used in this manuscritp is version
+2.9.4. The main features presented here
+were available since version 2.0.
 

### 2019-05-31 msnbase2.Rnw
use Rnw file and Makefile
start intro

--- 2019-04-14 msnbase2.tex
@@
 
+Mass spectrometry is a powerful technology to assays chemical and
+biological samples. It is used routinely, with well characterised
+protocol, as well a development platform, to improve on existing
+methods and devise new ones to analyse ever more complex sample in
+greater details. The complexity and diversity of mass spectrometry
+yields data that is itself complex and often times of considerable
+size, that requires non trivial processing before producing
+interpretable results. This is particularly relevant, and can
+constitue a significant challenge for method developers that, in
+addition to the development of sample processing and mass spectrometry
+methods, need to process and analyse these new data to demonstrate the
+improvement in their technical and analytical work.
+
+There exists a very diverse catalogue of software tools to explore,
+process and interpret mass spectrometry data. These range from low
+level software libraries such as vendor libraries, jmzML (ref),
+proteowizard (ref), ... that are aimed at programmers to develop new
+applications, to user-oriented applications,  such as ProteomeDiscoverer, MaxQuant, ...
+
+The former are used through application programming interfaces
+exclusively, while the latter generally featuring graphical user
+interfaces (GUI).
+
+Give examples of libraries re-used in user/gui focused application
+
+In this software note, we are going to focus on metabolomics and
+proteomics applications of mass spectrometry.
+
 What is \texttt{MSnbase}: \texttt{MSnbase} \cite{Gatto:2012} offers a
@@
 
-Count packages depending on MSnbase.
+Count packages depending on \texttt{MSnbase}.
 
 Future developments.
+
+The version of \texttt{MSnbase} used in this manuscritp is version
+\Sexpr{packageVersion("MSnbase")}. The main features presented here
+were available since version 2.0.
 

### 2019-06-03 msnbase2.Rnw
preliminary core of intro, bullet points rest

@@
 proteowizard (ref), ... that are aimed at programmers to develop new
-applications, to user-oriented applications,  such as ProteomeDiscoverer, MaxQuant, ...
+applications, to user-oriented applications, such as
+ProteomeDiscoverer, MaxQuant, ... that provide a limited and fixed set
+of functionality. The former are used through application programming
+interfaces exclusively, while the latter generally featuring graphical
+user interfaces (GUI).
 
-The former are used through application programming interfaces
-exclusively, while the latter generally featuring graphical user
-interfaces (GUI).
+TODO: Give examples of libraries re-used in user/gui focused
+application...
 
-Give examples of libraries re-used in user/gui focused application
-
-In this software note, we are going to focus on metabolomics and
-proteomics applications of mass spectrometry.
-
-What is \texttt{MSnbase}: \texttt{MSnbase} \cite{Gatto:2012} offers a
-solution between low level liraries such as jmzml, exclusively for the
-development of new applications and GUI software, that provides a
-limited and fixed set of functionality.
+In this software note, we present version 2 of the \texttt{MSnbase}
+\cite{Gatto:2012} R/Bioconductor software package. \texttt{MSnbase}
+offers a platform that lies between low level libraries and end-use
+software. It provides a flexible command line environment for
+metabolomics and proteomics mass spectrometry-based application, that
+allows a detailed step-by-step processing, analysis and exploration of
+the data and development of novel computational mass spectrometry
+methods.
 
@@
 
-Example applications: visualisation, ...
+Example applications:
+
+- visualisation
+- large scale data analyses (metabolomics, Johannes)
+- boxcar prototyping
 

### 2019-06-03 msnbase2.tex
preliminary core of intro, bullet points rest

@@
 
-What is \texttt{MSnbase}: \texttt{MSnbase} \cite{Gatto:2012} offers a
-solution between low level liraries such as jmzml, exclusively for the
-development of new applications and GUI software, that provides a
-limited and fixed set of functionality.
+Mass spectrometry is a powerful technology to assays chemical and
+biological samples. It is used routinely, with well characterised
+protocol, as well a development platform, to improve on existing
+methods and devise new ones to analyse ever more complex sample in
+greater details. The complexity and diversity of mass spectrometry
+yields data that is itself complex and often times of considerable
+size, that requires non trivial processing before producing
+interpretable results. This is particularly relevant, and can
+constitue a significant challenge for method developers that, in
+addition to the development of sample processing and mass spectrometry
+methods, need to process and analyse these new data to demonstrate the
+improvement in their technical and analytical work.
+
+There exists a very diverse catalogue of software tools to explore,
+process and interpret mass spectrometry data. These range from low
+level software libraries such as vendor libraries, jmzML (ref),
+proteowizard (ref), ... that are aimed at programmers to develop new
+applications, to user-oriented applications, such as
+ProteomeDiscoverer, MaxQuant, ... that provide a limited and fixed set
+of functionality. The former are used through application programming
+interfaces exclusively, while the latter generally featuring graphical
+user interfaces (GUI).
+
+TODO: Give examples of libraries re-used in user/gui focused
+application...
+
+In this software note, we present version 2 of the \texttt{MSnbase}
+\cite{Gatto:2012} R/Bioconductor software package. \texttt{MSnbase}
+offers a platform that lies between low level libraries and end-use
+software. It provides a flexible command line environment for
+metabolomics and proteomics mass spectrometry-based application, that
+allows a detailed step-by-step processing, analysis and exploration of
+the data and development of novel computational mass spectrometry
+methods.
 
@@
 The version of \texttt{MSnbase} used in this manuscritp is version
-2.9.4. The main features presented here
+2.10.0. The main features presented here
 were available since version 2.0.

### 2019-06-05 msnbase2.Rnw
intro to main section

@@
 
+We are going to present the new raw data infrastructure that has been
+available since \texttt{MSnbase} version 2.0, and then describe how it
+is leveraged to process large amount of data. We are then going to
+demonstrate some visualisation that make use of the raw data
+structures available in \texttt{MSnbase} and conlude with an example
+if flexible prototyping, by showing how to quickly combine spectra
+acquired using the Boxcar method (REF).
+
 \subsection{On disk backend}
@@
 
-- visualisation
 - large scale data analyses (metabolomics, Johannes)
+
+- visualisation: Jo's XIC plots, possibly also MSMap, with animation.
+
 - boxcar prototyping

### 2020-04-23 msnbase2.Rnw
add benchmarking scipts and results

@@
 
-Mass spectrometry is a powerful technology to assays chemical and
+Mass spectrometry is a powerful technology to assay chemical and
 biological samples. It is used routinely, with well characterised
-protocol, as well a development platform, to improve on existing
+protocols, as well a development platform, to improve on existing
 methods and devise new ones to analyse ever more complex sample in
@@
 yields data that is itself complex and often times of considerable
-size, that requires non trivial processing before producing
-interpretable results. This is particularly relevant, and can
-constitue a significant challenge for method developers that, in
-addition to the development of sample processing and mass spectrometry
-methods, need to process and analyse these new data to demonstrate the
-improvement in their technical and analytical work.
+size, requiring non trivial processing before producing interpretable
+results. This is particularly relevant, and can constitue a
+significant challenge for method developers that, in addition to the
+development of sample processing and mass spectrometry methods, need
+to process and analyse these new data to demonstrate the improvement
+in their technical and analytical work.
 
@@
 offers a platform that lies between low level libraries and end-use
-software. It provides a flexible command line environment for
-metabolomics and proteomics mass spectrometry-based application, that
-allows a detailed step-by-step processing, analysis and exploration of
-the data and development of novel computational mass spectrometry
-methods.
+software. It provides a flexible and interactive command line
+environment for metabolomics and proteomics mass spectrometry-based
+application. It allows a detailed step-by-step processing, analysis
+and exploration of the data and development of novel computational
+mass spectrometry methods.
 
@@
 if flexible prototyping, by showing how to quickly combine spectra
-acquired using the Boxcar method (REF).
+acquired using the Boxcar method \cite{Meier:2018}.
 

### 2020-04-23 S01-ondisk.R

@@ -1,0 +1,205 @@
+library("MSnbase")
+
+prep_data <- function(n,
+                      f = msdata::proteomics(full.names = TRUE, pattern = "20141210"),
+                      data_dir = "./data",
+                      clean = TRUE) {
+    if (clean && dir.exists(data_dir))
+        unlink(data_dir, recursive = TRUE)
+    if (!dir.exists(data_dir))
+        dir.create(data_dir)
+    fls <- paste0(data_dir, "/f", seq_len(n), ".mzML")
+    stopifnot(all(file.link(f, fls)))
+    fls
+}
+
+fls <- prep_data(100)
+
+## ----------------------------------------------------------------------
+## BENCHMARK: Reading time (in seconds) for a 1 file (default
+## settings, i.e. 6103 MS2 spectra for in memory and 1431 MS1 + 6103
+## MS2 spectra for on disk)
+## ----------------------------------------------------------------------
+
+t_read1 <- c(on_disk = system.time(x_dsk <- readMSData(fls[1], mode = "onDisk"))[["elapsed"]],
+             in_mem = system.time(x_mem <- readMSData(fls[1], mode = "inMemory"))[["elapsed"]])
+
+save(t_read1, file = "bench_t_read1.rda")
+
+## > t_read1
+## on_disk  in_mem 
+##  15.304  78.353
+
+
+## ----------------------------------------------------------------------
+## BENCHMARK: Reading time (in seconds) for a 1, 2, 5, 10, ... files
+## (default settings, i.e. 6103 MS2 spectra for in memory and 1431 MS1
+## + 6103 MS2 spectra for on disk)
+## ----------------------------------------------------------------------
+n <- c(1, 5, 10)
+t_read <- sapply(n, function(i) {
+    f <- fls[seq_len(i)]
+    c(on_disk = system.time(readMSData(f, mode = "onDisk"))[["elapsed"]],
+      in_mem = system.time(readMSData(f, mode = "inMemory"))[["elapsed"]])
+})
+
+save(t_read, file = "bench_t_read.rda")
+
+## > t_read
+##           [,1]    [,2]    [,3]
+## on_disk 14.902  74.757 151.029
+## in_mem  68.931 349.450 724.357
+
+
+## ----------------------------------------------------------------------
+## BENCHMARK: Memory consumption, compareing on disk and in memory for
+## 1, 5, 10, ... files
[…]

### 2020-04-24 msnbase2.Rnw
proceed with on-disk benchmarking section

--- 2019-06-03 msnbase2.Rnw
@@
-\documentclass[journal=jacsat,manuscript=article]{achemso}
+\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+%% \let\hlipl\hlkwb
+
+%% \usepackage{framed}
+%% \makeatletter
+%% \newenvironment{kframe}{%
+%%  \def\at@end@of@kframe{}%
+%%  \ifinner\ifhmode%
+%%   \def\at@end@of@kframe{\end{minipage}}%
+%%   \begin{minipage}{\columnwidth}%
+%%  \fi\fi%
+%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
+%%      % There is no \\@totalrightmargin, so:
+%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+%%  \MakeFramed {\advance\hsize-\width
+%%    \@totalleftmargin\z@ \linewidth\hsize
+%%    \@setminipage}}%
+%%  {\par\unskip\endMakeFramed%
+%%  \at@end@of@kframe}
+%% \makeatother
+
+%% \usepackage{xcolor}
+
+%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
+%% \definecolor{messagecolor}{rgb}{0, 0, 0}
+%% \definecolor{warningcolor}{rgb}{1, 0, 1}
+%% \definecolor{errorcolor}{rgb}{1, 0, 0}
+%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{subfig}
+\usepackage{graphicx}
+\usepackage{alltt}
 
@@
 \email{laurent.gatto@uclouvain.be}
-\affiliation[UCLouvain]{de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
+\affiliation[UCLouvain]{Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
 \author{Sebastian Gibb}
@@
   manipulation, processing and visualisation of mass spectrometry
-  data. Here we present how the new \textit{on disk} infrastructure
+  data. Here we present how the new \textit{on-disk} infrastructure
   allows the handling of hundreds on commodity hardware and present
@@
 \end{abstract}
+
+<<setup, include = FALSE, message = FALSE>>=
+knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
+library("xtable")
+library("microbenchmark")
+library("tidyverse")
+@
 
@@
 
-\subsection{On disk backend}
+\subsection{On-disk backend}
 
-Efficient low level access: in memory vs on disk mode (using mzR),
-benchmarking, used in ms-based proteomics and metabolomics.
+The main feature in version 2 of the \texttt{MSnbase} package was the
+addition of different backends for raw data storage, namely
+\textit{in-memory} and on-disk. The following code chunk
+demonstrates how to create two \texttt{MSnExp} objects, tailored to
+manage mass spectrometry experiments, storing data in-memory or
+on-disk.
 
-\subsection{Use cases}
+<<readMSData, eval = FALSE, echo = TRUE>>=
+library("MSnbase")
+raw_mem <- readMSData("file.mzML", mode = "inMemory")
+raw_dsk <- readMSData("file.mzML", mode = "onDisk")
+@
 
-Example applications:
+The former is the legacy storage mode, implemented in the first
+version of the package, that loads all the raw data and the metadata
+in memory. This solution doesn't scale for modern large dataset, and
+was complemented by the on-disk backend, that only load
+metadata in memory and accesses the spectra in the original files when
+needed. There are two direct benefits using the on-disk
+backend. Table \ref{tab:bench1} shows 4-fold faster reading times
+(left) and 10-fold smaller memory footprint (right).
 
-- visualisation
-- large scale data analyses (metabolomics, Johannes)
-- boxcar prototyping
+<<time_sz>>=
+load("bench_time_sz.rda")
+time <- sapply(time_sz, function(x) x[, "time"])
+sz <- sapply(time_sz, function(x) x[, "sz"])
+colnames(sz) <- colnames(time) <- c(1, 5, 10)
+rownames(sz) <- rownames(time) <- c("in-memory", "on-disk")
+sz <- round(sz/(1014^2), 2)
+print(xtable(time), floating = FALSE, file = "tabs/t_read.tex")
+print(xtable(sz), floating = FALSE, include.rownames = FALSE, file = "tabs/sz.tex")
+@
+
+\begin{table}[ht]
+\centering
+\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
+\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
+\caption{Reading time (in seconds) (left) and data size in memory (in
+  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
+  only) and 6103 MS2 (on-disk and in-memory) spectra. }
+\label{tab:bench1}
+\end{table}
+
+The on-disk backend also offers efficient data manipulation by way of
+\textit{lazy processing}. Operations on the raw data are stored in a
+processing queue and only effectively applied when raw data is
+accessed on disk. As an example, the following short analysis
+pipeline, that can equally be applied to on in-memory or on-disk data
+retains MS2 spectra acquired between 1000 and 3000 seconds, extract
+the M/Z range corresponding to the TMT 6-plex range and focuses on the
+MS2 spectra with a precursor intensity greater than $11 \times 10^6$
+(the median precursor intensity).
+
+<<filter, eval = FALSE, echo = TRUE>>=
+x <- x_dsk %>%
+    filterRt(c(1000, 3000)) %>%
+    filterMz(120, 135)
+x[precursorIntensity(x) > 11e6, ]
+@
+
+As shown on figure \ref{fig:filt}, this lazy mechanism si
+significantly faster than its application on in-memory data.
+
+\begin{figure}[ht]
+  \centering
+<<plot_filter, message = FALSE, out.width = '50%', fig.width = 5, fig.height = 3>>=
+load("bench_t_filt.rda")
+ggplot2::autoplot(t_filt)
+@
+\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
+\label{fig:filt}
+\end{figure}
+
+The advantageous reading and execution times and memory footprint of
+the on-disk backend are possible by avoiding unnecessary access to the
+raw data. Once access to the spectra M/Z and intensity values become
+mandatory (for example for plotting), then the in-memory backend
+becomes more efficient, as illustrated on figure
+\ref{fig:access}. This gain is maximal when the whole dataset is the
+be accessed (i.e. all spectra are already in memory) and negligeable
+when large fractions of the data need to be subset.
+
+\begin{figure}[ht]
+  \centering
+<<plot_access, message = FALSE, out.width = '75%', fig.width = 5, fig.height = 3>>=
+load("bench_t_access.rda")
+tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
+       expr = as.character(t_access$expr)) %>%
+    mutate(n = sub("^access_.+_", "", expr)) %>%
+    mutate(n = sub("all", 6103, n)) %>%
+    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
+    ggplot(aes(x = n, y = time, fill = log10(as.numeric(n)))) + 
+    ggplot2::geom_violin() + 
+    ggplot2::scale_y_log10() +
+    ylab("Time [millisec]") +
+    xlab("Number of spectra (out of 6103)") +
+    facet_wrap(~ mode) +
+    theme(legend.position = "none")
+@
+\caption{Access time to spectra for the in-memory (left) and on-disk
+  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
+\label{fig:access}
+\end{figure}
+
+Enables large scale data analyses (metabolomics example, see Johannes)
+
+\subsection{Prototyping}
+
+See https://github.com/lgatto/msnbase\_boxcar.
+
+\subsection{Visualisation}
+
+Examples: 3D MSmap, boxcar, centroiding vignette
 
@@
 The version of \texttt{MSnbase} used in this manuscritp is version
-\Sexpr{packageVersion("MSnbase")}. The main features presented here
+2.10.0. The main features presented here
 were available since version 2.0.

### 2020-04-24 msnbase2.tex
add benchmarking scipts and results
proceed with on-disk benchmarking section

@@
 \documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
 \makeatletter
@@
 
+\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+%% \let\hlipl\hlkwb
+
+%% \usepackage{framed}
+%% \makeatletter
+%% \newenvironment{kframe}{%
+%%  \def\at@end@of@kframe{}%
+%%  \ifinner\ifhmode%
+%%   \def\at@end@of@kframe{\end{minipage}}%
+%%   \begin{minipage}{\columnwidth}%
+%%  \fi\fi%
+%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
+%%      % There is no \\@totalrightmargin, so:
+%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+%%  \MakeFramed {\advance\hsize-\width
+%%    \@totalleftmargin\z@ \linewidth\hsize
+%%    \@setminipage}}%
+%%  {\par\unskip\endMakeFramed%
+%%  \at@end@of@kframe}
+%% \makeatother
+
+%% \usepackage{xcolor}
+
+%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
+%% \definecolor{messagecolor}{rgb}{0, 0, 0}
+%% \definecolor{warningcolor}{rgb}{1, 0, 1}
+%% \definecolor{errorcolor}{rgb}{1, 0, 0}
+%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{subfig}
+\usepackage{graphicx}
 \usepackage{alltt}
@@
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-\subsection{On disk backend}
+\subsection{On-disk backend}
 
-Efficient low level access: in memory vs on disk mode (using mzR),
-benchmarking, used in ms-based proteomics and metabolomics.
+The main feature in version 2 of the \texttt{MSnbase} package was the
+addition of different backends for raw data storage, namely
+\textit{in-memory} and on-disk. The following code chunk
+demonstrates how to create two \texttt{MSnExp} objects, tailored to
+manage mass spectrometry experiments, storing data in-memory or
+on-disk.
 
-\subsection{Use cases}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(}\hlstr{"MSnbase"}\hlstd{)}
+\hlstd{raw_mem} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"file.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"inMemory"}\hlstd{)}
+\hlstd{raw_dsk} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"file.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"onDisk"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
-Example applications:
+The former is the legacy storage mode, implemented in the first
+version of the package, that loads all the raw data and the metadata
+in memory. This solution doesn't scale for modern large dataset, and
+was complemented by the on-disk backend, that only load
+metadata in memory and accesses the spectra in the original files when
+needed. There are two direct benefits using the on-disk
+backend. Table \ref{tab:bench1} shows 4-fold faster reading times
+(left) and 10-fold smaller memory footprint (right).
 
-- visualisation
-- large scale data analyses (metabolomics, Johannes)
-- boxcar prototyping
+\begin{table}[ht]
+\centering
+\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
+\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
+\caption{Reading time (in seconds) (left) and data size in memory (in
+  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
+  only) and 6103 MS2 (on-disk and in-memory) spectra. }
+\label{tab:bench1}
+\end{table}
+
+The on-disk backend also offers efficient data manipulation by way of
+\textit{lazy processing}. Operations on the raw data are stored in a
+processing queue and only effectively applied when raw data is
+accessed on disk. As an example, the following short analysis
+pipeline, that can equally be applied to on in-memory or on-disk data
+retains MS2 spectra acquired between 1000 and 3000 seconds, extract
+the M/Z range corresponding to the TMT 6-plex range and focuses on the
+MS2 spectra with a precursor intensity greater than $11 \times 10^6$
+(the median precursor intensity).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{x} \hlkwb{<-} \hlstd{x_dsk} \hlopt{%>%}
+    \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1000}\hlstd{,} \hlnum{3000}\hlstd{))} \hlopt{%>%}
+    \hlkwd{filterMz}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{135}\hlstd{)}
+\hlstd{x[}\hlkwd{precursorIntensity}\hlstd{(x)} \hlopt{>} \hlnum{11e6}\hlstd{, ]}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+As shown on figure \ref{fig:filt}, this lazy mechanism si
+significantly faster than its application on in-memory data.
+
+\begin{figure}[ht]
+  \centering
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=0.5\linewidth]{figure/plot_filter-1}
+
+\end{knitrout}
+\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
+\label{fig:filt}
+\end{figure}
+
+The advantageous reading and execution times and memory footprint of
+the on-disk backend are possible by avoiding unnecessary access to the
+raw data. Once access to the spectra M/Z and intensity values become
+mandatory (for example for plotting), then the in-memory backend
+becomes more efficient, as illustrated on figure
+\ref{fig:access}. This gain is maximal when the whole dataset is the
+be accessed (i.e. all spectra are already in memory) and negligeable
+when large fractions of the data need to be subset.
+
+\begin{figure}[ht]
+  \centering
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=0.75\linewidth]{figure/plot_access-1}
+
+\end{knitrout}
+\caption{Access time to spectra for the in-memory (left) and on-disk
+  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
+\label{fig:access}
+\end{figure}
+
+Enables large scale data analyses (metabolomics example, see Johannes)
+
+\subsection{Prototyping}
+
+See https://github.com/lgatto/msnbase\_boxcar.
+
+\subsection{Visualisation}
+
+Examples: 3D MSmap, boxcar, centroiding vignette
 

### 2020-04-24 S01-ondisk.R

@@ -15,60 +15,26 @@ prep_data <- function(n,
 
 fls <- prep_data(100)
 
-## ----------------------------------------------------------------------
-## BENCHMARK: Reading time (in seconds) for a 1 file (default
-## settings, i.e. 6103 MS2 spectra for in memory and 1431 MS1 + 6103
-## MS2 spectra for on disk)
-## ----------------------------------------------------------------------
-
-t_read1 <- c(on_disk = system.time(x_dsk <- readMSData(fls[1], mode = "onDisk"))[["elapsed"]],
-             in_mem = system.time(x_mem <- readMSData(fls[1], mode = "inMemory"))[["elapsed"]])
-
-save(t_read1, file = "bench_t_read1.rda")
-
-## > t_read1
-## on_disk  in_mem 
-##  15.304  78.353
-
 
 ## ----------------------------------------------------------------------
-## BENCHMARK: Reading time (in seconds) for a 1, 2, 5, 10, ... files
-## (default settings, i.e. 6103 MS2 spectra for in memory and 1431 MS1
-## + 6103 MS2 spectra for on disk)
+## BENCHMARK:
+## - Reading time (in seconds) for a 1, 5, 10, ... files (default
+##   settings, i.e. 6103 MS2 spectra for in memory and 1431 MS1 + 6103
+##   MS2 spectra for on disk)
+## - Memory consumption, compareing on disk and in memory for 1, 5,
+##   10, ... files
 ## ----------------------------------------------------------------------
 n <- c(1, 5, 10)
-t_read <- sapply(n, function(i) {
+time_sz <- lapply(n, function(i) {
     f <- fls[seq_len(i)]
-    c(on_disk = system.time(readMSData(f, mode = "onDisk"))[["elapsed"]],
-      in_mem = system.time(readMSData(f, mode = "inMemory"))[["elapsed"]])
+    time <- c(in_mem = system.time(x_mem <- readMSData(f, mode = "inMemory"))[["elapsed"]],
+              on_disk = system.time(x_dsk <- readMSData(f, mode = "onDisk"))[["elapsed"]])
+    sz <- c(in_mem = pryr::object_size(x_mem),
+            on_disk = pryr::object_size(x_dsk))
+    cbind(time, sz, n = i)
 })
 
-save(t_read, file = "bench_t_read.rda")
-
-## > t_read
-##           [,1]    [,2]    [,3]
-## on_disk 14.902  74.757 151.029
-## in_mem  68.931 349.450 724.357
-
-
[…]

### 2020-04-25 msnbase2.Rnw
replace bench1 tab with figure

@@
 library("tidyverse")
+library("patchwork")
 @
@@
 in memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only load
-metadata in memory and accesses the spectra in the original files when
+was complemented by the on-disk backend, that only load metadata in
+memory and accesses the spectra in the original files when
 needed. There are two direct benefits using the on-disk
-backend. Table \ref{tab:bench1} shows 4-fold faster reading times
+backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
 (left) and 10-fold smaller memory footprint (right).
@@
 sz <- round(sz/(1014^2), 2)
-print(xtable(time), floating = FALSE, file = "tabs/t_read.tex")
-print(xtable(sz), floating = FALSE, include.rownames = FALSE, file = "tabs/sz.tex")
+
+p1 <- data.frame(time) %>% 
+    rownames_to_column() %>% 
+    pivot_longer(names_to = "files", 
+                 values_to = "time", 
+                 -rowname) %>% 
+    rename(backend = rowname) %>% 
+    mutate(files = as.numeric(sub("X", "", files))) %>% 
+    ggplot(aes(x = files, y = time, colour = backend)) + 
+    geom_point() + 
+    geom_line() + 
+    scale_x_continuous(breaks=c(1, 5, 10)) + 
+    ylab("Reading time (s)") + 
+    theme(legend.position = "none")
+
+p2 <- data.frame(sz) %>% 
+    rownames_to_column() %>% 
+    pivot_longer(names_to = "files", 
+                 values_to = "size",
+                 -rowname) %>% 
+    rename(backend = rowname) %>% 
+    mutate(files = as.numeric(sub("X", "", files))) %>% 
+    ggplot(aes(x = files, y = size, colour = backend)) + 
+    geom_point() + 
+    geom_line() + 
+    scale_x_continuous(breaks=c(1, 5, 10)) + 
+    ylab("Object size (MB)") 
 @
 
-\begin{table}[ht]
+\begin{figure}[ht]
   \centering
-\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
-\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
-\caption{Reading time (in seconds) (left) and data size in memory (in
+<<plot_bench1, out.width = '100%', fig.width = 7, fig.height = 3>>=
+p1 + p2 + plot_annotation(tag_levels = 'a')
+@
+\caption{(a) Reading time (in seconds) and (b) data size in memory (in
   MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
   only) and 6103 MS2 (on-disk and in-memory) spectra. }
-\label{tab:bench1}
-\end{table}
+\label{fig:bench1}
+\end{figure}
 
@@
 
-As shown on figure \ref{fig:filt}, this lazy mechanism si
+As shown on figure~\ref{fig:filt}, this lazy mechanism si
 significantly faster than its application on in-memory data.
@@
 mandatory (for example for plotting), then the in-memory backend
-becomes more efficient, as illustrated on figure
-\ref{fig:access}. This gain is maximal when the whole dataset is the
-be accessed (i.e. all spectra are already in memory) and negligeable
-when large fractions of the data need to be subset.
+becomes more efficient, as illustrated on
+figure~\ref{fig:access}. This gain is maximal when the whole dataset
+is the be accessed (i.e. all spectra are already in memory) and
+negligeable when large fractions of the data need to be subset.
 
@@
 
-Enables large scale data analyses (metabolomics example, see Johannes)
+This new on-disk infrastructure enables large scale data analyses
+using \texttt{MSnbase} (metabolomics example, see Johannes).
 

### 2020-04-25 msnbase2.tex
replace bench1 tab with figure

@@
 in memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only load
-metadata in memory and accesses the spectra in the original files when
+was complemented by the on-disk backend, that only load metadata in
+memory and accesses the spectra in the original files when
 needed. There are two direct benefits using the on-disk
-backend. Table \ref{tab:bench1} shows 4-fold faster reading times
+backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
 (left) and 10-fold smaller memory footprint (right).
 
-\begin{table}[ht]
+\begin{figure}[ht]
   \centering
-\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
-\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
-\caption{Reading time (in seconds) (left) and data size in memory (in
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=1\linewidth]{figure/plot_bench1-1}
+
+\end{knitrout}
+\caption{(a) Reading time (in seconds) and (b) data size in memory (in
   MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
   only) and 6103 MS2 (on-disk and in-memory) spectra. }
-\label{tab:bench1}
-\end{table}
+\label{fig:bench1}
+\end{figure}
 

### 2020-04-25 msnbase2.Rnw
compose one bench figure

@@
 
-The former is the legacy storage mode, implemented in the first
-version of the package, that loads all the raw data and the metadata
-in memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only load metadata in
-memory and accesses the spectra in the original files when
-needed. There are two direct benefits using the on-disk
-backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
-(left) and 10-fold smaller memory footprint (right).
-
 <<time_sz>>=
@@
 
-p1 <- data.frame(time) %>% 
+p_read <- 
+    data.frame(time) %>% 
     rownames_to_column() %>% 
@@
     ylab("Reading time (s)") + 
+    xlab("Number of files") +
     theme(legend.position = "none")
 
-p2 <- data.frame(sz) %>% 
+p_sz <- 
+    data.frame(sz) %>% 
     rownames_to_column() %>% 
@@
     scale_x_continuous(breaks=c(1, 5, 10)) + 
-    ylab("Object size (MB)") 
+    ylab("Object size (MB)") +
+    xlab("Number of files") +
+    theme(legend.position = "none")    
 @
 
-\begin{figure}[ht]
+<<filt>>=
+load("bench_t_filt.rda")
+p_filt <- 
+    tibble(time = microbenchmark:::convert_to_unit(t_filt$time, "ms"),
+           expr = as.character(t_filt$expr)) %>%    
+    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
+    ggplot(aes(x = mode, y = time, fill = mode)) + 
+    ggplot2::geom_violin() + 
+    ggplot2::scale_y_log10() +
+    ylab("Filtering time [millisec]") + 
+    xlab("Backend") +    
+    theme(legend.position = "none")
+
+@
+
+<<access>>=
+load("bench_t_access.rda")
+p_access <- 
+    tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
+           expr = as.character(t_access$expr)) %>%
+    mutate(n = sub("^access_.+_", "", expr)) %>%
+    mutate(n = sub("all", 6103, n)) %>%
+    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
+    ggplot(aes(x = n, y = time, fill = mode)) + 
+    ggplot2::geom_violin() + 
+    ggplot2::scale_y_log10() +
+    ylab("Raw data access [millisec]") +
+    xlab("Number of spectra (out of 6103)") +
+    facet_wrap(~ mode) +
+    theme(legend.position = "none")
+@
+
+\begin{figure}[p]
   \centering
-<<plot_bench1, out.width = '100%', fig.width = 7, fig.height = 3>>=
-p1 + p2 + plot_annotation(tag_levels = 'a')
+<<plot_bench>>=
+(p_read + p_sz + p_filt) / p_access + 
+    plot_layout(heights = c(1, 0.75)) + 
+    plot_annotation(tag_levels = 'a') &
+    theme(axis.title = element_text(size = 10))
 @
@@
   MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
-  only) and 6103 MS2 (on-disk and in-memory) spectra. }
-\label{fig:bench1}
+  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
+  benchmark assessed over 10 interations on in-memory and on-disk data
+  containing 6103 MS2 spectra.  (d) Access time to spectra for the
+  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
+  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
+  red. }
+\label{fig:bench}
 \end{figure}
+
+The former is the legacy storage mode, implemented in the first
+version of the package, that loads all the raw data and the metadata
+in memory. This solution doesn't scale for modern large dataset, and
+was complemented by the on-disk backend, that only load metadata in
+memory and accesses the spectra in the original files when
+needed. There are two direct benefits using the on-disk
+backend. Figure \ref{fig:bench} shows 4-fold faster reading times (a)
+and 10-fold smaller memory footprint (b).
 
@@
 
-As shown on figure~\ref{fig:filt}, this lazy mechanism si
-significantly faster than its application on in-memory data.
-
-\begin{figure}[ht]
-  \centering
-<<plot_filter, message = FALSE, out.width = '50%', fig.width = 5, fig.height = 3>>=
-load("bench_t_filt.rda")
-ggplot2::autoplot(t_filt)
-@
-\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
-\label{fig:filt}
-\end{figure}
-
-The advantageous reading and execution times and memory footprint of
-the on-disk backend are possible by avoiding unnecessary access to the
-raw data. Once access to the spectra M/Z and intensity values become
+As shown on figure~\ref{fig:bench} (c), this lazy mechanism si
+significantly faster than its application on in-memory data. The
+advantageous reading and execution times and memory footprint of the
+on-disk backend are possible by avoiding unnecessary access to the raw
+data. Once access to the spectra M/Z and intensity values become
 mandatory (for example for plotting), then the in-memory backend
-becomes more efficient, as illustrated on
-figure~\ref{fig:access}. This gain is maximal when the whole dataset
-is the be accessed (i.e. all spectra are already in memory) and
-negligeable when large fractions of the data need to be subset.
-
-\begin{figure}[ht]
-  \centering
-<<plot_access, message = FALSE, out.width = '75%', fig.width = 5, fig.height = 3>>=
-load("bench_t_access.rda")
-tibble(time = microbenchmark:::convert_to_unit(t_access$time, "ms"),
-       expr = as.character(t_access$expr)) %>%
-    mutate(n = sub("^access_.+_", "", expr)) %>%
-    mutate(n = sub("all", 6103, n)) %>%
-    mutate(mode = if_else(grepl("mem", expr), "in-memory", "on-disk")) %>%
-    ggplot(aes(x = n, y = time, fill = log10(as.numeric(n)))) + 
-    ggplot2::geom_violin() + 
-    ggplot2::scale_y_log10() +
-    ylab("Time [millisec]") +
-    xlab("Number of spectra (out of 6103)") +
-    facet_wrap(~ mode) +
-    theme(legend.position = "none")
-@
-\caption{Access time to spectra for the in-memory (left) and on-disk
-  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
-\label{fig:access}
-\end{figure}
+becomes more efficient, as illustrated on figure~\ref{fig:bench}
+(d). This gain is maximal when the whole dataset is the be accessed
+(i.e. all spectra are already in memory) and negligeable when large
+fractions of the data need to be subset.
 

### 2020-04-25 msnbase2.tex
compose one bench figure

@@
 
+\begin{figure}[p]
+  \centering
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/plot_bench-1}
+
+\end{knitrout}
+\caption{(a) Reading time (in seconds) and (b) data size in memory (in
+  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
+  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
+  benchmark assessed over 10 interations on in-memory and on-disk data
+  containing 6103 MS2 spectra.  (d) Access time to spectra for the
+  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
+  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
+  red. }
+\label{fig:bench}
+\end{figure}
+
 The former is the legacy storage mode, implemented in the first
@@
 needed. There are two direct benefits using the on-disk
-backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
-(left) and 10-fold smaller memory footprint (right).
-
-\begin{figure}[ht]
-  \centering
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=1\linewidth]{figure/plot_bench1-1}
-
-\end{knitrout}
-\caption{(a) Reading time (in seconds) and (b) data size in memory (in
-  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
-  only) and 6103 MS2 (on-disk and in-memory) spectra. }
-\label{fig:bench1}
-\end{figure}
+backend. Figure \ref{fig:bench} shows 4-fold faster reading times (a)
+and 10-fold smaller memory footprint (b).
 
@@
 
-As shown on figure~\ref{fig:filt}, this lazy mechanism si
-significantly faster than its application on in-memory data.
-
-\begin{figure}[ht]
-  \centering
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=0.5\linewidth]{figure/plot_filter-1}
-
-\end{knitrout}
-\caption{Filtering benchmark assessed over 10 interations on in-memory (top) and on-disk (bottom) data containing 6103 MS2 spectra.}
-\label{fig:filt}
-\end{figure}
-
-The advantageous reading and execution times and memory footprint of
-the on-disk backend are possible by avoiding unnecessary access to the
-raw data. Once access to the spectra M/Z and intensity values become
+As shown on figure~\ref{fig:bench} (c), this lazy mechanism si
+significantly faster than its application on in-memory data. The
+advantageous reading and execution times and memory footprint of the
+on-disk backend are possible by avoiding unnecessary access to the raw
+data. Once access to the spectra M/Z and intensity values become
 mandatory (for example for plotting), then the in-memory backend
-becomes more efficient, as illustrated on
-figure~\ref{fig:access}. This gain is maximal when the whole dataset
-is the be accessed (i.e. all spectra are already in memory) and
-negligeable when large fractions of the data need to be subset.
-
-\begin{figure}[ht]
-  \centering
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=0.75\linewidth]{figure/plot_access-1}
-
-\end{knitrout}
-\caption{Access time to spectra for the in-memory (left) and on-disk
-  (right) backends for 1, 10, 100 1000, 5000 and all 6103 spectra.}
-\label{fig:access}
-\end{figure}
+becomes more efficient, as illustrated on figure~\ref{fig:bench}
+(d). This gain is maximal when the whole dataset is the be accessed
+(i.e. all spectra are already in memory) and negligeable when large
+fractions of the data need to be subset.
 

### 2020-04-26 msnbase2.Rnw
add boxcar section

@@
 
+In \texttt{MSnbase}, mass spectrometry experiments are handled as
+\texttt{MSnExp} objects. While the implementation is more complex, it
+is useful to schematise a raw data experiment as being composed of raw
+data, i.e. a collection of individual spectra, as well as
+spectra-level metadata. Each spectrum is composed of m/z values and
+associated intensities. The metadata are represented by a table with
+variables along the columns and each row associated to a
+spectrum. Among the metadata available for each spectrum, we there is
+its MS level, acquisition number, retention time, precursor m/z and
+intensity (for MS level 2 and above), and many more. \texttt{MSnbase}
+provides a rich interface to manipulate such objects. The code chunk
+below illustrates such an object as displayed in the R console and an
+enumeration of the metadata.
+
+<<msnexp, echo = TRUE, eval = FALSE>>=
+> show(ms)
+MSn experiment data ("OnDiskMSnExp")
+Object size in memory: 0.54 Mb
+- - - Spectra data - - -
+ MS level(s): 1 2 3 
+ Number of spectra: 994 
+ MSn retention times: 45:27 - 47:6 minutes
+- - - Processing information - - -
+Data loaded [Sun Apr 26 15:40:58 2020] 
+ MSnbase version: 2.13.6 
+- - - Meta data  - - -
+phenoData
+  rowNames: MS3TMT11.mzML
+  varLabels: sampleNames
+  varMetadata: labelDescription
+Loaded from:
+  MS3TMT11.mzML 
+protocolData: none
+featureData
+  featureNames: F1.S001 F1.S002 ... F1.S994 (994 total)
+  fvarLabels: fileIdx spIdx ... spectrum (35 total)
+  fvarMetadata: labelDescription
+experimentData: use 'experimentData(object)'
+> fvarLabels(ms)
+ [1] "fileIdx"                    "spIdx"
+ [3] "smoothed"                   "seqNum"
+ [5] "acquisitionNum"             "msLevel"
+ [7] "polarity"                   "originalPeaksCount"
+ [9] "totIonCurrent"              "retentionTime"
+[11] "basePeakMZ"                 "basePeakIntensity"
+[13] "collisionEnergy"            "ionisationEnergy"
+[15] "lowMZ"                      "highMZ"
+[17] "precursorScanNum"           "precursorMZ"
+[19] "precursorCharge"            "precursorIntensity"
+[21] "mergedScan"                 "mergedResultScanNum"
+[23] "mergedResultStartScanNum"   "mergedResultEndScanNum"
+[25] "injectionTime"              "filterString"
+[27] "spectrumId"                 "centroided"
+[29] "ionMobilityDriftTime"       "isolationWindowTargetMZ"
+[31] "isolationWindowLowerOffset" "isolationWindowUpperOffset"
+[33] "scanWindowLowerLimit"       "scanWindowUpperLimit"
+[35] "spectrum"
+@
+
 \subsection{On-disk backend}
@@
 addition of different backends for raw data storage, namely
-\textit{in-memory} and on-disk. The following code chunk
+\textit{in-memory} and \textit{on-disk}. The following code chunk
 demonstrates how to create two \texttt{MSnExp} objects, tailored to
@@
 in memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only load metadata in
+was complemented by the on-disk backend, that only loads metadata into
 memory and accesses the spectra in the original files when
-needed. There are two direct benefits using the on-disk
-backend. Figure \ref{fig:bench} shows 4-fold faster reading times (a)
-and 10-fold smaller memory footprint (b).
+needed. There are two direct benefits using the on-disk backend,
+namely faster reading and reduced memory footpring. Figure
+\ref{fig:bench} shows 5-fold faster reading times (a) and over a
+10-fold reduction in memory usage (b).
 
@@
 retains MS2 spectra acquired between 1000 and 3000 seconds, extract
-the M/Z range corresponding to the TMT 6-plex range and focuses on the
+the m/z range corresponding to the TMT 6-plex range and focuses on the
 MS2 spectra with a precursor intensity greater than $11 \times 10^6$
@@
 
-As shown on figure~\ref{fig:bench} (c), this lazy mechanism si
+As shown on Figure~\ref{fig:bench} (c), this lazy mechanism si
 significantly faster than its application on in-memory data. The
@@
 on-disk backend are possible by avoiding unnecessary access to the raw
-data. Once access to the spectra M/Z and intensity values become
+data. Once access to the spectra m/z and intensity values become
 mandatory (for example for plotting), then the in-memory backend
-becomes more efficient, as illustrated on figure~\ref{fig:bench}
+becomes more efficient, as illustrated on Figure~\ref{fig:bench}
 (d). This gain is maximal when the whole dataset is the be accessed
@@
 
-See https://github.com/lgatto/msnbase\_boxcar.
+The \texttt{MSnExp} data structure and its interface constitute a
+efficient prototyping environment for computational method
+development. We illustrate this by demonstrating how to implement the
+BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
+acquisition aims at improving the detection of intact precursor ions
+by distributing the charge capacity over multiple narrow m/z segments
+and thus limiting the proportion of highly abundant precursors in each
+segment. A full scan is reconstructed by combining the respective
+adjacent segments of the BoxCar acquisitions. The
+\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
+that demonstrates this. The simple method is composed of three steps:
+
+\begin{enumerate}
+
+\item Identify and filter the groups of spectra that represent
+  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)). This can be
+  done using the `filterString` metadata variable (see code chunk
+  above) that identifies BoxCar spectra by their adjacent M/Z segments
+  with the \texttt{bc\_groups()} function from \texttt{MSnbaseBoxCar}.
+
+\item Remove any signal outside the BoxCar segnments using the
+  \texttt{bc\_zero\_out\_box()} function from \texttt{MSnbaseBoxCar}
+  (Figures~\ref{fig:bc}~(c) and (d)).
+
+\item Using the \texttt{combineSpectra} function from the
+  \texttt{MSnbase}, combine the cleaned BoxCar spectra into a new,
+  full spectrum (Figure~\ref{fig:bc}~(e)).
+
+\end{enumerate}
+
+\begin{figure}[p]
+  \centering
+<<plot_boxcar, message=FALSE, warning=FALSE, fig.height=8>>=
+load("boxcar.rda")
+p1 + p2 + p3 + p4 + 
+    plot_layout(ncol = 1, 
+                heights = c(0.7, 1.2, 0.7, 0.7)) + 
+    plot_annotation(tag_levels = 'a') &
+    theme(axis.title = element_text(size = 10),
+          axis.text = element_text(size = 6),
+          plot.margin = margin(0, 0, 0, 0, "mm"))
+@
+\caption{BoxCar processing with \texttt{MSnbase}. (a) Standard full
+  scan with (b) three corresponding BoxCar scans showing the adjacent
+  segments. Figure (c) shows the overlapping intact BoxCar segments
+  and (d) the same segments after cleaning, i.e. where peaks outside
+  of the segments were removed. The reconstructed full scan is shown
+  on panel (e).}
+\label{fig:bc}
+\end{figure}
 

### 2020-04-26 msnbase2.tex
add boxcar section

@@
 
+In \texttt{MSnbase}, mass spectrometry experiments are handled as
+\texttt{MSnExp} objects. While the implementation is more complex, it
+is useful to schematise a raw data experiment as being composed of raw
+data, i.e. a collection of individual spectra, as well as
+spectra-level metadata. Each spectrum is composed of m/z values and
+associated intensities. The metadata are represented by a table with
+variables along the columns and each row associated to a
+spectrum. Among the metadata available for each spectrum, we there is
+its MS level, acquisition number, retention time, precursor m/z and
+intensity (for MS level 2 and above), and many more. \texttt{MSnbase}
+provides a rich interface to manipulate such objects. The code chunk
+below illustrates such an object as displayed in the R console and an
+enumeration of the metadata.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+> \hlkwd{show}(ms)
+MSn experiment \hlkwd{data} (\hlstr{"OnDiskMSnExp"})
+Object size in memory: 0.54 Mb
+- - - Spectra data - - -
+ MS \hlkwd{level}(s): 1 2 3 
+ Number of spectra: 994 
+ MSn retention times: 45:27 - 47:6 minutes
+- - - Processing information - - -
+Data loaded [Sun Apr 26 15:40:58 2020] 
+ MSnbase version: 2.13.6 
+- - - Meta data  - - -
+phenoData
+  rowNames: MS3TMT11.mzML
+  varLabels: sampleNames
+  varMetadata: labelDescription
+Loaded from:
+  MS3TMT11.mzML 
+protocolData: none
+featureData
+  featureNames: F1.S001 F1.S002 ... \hlkwd{F1.S994} (994 total)
+  fvarLabels: fileIdx spIdx ... \hlkwd{spectrum} (35 total)
+  fvarMetadata: labelDescription
+experimentData: use \hlstr{'\hlkwd{experimentData}(object)'}
+> \hlkwd{fvarLabels}(ms)
+ [1] \hlstr{"fileIdx"}                    \hlstr{"spIdx"}
+ [3] \hlstr{"smoothed"}                   \hlstr{"seqNum"}
+ [5] \hlstr{"acquisitionNum"}             \hlstr{"msLevel"}
+ [7] \hlstr{"polarity"}                   \hlstr{"originalPeaksCount"}
+ [9] \hlstr{"totIonCurrent"}              \hlstr{"retentionTime"}
+[11] \hlstr{"basePeakMZ"}                 \hlstr{"basePeakIntensity"}
+[13] \hlstr{"collisionEnergy"}            \hlstr{"ionisationEnergy"}
+[15] \hlstr{"lowMZ"}                      \hlstr{"highMZ"}
+[17] \hlstr{"precursorScanNum"}           \hlstr{"precursorMZ"}
+[19] \hlstr{"precursorCharge"}            \hlstr{"precursorIntensity"}
+[21] \hlstr{"mergedScan"}                 \hlstr{"mergedResultScanNum"}
+[23] \hlstr{"mergedResultStartScanNum"}   \hlstr{"mergedResultEndScanNum"}
+[25] \hlstr{"injectionTime"}              \hlstr{"filterString"}
+[27] \hlstr{"spectrumId"}                 \hlstr{"centroided"}
+[29] \hlstr{"ionMobilityDriftTime"}       \hlstr{"isolationWindowTargetMZ"}
+[31] \hlstr{"isolationWindowLowerOffset"} \hlstr{"isolationWindowUpperOffset"}
+[33] \hlstr{"scanWindowLowerLimit"}       \hlstr{"scanWindowUpperLimit"}
+[35] \hlstr{"spectrum"}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
 \subsection{On-disk backend}
@@
 
-See https://github.com/lgatto/msnbase\_boxcar.
+The \texttt{MSnExp} data structure and its interface constitute a
+efficient prototyping environment for computational method
+development. We illustrate this by demonstrating how to implement the
+BoxCar\cite{Meier:2018} acquisition method. In a nutshell, BoxCar
+acquisition aims at improving the detection of intact precursor ions
+by distributing the charge capacity over multiple narrow m/z segments
+and thus limiting the proportion of highly abundant precursors in each
+segment. A full scan is reconstructed by combining the respective
+adjacent segments of the BoxCar acquisitions. The
+\texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
+that demonstrates this. The simple method is composed of three steps:
+
+\begin{enumerate}
+
+\item Identify and filter the groups of spectra that represent
+  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)). This can be
+  done using the `filterString` metadata variable (see code chunk
+  above) that identifies BoxCar spectra by their adjacent M/Z segments
+  with the \texttt{bc\_groups()} function from \texttt{MSnbaseBoxCar}.
+
+\item Remove any signal outside the BoxCar segnments using the
+  \texttt{bc\_zero\_out\_box()} function from \texttt{MSnbaseBoxCar}
+  (Figures~\ref{fig:bc}~(c) and (d)).
+
+\item Using the \texttt{combineSpectra} function from the
+  \texttt{MSnbase}, combine the cleaned BoxCar spectra into a new,
+  full spectrum (Figure~\ref{fig:bc}~(e)).
+
+\end{enumerate}
+
+\begin{figure}[p]
+  \centering
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/plot_boxcar-1}
+
+\end{knitrout}
+\caption{BoxCar processing with \texttt{MSnbase}. (a) Standard full
+  scan with (b) three corresponding BoxCar scans showing the adjacent
+  segments. Figure (c) shows the overlapping intact BoxCar segments
+  and (d) the same segments after cleaning, i.e. where peaks outside
+  of the segments were removed. The reconstructed full scan is shown
+  on panel (e).}
+\label{fig:bc}
+\end{figure}
 

### 2020-04-26 S02-boxcar.R

@@ -1,0 +1,51 @@
+## -------------------------------------
+## Based on the MSnbaseBoxCar vignette
+## -------------------------------------
+
+library("MSnbaseBoxCar")
+library("magrittr")
+library("ggplot2")
+library("patchwork")
+
+f <- dir(system.file("extdata", package = "MSnbaseBoxCar"),
+         pattern = "boxcar.mzML",
+         full.names = TRUE)
+basename(f)
+x <- readMSData(f, mode = "onDisk") %>%
+    filterMz(c(299, 555))
+
+fData(x)$groups <- bc_groups(x)
+xbc <- x[bc_is_boxcar(x)]
+xbc2 <- xbc %>%
+    bc_zero_out_box(offset = 0)
+
+res  <- combineSpectra(xbc,
+                       fcol = "groups",
+                       method = boxcarCombine)
+
+
+p1 <- plot(x[[1]], plot = FALSE) +
+    labs(title = element_text("")) +
+    xlab(element_text(""))
+
+p2 <- plot(x[2:4], plot = FALSE) +
+    labs(title = element_text("")) +
+    xlab(element_text(""))
+
+p01 <- bc_plot(xbc[2:4]) +
+    xlim(290, 455) +
+    xlab(element_text(""))
+p02 <- bc_plot(xbc2[2:4]) +
+    xlim(290, 455) +
+    xlab(element_text("")) + 
+    ylab(element_text(""))
+p3 <- p01 + p02
+
+p4 <- plot(res[[1]], plot = FALSE) +
+    labs(title = element_text(""))
+
+
+save(p1, p2, p3, p4, file = "boxcar.rda")
+
+
+

### 2020-04-27 msnbase2.Rnw
add boxcar code, finish section

@@
 <<filter, eval = FALSE, echo = TRUE>>=
-x <- x_dsk %>%
+ms <- ms %>%
     filterRt(c(1000, 3000)) %>%
     filterMz(120, 135)
-x[precursorIntensity(x) > 11e6, ]
+ms[precursorIntensity(ms) > 11e6, ]
 @
@@
 \texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
-that demonstrates this. The simple method is composed of three steps:
+that demonstrates this. The simple method is composed of three steps
+is described below and illustrated with code from
+\texttt{MSnbaseBoxCar} in the following code chunk.
 
@@
 \item Identify and filter the groups of spectra that represent
-  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)). This can be
-  done using the `filterString` metadata variable (see code chunk
-  above) that identifies BoxCar spectra by their adjacent M/Z segments
-  with the \texttt{bc\_groups()} function from \texttt{MSnbaseBoxCar}.
+  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)).. This can be
+  done using the `filterString` metadata variable that identifies
+  BoxCar spectra by their adjacent M/Z segments with the
+  \texttt{bc\_groups()} function and filtering relevant spectra with
+  the \texttt{filterBoxCar()}.
 
@@
 \end{enumerate}
+
+<<bc1, echo=TRUE, eval=FALSE>>=
+raw_bc %>% 
+    bc_groups() %>%       ## identify BoxCar groups (bc_groups)
+    filterBoxCar() %>%    ## keep only BoxCar spectra
+    bc_zero_out_box() %>% ## remove signal outside of BoxCar segments
+    combineSpectra(fcol = "bc_groups", ## reconstruct full spectrum
+                   method = boxcarCompbine)
+@
 
@@
 
+All the functions for the processing of BoxCar spectra and segments in
+\texttt{MSnbaseBoxCar} were developed using existing functionality
+implemented in \texttt{MSnbase}, illustrating the flexibility and
+adaptability of the \texttt{MSnbase} package for computational mass
+spectrometry method development.
+
 \subsection{Visualisation}
 
-Examples: 3D MSmap, boxcar, centroiding vignette
+The R environment is well known for the quality of its visualiation
+capacity. This also holds true for mass
+spectrometry\cite{Gatto:2015,protViz}.
 

### 2020-04-27 msnbase2.tex
add boxcar code, finish section

@@
 \begin{alltt}
-\hlstd{x} \hlkwb{<-} \hlstd{x_dsk} \hlopt{%>%}
+\hlstd{ms} \hlkwb{<-} \hlstd{ms} \hlopt{%>%}
     \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{1000}\hlstd{,} \hlnum{3000}\hlstd{))} \hlopt{%>%}
     \hlkwd{filterMz}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{135}\hlstd{)}
-\hlstd{x[}\hlkwd{precursorIntensity}\hlstd{(x)} \hlopt{>} \hlnum{11e6}\hlstd{, ]}
+\hlstd{ms[}\hlkwd{precursorIntensity}\hlstd{(ms)} \hlopt{>} \hlnum{11e6}\hlstd{, ]}
 \end{alltt}
@@
 \end{enumerate}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{raw_bc} \hlopt{%>%}
+    \hlkwd{bc_groups}\hlstd{()} \hlopt{%>%}       \hlcom{## identify BoxCar groups (bc_groups)}
+    \hlkwd{filterBoxCar}\hlstd{()} \hlopt{%>%}    \hlcom{## keep only BoxCar spectra}
+    \hlkwd{bc_zero_out_box}\hlstd{()} \hlopt{%>%} \hlcom{## remove signal outside of BoxCar segments}
+    \hlkwd{combineSpectra}\hlstd{(}\hlkwc{fcol} \hlstd{=} \hlstr{"bc_groups"}\hlstd{,} \hlcom{## reconstruct full spectrum}
+                   \hlkwc{method} \hlstd{= boxcarCompbine)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 
+All the functions for the processing of BoxCar spectra and segments in
+\texttt{MSnbaseBoxCar} were developed using existing functionality
+implemented in \texttt{MSnbase}, illustrating the flexibility and
+adaptability of the \texttt{MSnbase} package for computational mass
+spectrometry method development.
+
 \subsection{Visualisation}
 
-Examples: 3D MSmap, boxcar, centroiding vignette
+The R environment is well known for the quality of its visualiation
+capacity. This also holds true for mass spectrometry
+data~\cite{Gatto:2015,protViz}.
 

### 2020-04-27 S02-boxcar.R

@@ -14,13 +14,11 @@ basename(f)
 x <- readMSData(f, mode = "onDisk") %>%
     filterMz(c(299, 555))
 
-fData(x)$groups <- bc_groups(x)
-xbc <- x[bc_is_boxcar(x)]
-xbc2 <- xbc %>%
-    bc_zero_out_box(offset = 0)
-
-res  <- combineSpectra(xbc,
-                       fcol = "groups",
+x <- bc_groups(x)
+xbc <- filterBoxCar(x)
+xbc2 <- bc_zero_out_box(xbc, offset = 0)
+res  <- combineSpectra(xbc2,
+                       fcol = "bc_groups",
                        method = boxcarCombine)
 
 

### 2020-04-27 S03-vis.R

@@ -1,0 +1,18 @@
+f <- dir(system.file("extdata", package = "MSnbaseBoxCar"),
+         pattern = "boxcar.mzML",
+         full.names = TRUE)
+basename(f)
+
+x <- readMSData(f, mode = "onDisk") %>%    
+    bc_groups()
+
+fData(x)$msLevel <- ifelse(is.na(fData(x)$bc_groups), 1, 2)
+
+map <- MSmap(x, 1:8, 200, 1000, .005)
+
+vis0 <- plot3D(map)
+
+
+
+
+

### 2020-04-27 msnbase2.Rnw
vis fig (from centroiding vignette)

@@
 <<bc1, echo=TRUE, eval=FALSE>>=
-raw_bc %>% 
+bc <- readMSData("boxcar.mzML", mode = "onDisk") %>%
       bc_groups() %>%       ## identify BoxCar groups (bc_groups)
@@
 capacity. This also holds true for mass
-spectrometry\cite{Gatto:2015,protViz}.
+spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
+conclude the overview of version 2 of the \texttt{MSnbase} package by
+highlighting the flexibility of the software to visualise and assess
+the efficiency of raw data processing.
 
@@
 \item{potential for reuse}: see
-  \cite{Wieczorek:2017,Griss:2019,Smith:2006} for examples.
+  \cite{Smith:2006,Wieczorek:2017,Griss:2019} for examples.
 \item{general limitations}

### 2020-04-27 S02-boxcar.R

@@ -43,7 +43,20 @@ p4 <- plot(res[[1]], plot = FALSE) +
     labs(title = element_text(""))
 
 
-save(p1, p2, p3, p4, file = "boxcar.rda")
 
 
+## 3D MS map
+x <- readMSData(f, mode = "onDisk") %>%    
+    bc_groups()
+## use MS level to draw the full and BoxCar spectra in different
+## colours
+fData(x)$msLevel <- ifelse(is.na(fData(x)$bc_groups), 1, 2)
 
+map <- MSmap(x, 1:12, 200, 1000, .005)
+
+lattice::trellis.par.set("axis.line",
+                         list(col = NA,lty = 1,lwd = 1))
+pmap <- plot3D(map)
+
+
+save(p1, p2, p3, p4, pmap, file = "boxcar.rda")

### 2020-04-27 S03-vis.R

@@ -1,18 +1,40 @@
-f <- dir(system.file("extdata", package = "MSnbaseBoxCar"),
-         pattern = "boxcar.mzML",
-         full.names = TRUE)
-basename(f)
+library("MSnbase")
+library("msdata")
+library("magrittr")
 
-x <- readMSData(f, mode = "onDisk") %>%    
-    bc_groups()
+fl <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)[2]
+basename(fl)
+data_prof <- readMSData(fl, mode = "onDisk", centroided = FALSE)
 
-fData(x)$msLevel <- ifelse(is.na(fData(x)$bc_groups), 1, 2)
+## Define the mz and retention time ranges
+serine_mz <- 106.049871
+mzr <- c(serine_mz - 0.01, serine_mz + 0.01)
+rtr <- c(175, 187)
 
-map <- MSmap(x, 1:8, 200, 1000, .005)
-
-vis0 <- plot3D(map)
-
-
-
+## Filtering the object
+serine <- data_prof %>%
+    filterRt(rtr) %>%
+    filterMz(mzr)
 
 
+data_sg <- data_prof %>%
+    smooth(method = "SavitzkyGolay", halfWindowSize = 4L)
+
+
+## Use pickPeaks with descendPeak m/z refinement
+data_sg_cent_mz <- data_sg %>%
+    pickPeaks(refineMz = "descendPeak")
+
+serine_sg_cent_mz <- data_sg_cent_mz %>%
+    filterRt(rtr) %>%
+    filterMz(mzr)
+
+
+pdf("./figure/centroiding.pdf", width = 12, height = 7)
+layout(matrix(1:4, ncol = 2))
+plot(serine, type = "XIC", layout = NULL)
+abline(h = serine_mz, col = "red", lty = 2)
+plot(serine_sg_cent_mz, type = "XIC", layout = NULL)
+abline(h = serine_mz, col = "red", lty = 2)
+abline(v = rtime(serine_sg_cent_mz)[22], col = "red", lty = 3)
+dev.off()

### 2020-04-27 msnbase2.Rnw
add vis/proc section

@@
 @
+
+We will also illustrate how \texttt{MSnbase} makes use of the
+forward-pipe operator (\texttt{\%>\%}) defined in the
+\texttt{magrittr} package. This operator has proved useful to develop
+non-trivial pipelines by combining individual functions into easily
+readable operations.
 
@@
 highlighting the flexibility of the software to visualise and assess
-the efficiency of raw data processing.
+the efficiency of raw data processing. Figure \ref{fig:cent} compares
+the raw MS profile data for the serine and the same data after
+smoothing, centroiding and M/Z refinement.
+
+<<serine_processing, echo=TRUE, eval=FALSE>>=
+serine_mz <- 106.049871
+serine_proc <- ms %>%
+    smooth(method = "SavitzkyGolay", halfWindowSize = 4L) %>% 
+    pickPeaks(refineMz = "descendPeak") %>%
+    filterRt(c(serine_mz - 0.01, serine_mz + 0.01)) %>%
+    filterMz(c(175, 187))
+@
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{./figure/centroiding.pdf}
+  \caption{Visualisation of data smoothing and M/Z refinement using
+    \texttt{MSnbase}. (a) Raw MS profile data for serine. Upper panel
+    shows the base peak chromatogram (BPC), lower panel the individual
+    signals in the retention time -- M/Z space. The horizontal dashed
+    red line indicates the theoretical M/Z of the [M+H]+ adduct of
+    serine. (b) Smoothed and centroided data for serine with M/Z
+    refinement. The horizontal red dashed line indicates the
+    theoretical M/Z for the [M+H]+ ion and the vertical red dotted
+    line the position of the maximum signal. }
+  \label{fig:cent}
+\end{figure}
 

### 2020-04-27 msnbase2.tex
vis fig (from centroiding vignette)
add vis/proc section

@@
 
+We will also illustrate how \texttt{MSnbase} makes use of the
+forward-pipe operator (\texttt{\%>\%}) defined in the
+\texttt{magrittr} package. This operator has proved useful to develop
+non-trivial pipelines by combining individual functions into easily
+readable operations.
+
 \subsection{On-disk backend}
@@
 \begin{alltt}
-\hlstd{raw_bc} \hlopt{%>%}
+\hlstd{bc} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"boxcar.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"onDisk"}\hlstd{)} \hlopt{%>%}
       \hlkwd{bc_groups}\hlstd{()} \hlopt{%>%}       \hlcom{## identify BoxCar groups (bc_groups)}
@@
 The R environment is well known for the quality of its visualiation
-capacity. This also holds true for mass spectrometry
-data~\cite{Gatto:2015,protViz}.
+capacity. This also holds true for mass
+spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
+conclude the overview of version 2 of the \texttt{MSnbase} package by
+highlighting the flexibility of the software to visualise and assess
+the efficiency of raw data processing. Figure \ref{fig:cent} compares
+the raw MS profile data for the serine and the same data after
+smoothing, centroiding and M/Z refinement.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{serine_mz} \hlkwb{<-} \hlnum{106.049871}
+\hlstd{serine_proc} \hlkwb{<-} \hlstd{ms} \hlopt{%>%}
+    \hlkwd{smooth}\hlstd{(}\hlkwc{method} \hlstd{=} \hlstr{"SavitzkyGolay"}\hlstd{,} \hlkwc{halfWindowSize} \hlstd{=} \hlnum{4L}\hlstd{)} \hlopt{%>%}
+    \hlkwd{pickPeaks}\hlstd{(}\hlkwc{refineMz} \hlstd{=} \hlstr{"descendPeak"}\hlstd{)} \hlopt{%>%}
+    \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(serine_mz} \hlopt{-} \hlnum{0.01}\hlstd{, serine_mz} \hlopt{+} \hlnum{0.01}\hlstd{))} \hlopt{%>%}
+    \hlkwd{filterMz}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{175}\hlstd{,} \hlnum{187}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{./figure/centroiding.pdf}
+  \caption{Visualisation of data smoothing and M/Z refinement using
+    \texttt{MSnbase}. (a) Raw MS profile data for serine. Upper panel
+    shows the base peak chromatogram (BPC), lower panel the individual
+    signals in the retention time -- M/Z space. The horizontal dashed
+    red line indicates the theoretical M/Z of the [M+H]+ adduct of
+    serine. (b) Smoothed and centroided data for serine with M/Z
+    refinement. The horizontal red dashed line indicates the
+    theoretical M/Z for the [M+H]+ ion and the vertical red dotted
+    line the position of the maximum signal. }
+  \label{fig:cent}
+\end{figure}
 

### 2020-04-27 S03-vis.R

@@ -11,30 +11,25 @@ serine_mz <- 106.049871
 mzr <- c(serine_mz - 0.01, serine_mz + 0.01)
 rtr <- c(175, 187)
 
-## Filtering the object
+## Raw MS profile
 serine <- data_prof %>%
     filterRt(rtr) %>%
     filterMz(mzr)
 
-
-data_sg <- data_prof %>%
-    smooth(method = "SavitzkyGolay", halfWindowSize = 4L)
-
-
-## Use pickPeaks with descendPeak m/z refinement
-data_sg_cent_mz <- data_sg %>%
-    pickPeaks(refineMz = "descendPeak")
-
-serine_sg_cent_mz <- data_sg_cent_mz %>%
+## Processing
+serine_proc <- data_prof %>%
+    smooth(method = "SavitzkyGolay", halfWindowSize = 4L) %>% 
+    pickPeaks(refineMz = "descendPeak") %>%
     filterRt(rtr) %>%
     filterMz(mzr)
 
 
-pdf("./figure/centroiding.pdf", width = 12, height = 7)
+pdf("./figure/centroiding.pdf", width = 9, height = 6)
 layout(matrix(1:4, ncol = 2))
 plot(serine, type = "XIC", layout = NULL)
 abline(h = serine_mz, col = "red", lty = 2)
-plot(serine_sg_cent_mz, type = "XIC", layout = NULL)
+plot(serine_proc, type = "XIC", layout = NULL)
 abline(h = serine_mz, col = "red", lty = 2)
 abline(v = rtime(serine_sg_cent_mz)[22], col = "red", lty = 3)
 dev.off()
+

### 2020-04-27 msnbase2.Rnw
add discussion

@@
 In this software note, we present version 2 of the \texttt{MSnbase}
-\cite{Gatto:2012} R/Bioconductor software package. \texttt{MSnbase}
+\cite{Gatto:2012} software package, available from the
+Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase}
 offers a platform that lies between low level libraries and end-use
@@
 the raw MS profile data for the serine and the same data after
-smoothing, centroiding and M/Z refinement.
+smoothing, centroiding and M/Z refinement. Detailed execution and
+description of these operations can be found in the \emph{MSnbase:
+  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.
 
@@
 
-To address (from guidelines):
+The first public commit to the \texttt{MSnbase} GitHub repository was
+in October 2010. Since then, the package benefited from 12
+contributors\cite{contribs} that added various features, some
+particularly significant ones such as the on-disk backend described
+above. According to the package's Bioconductor page, there are 36
+Bioconductor packages that depend, import or suggest it. Among these
+are \texttt{pRoloc}~\cite{Gatto:2014a}, \texttt{DAPAR} and
+\texttt{ProStaR}~\cite{Wieczorek:2017},
+\texttt{RMassBank}~\cite{Stravs:2013},
+\texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
+\texttt{MSstatsQC}~\cite{Dogu:2017} and the widerly used
+\texttt{xcms}~\cite{Smith:2006}. \texttt{MSnbase} is also used in
+packages that are not on Bioconductor, such as
+IsoProt~\cite{Griss:2019}, for examples. The package currently ranks
+101 out of 1823 packages based on the monthly downloads from unique IP
+addresses, tallying over 1000 downloads from unique IP addresses each
+months.
 
-\begin{itemize}
+As is custom with Bioconductor packages, \texttt{MSnbase} comes with
+ample documentation. Every user-accessible function is documented in a
+dedicated manual page. In addition, the package offers 5 vignettes,
+including one aimed at developers. Question from users and developers
+are answered on the Bioconductor support forum as well as on the
+package GitHub page. The package provides several sample and
+benchmarking datasets, and relies on other dedicated
+\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
+raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
+data.
 
-\item{potential for reuse}: see
-  \cite{Smith:2006,Wieczorek:2017,Griss:2019} for examples.
-\item{general limitations}
-\item{system limitations}
-\item{end-user documentation}
-\item{developer documentation}
-\item{sample data}
-\item{benchmark data set}
-\item{availability}
-\item{license information}
-\item{system requirements}
-
-\end{itemize}
-
-Collaborative development, 11 contributors since creation (see blog
-post).
-
-Count packages depending on \texttt{MSnbase}.
+\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
+open source Artistic 2.0 license and easily installable using standard
+installation procedures.
 
@@
 
+<<version>>=
+v <- packageVersion("MSnbase")
+@
+
 The version of \texttt{MSnbase} used in this manuscritp is version
-2.10.0. The main features presented here
-were available since version 2.0.
+\Sexpr{v}. The main features presented here were available since
+version 2.0.
 

### 2020-04-27 msnbase2.tex
add discussion

@@
 The version of \texttt{MSnbase} used in this manuscritp is version
-2.10.0. The main features presented here
-were available since version 2.0.
+2.13.9. The main features presented here were available since
+version 2.0.
 

### 2020-04-27 msnbase2.Rnw
first draft ready

@@
 \makeatother
-
-%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-%% \let\hlipl\hlkwb
-
-%% \usepackage{framed}
-%% \makeatletter
-%% \newenvironment{kframe}{%
-%%  \def\at@end@of@kframe{}%
-%%  \ifinner\ifhmode%
-%%   \def\at@end@of@kframe{\end{minipage}}%
-%%   \begin{minipage}{\columnwidth}%
-%%  \fi\fi%
-%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
-%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
-%%      % There is no \\@totalrightmargin, so:
-%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
-%%  \MakeFramed {\advance\hsize-\width
-%%    \@totalleftmargin\z@ \linewidth\hsize
-%%    \@setminipage}}%
-%%  {\par\unskip\endMakeFramed%
-%%  \at@end@of@kframe}
-%% \makeatother
-
-%% \usepackage{xcolor}
-
-%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
-%% \definecolor{messagecolor}{rgb}{0, 0, 0}
-%% \definecolor{warningcolor}{rgb}{1, 0, 1}
-%% \definecolor{errorcolor}{rgb}{1, 0, 0}
-%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
 
@@
 \title[MSnbase version 2]
-  {\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data}
+  {\texttt{MSnbase}, efficient R-based processing and visualisation of raw mass spectrometry data}
 
@@
 Mass spectrometry is a powerful technology to assays chemical and
-biological samples. It is used routinely, with well characterised
+biological samples. It is used routinely with well characterised
 protocol, as well a development platform, to improve on existing
-methods and devise new ones to analyse ever more complex sample in
-greater details. The complexity and diversity of mass spectrometry
-yields data that is itself complex and often times of considerable
-size, that requires non trivial processing before producing
-interpretable results. This is particularly relevant, and can
-constitue a significant challenge for method developers that, in
+methods and devise new ones. The complexity and diversity of mass
+spectrometry yields data that is itself complex and often times of
+considerable size, that requires non trivial processing before
+producing interpretable results. This is particularly relevant, and
+can constitute a significant challenge for method developers: in
 addition to the development of sample processing and mass spectrometry
-methods, need to process and analyse these new data to demonstrate the
-improvement in their technical and analytical work.
+methods, there is a need to process, analyse, interpret and assess
+these new data to demonstrate the improvement in their technical and
+analytical workflows.
 
@@
 process and interpret mass spectrometry data. These range from low
-level software libraries such as vendor libraries, jmzML (ref),
-proteowizard (ref), ... that are aimed at programmers to develop new
-applications, to user-oriented applications, such as
-ProteomeDiscoverer, MaxQuant, ... that provide a limited and fixed set
-of functionality. The former are used through application programming
-interfaces exclusively, while the latter generally featuring graphical
-user interfaces (GUI).
-
-TODO: Give examples of libraries re-used in user/gui focused
-application...
+level software libraries that are aimed at programmers to develop new
+applications, to user-oriented applications. Example of software
+libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
+ProteoWizard~\cite{Chambers:2012}, that are used as application
+programming interfaces. ProteomeDiscoverer (Thermo Scientific),
+MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
+among the most widely user-centric application. They generally provide
+a limited and fixed set of functionality featuring often advanced
+graphical user interfaces (GUI).
 
@@
 \cite{Gatto:2012} software package, available from the
-Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase}
-offers a platform that lies between low level libraries and end-use
-software. It provides a flexible command line environment for
-metabolomics and proteomics mass spectrometry-based application, that
+Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase},
+like other Python-based software packages such as
+{pyOpenMS}~\cite{Rost:2014}, spectrum\_utils~\cite{Bittremieux:2020}
+or Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies
+between low level libraries and end-use software. \texttt{MSnbase}
+provides a flexible R~\cite{R} command line environment for
+metabolomics and proteomics mass spectrometry-based application. It
 allows a detailed step-by-step processing, analysis and exploration of
@@
 variables along the columns and each row associated to a
-spectrum. Among the metadata available for each spectrum, we there is
-its MS level, acquisition number, retention time, precursor m/z and
-intensity (for MS level 2 and above), and many more. \texttt{MSnbase}
-provides a rich interface to manipulate such objects. The code chunk
-below illustrates such an object as displayed in the R console and an
-enumeration of the metadata.
+spectrum. Among the metadata available for each spectrum, there are MS
+level, acquisition number, retention time, precursor m/z and intensity
+(for MS level 2 and above), and many more. \texttt{MSnbase} provides a
+rich interface to manipulate such objects. The code chunk below
+illustrates such an object as displayed in the R console and an
+enumeration of the metadata fields.
 
@@
 
-We will also illustrate how \texttt{MSnbase} makes use of the
-forward-pipe operator (\texttt{\%>\%}) defined in the
-\texttt{magrittr} package. This operator has proved useful to develop
-non-trivial pipelines by combining individual functions into easily
-readable operations.
+In the following sections, we will describe demonstrate how
+\texttt{MSnbase} effichiently manages large raw mass spectrometry data
+and how it is used for data processing and visualisation. We will also
+illustrate how it makes use of the forward-pipe operator
+(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
+operator has proved useful to develop non-trivial analyses by
+combining individual functions into easily readable pipelines.
 
@@
   done using the `filterString` metadata variable that identifies
-  BoxCar spectra by their adjacent M/Z segments with the
+  BoxCar spectra by their adjacent m/z segments with the
   \texttt{bc\_groups()} function and filtering relevant spectra with
@@
 the raw MS profile data for the serine and the same data after
-smoothing, centroiding and M/Z refinement. Detailed execution and
+smoothing, centroiding and m/z refinement. Detailed execution and
 description of these operations can be found in the \emph{MSnbase:
@@
   \includegraphics[width=\linewidth]{./figure/centroiding.pdf}
-  \caption{Visualisation of data smoothing and M/Z refinement using
+  \caption{Visualisation of data smoothing and m/z refinement using
     \texttt{MSnbase}. (a) Raw MS profile data for serine. Upper panel
     shows the base peak chromatogram (BPC), lower panel the individual
-    signals in the retention time -- M/Z space. The horizontal dashed
-    red line indicates the theoretical M/Z of the [M+H]+ adduct of
-    serine. (b) Smoothed and centroided data for serine with M/Z
+    signals in the retention time -- m/z space. The horizontal dashed
+    red line indicates the theoretical m/z of the [M+H]+ adduct of
+    serine. (b) Smoothed and centroided data for serine with m/z
     refinement. The horizontal red dashed line indicates the
-    theoretical M/Z for the [M+H]+ ion and the vertical red dotted
+    theoretical m/z for the [M+H]+ ion and the vertical red dotted
     line the position of the maximum signal. }
@@
 Bioconductor packages that depend, import or suggest it. Among these
-are \texttt{pRoloc}~\cite{Gatto:2014a}, \texttt{DAPAR} and
-\texttt{ProStaR}~\cite{Wieczorek:2017},
-\texttt{RMassBank}~\cite{Stravs:2013},
+are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
+spectrometry-based spatial proteomics data,
 \texttt{msmsTests}~\cite{msmsTests}, \texttt{DEP}~\cite{Zhang:2018},
-\texttt{MSstatsQC}~\cite{Dogu:2017} and the widerly used
-\texttt{xcms}~\cite{Smith:2006}. \texttt{MSnbase} is also used in
-packages that are not on Bioconductor, such as
-IsoProt~\cite{Griss:2019}, for examples. The package currently ranks
-101 out of 1823 packages based on the monthly downloads from unique IP
-addresses, tallying over 1000 downloads from unique IP addresses each
-months.
+\texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
+statistical analysis for quantitative proteomics data,
+\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
+MS files and build MassBank records,
+\texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
+suitability monitoring and quality control of targeted proteomic
+experiments and the widely used \texttt{xcms}~\cite{Smith:2006} packge
+for the processing and analysis of metabolomics data. \texttt{MSnbase}
+is also used in non-R/Bioconductor software, such as for example
+IsoProt~\cite{Griss:2019}, that provides a reproducible workflow for
+iTRAQ/TMT experiments. \texttt{MSnbase} currently ranks 101 out of
+1823 packages based on the monthly downloads from unique IP addresses,
+tallying over 1000 downloads from unique IP addresses each months.
 
@@
 raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
-data.
-
-\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
-open source Artistic 2.0 license and easily installable using standard
-installation procedures.
-
-Future developments.
+data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under
+the open source Artistic 2.0 license and easily installable using
+standard installation procedures.
 
@@
 
-The version of \texttt{MSnbase} used in this manuscritp is version
+The version of \texttt{MSnbase} used in this manuscript is
 \Sexpr{v}. The main features presented here were available since

### 2020-04-27 msnbase2.tex
first draft ready

@@
 
-%% \definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-%% \newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-%% \newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-%% \newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-%% \newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-%% \newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-%% \newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-%% \newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-%% \newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-%% \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-%% \let\hlipl\hlkwb
-
-%% \usepackage{framed}
-%% \makeatletter
-%% \newenvironment{kframe}{%
-%%  \def\at@end@of@kframe{}%
-%%  \ifinner\ifhmode%
-%%   \def\at@end@of@kframe{\end{minipage}}%
-%%   \begin{minipage}{\columnwidth}%
-%%  \fi\fi%
-%%  \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
-%%  \colorbox{shadecolor}{##1}\hskip-\fboxsep
-%%      % There is no \\@totalrightmargin, so:
-%%      \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
-%%  \MakeFramed {\advance\hsize-\width
-%%    \@totalleftmargin\z@ \linewidth\hsize
-%%    \@setminipage}}%
-%%  {\par\unskip\endMakeFramed%
-%%  \at@end@of@kframe}
-%% \makeatother
-
-%% \usepackage{xcolor}
-
-%% \definecolor{shadecolor}{rgb}{.97, .97, .97}
-%% \definecolor{messagecolor}{rgb}{0, 0, 0}
-%% \definecolor{warningcolor}{rgb}{1, 0, 1}
-%% \definecolor{errorcolor}{rgb}{1, 0, 0}
-%% \newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
 \usepackage{subfig}
@@
 raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
-data.
+data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under
+the open source Artistic 2.0 license and easily installable using
+standard installation procedures.
 
-\texttt{MSnbase} is available on Windows, Mac OS and Linux under the
-open source Artistic 2.0 license and easily installable using standard
-installation procedures.
-
-Future developments.
-
-The version of \texttt{MSnbase} used in this manuscritp is version
+The version of \texttt{MSnbase} used in this manuscript is
 2.13.9. The main features presented here were available since

### 2020-04-27 msnbase2.Rnw
fix one of many typos

@@
 The authors thank the various contributors and users who have provided
-constructuve input and feedback that have helped, over the years, the
+constructive input and feedback that have helped, over the years, the
 improvement of the package. The authors declare no conflict of

### 2020-04-27 msnbase2.Rnw
minor fixes

@@
 a limited and fixed set of functionality featuring often advanced
-graphical user interfaces (GUI).
+graphical user interfaces.
 
@@
 \cite{Gatto:2012} software package, available from the
-Bioconductor~\cite{Huber:2015} project since 2011. \texttt{MSnbase},
-like other Python-based software packages such as
-{pyOpenMS}~\cite{Rost:2014}, spectrum\_utils~\cite{Bittremieux:2020}
-or Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies
-between low level libraries and end-use software. \texttt{MSnbase}
-provides a flexible R~\cite{R} command line environment for
-metabolomics and proteomics mass spectrometry-based application. It
-allows a detailed step-by-step processing, analysis and exploration of
-the data and development of novel computational mass spectrometry
-methods.
+Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
+Python-based software packages such as {pyOpenMS}~\cite{Rost:2014},
+spectrum\_utils~\cite{Bittremieux:2020} or
+Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
+low level libraries and end-use software. \texttt{MSnbase} provides a
+flexible R~\cite{R} command line environment for metabolomics and
+proteomics mass spectrometry-based application. It allows a detailed
+step-by-step processing, analysis and exploration of the data and
+development of novel computational mass spectrometry methods.
 

### 2020-04-28 msnbase2.Rnw
various updates, mention ref from issue 1

@@
   manipulation, processing and visualisation of mass spectrometry
-  data. Here we present how the new \textit{on-disk} infrastructure
-  allows the handling of hundreds on commodity hardware and present
-  some application of the package.
+  data. We focus on the new \textit{on-disk} infrastructure, that
+  allows the handling of large raw mass spectrometry experiment on
+  commodity hardware and illustrate how the package is used for data
+  processing, method development and visualisation.
 \end{abstract}
@@
 knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
-library("xtable")
-library("microbenchmark")
 library("tidyverse")
@@
 
-This new on-disk infrastructure enables large scale data analyses
-using \texttt{MSnbase} (metabolomics example, see Johannes).
-
 \subsection{Prototyping}
@@
 
+We have presented here some some important functionality of the
+version 2 of the \texttt{MSnbase} package. The new on-disk
+infrastructure enables large scale data analyses~\cite{Nothias:2020}
+using \texttt{MSnbase} or other packages that rely on it such as
+\texttt{xcms}. We have also illustrated how \texttt{MSnbase} can be
+used for standard data analysis and visualisation, and how it can be
+used for method development and prototyping.
+
 The first public commit to the \texttt{MSnbase} GitHub repository was
@@
 dedicated manual page. In addition, the package offers 5 vignettes,
-including one aimed at developers. Question from users and developers
+including one aimed at developers. Questions from users and developers
 are answered on the Bioconductor support forum as well as on the
@@
 \Sexpr{v}. The main features presented here were available since
-version 2.0.
+version 2.0. The code to reproduce the analyses and figures in this
+article is available at
+\url{https://github.com/lgatto/2020-msnbase-v2/}.
 

### 2020-04-28 msnbase2.Rnw
Smaller fixes and reformulations

--- 2020-04-27 msnbase2.Rnw
@@
 \author{Johannes Rainer}
-\affiliation[EURAC]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}
+\affiliation[Eurac Research]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}
 
@@
 
-Mass spectrometry is a powerful technology to assays chemical and
+Mass spectrometry is a powerful technology to assay chemical and
 biological samples. It is used routinely with well characterised
@@
 
-There exists a very diverse catalogue of software tools to explore,
-process and interpret mass spectrometry data. These range from low
-level software libraries that are aimed at programmers to develop new
-applications, to user-oriented applications. Example of software
+There exists a very diverse catalogue of software tools to explore, process and
+interpret mass spectrometry data. These range from low level software libraries
+that are aimed at programmers to enable development of new applications, to more
+user-oriented applications with graphical user interfaces. Example of software
 libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}, that are used as application
-programming interfaces. ProteomeDiscoverer (Thermo Scientific),
-MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
-among the most widely user-centric application. They generally provide
-a limited and fixed set of functionality featuring often advanced
-graphical user interfaces.
+ProteoWizard~\cite{Chambers:2012}, that are used as application programming
+interfaces. ProteomeDiscoverer (Thermo Scientific), MaxQuant~\cite{Cox:2008} and
+PeptideShaker~\cite{Vaudel:2015} are among the most widely used user-centric
+applications. They generally provide a limited and fixed set of functionality
+featuring often advanced graphical user interfaces.
 
@@
 Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
-Python-based software packages such as {pyOpenMS}~\cite{Rost:2014},
+software packages such as Python-based {pyOpenMS}~\cite{Rost:2014},
 spectrum\_utils~\cite{Bittremieux:2020} or
@@
 flexible R~\cite{R} command line environment for metabolomics and
-proteomics mass spectrometry-based application. It allows a detailed
+proteomics mass spectrometry-based applications. It allows a detailed
 step-by-step processing, analysis and exploration of the data and
-development of novel computational mass spectrometry methods.
+development of novel computational mass spectrometry methods~\cite{Stanstrup:2019}.
 
@@
 
-In the following sections, we will describe demonstrate how
-\texttt{MSnbase} effichiently manages large raw mass spectrometry data
-and how it is used for data processing and visualisation. We will also
-illustrate how it makes use of the forward-pipe operator
-(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by
+In the following sections, we describe \texttt{MSnbase}'s ability to efficiently
+manage very large mass spectrometry data and experiments and how it can be used
+for data processing and visualisation. We will also illustrate how it makes use
+of the forward-pipe operator (\texttt{\%>\%}) defined in the \texttt{magrittr}
+package. This operator has proved useful to develop non-trivial analyses by
 combining individual functions into easily readable pipelines.
@@
 demonstrates how to create two \texttt{MSnExp} objects, tailored to
-manage mass spectrometry experiments, storing data in-memory or
+manage mass spectrometry experiments, storing data either in-memory or
 on-disk.
@@
 version of the package, that loads all the raw data and the metadata
-in memory. This solution doesn't scale for modern large dataset, and
+into memory. This solution doesn't scale for modern large dataset, and
 was complemented by the on-disk backend, that only loads metadata into
-memory and accesses the spectra in the original files when
+memory and accesses the spectra data (i.e. m/z and intensity values) in the original files when
 needed. There are two direct benefits using the on-disk backend,
-namely faster reading and reduced memory footpring. Figure
+namely faster reading and reduced memory footprint. Figure
 \ref{fig:bench} shows 5-fold faster reading times (a) and over a
@@
 \textit{lazy processing}. Operations on the raw data are stored in a
-processing queue and only effectively applied when raw data is
-accessed on disk. As an example, the following short analysis
-pipeline, that can equally be applied to on in-memory or on-disk data
-retains MS2 spectra acquired between 1000 and 3000 seconds, extract
+processing queue and only effectively applied when the user accesses m/z or
+intensity values. As an example, the following short analysis
+pipeline, that can equally be applied to in-memory or on-disk data,
+retains MS2 spectra acquired between 1000 and 3000 seconds, extracts
 the m/z range corresponding to the TMT 6-plex range and focuses on the
@@
 
-As shown on Figure~\ref{fig:bench} (c), this lazy mechanism si
+As shown on Figure~\ref{fig:bench} (c), this lazy mechanism is
 significantly faster than its application on in-memory data. The
 advantageous reading and execution times and memory footprint of the
-on-disk backend are possible by avoiding unnecessary access to the raw
+on-disk backend are possible by retrieving only spectra data from the selected
+subset hence avoiding access to the full raw
 data. Once access to the spectra m/z and intensity values become
@@
 becomes more efficient, as illustrated on Figure~\ref{fig:bench}
-(d). This gain is maximal when the whole dataset is the be accessed
-(i.e. all spectra are already in memory) and negligeable when large
-fractions of the data need to be subset.
+(d). This gain is maximal when the whole dataset is accessed
+(i.e. all spectra are already in memory) and negligible when only subsets of the
+data are requested.
 
-This new on-disk infrastructure enables large scale data analyses
-using \texttt{MSnbase} (metabolomics example, see Johannes).
+This new on-disk infrastructure enables analyses also of very large scale
+experiments using \texttt{MSnbase}). An example analysis of an intermediate
+scale metabolomics experiment (~2000 raw data files) using the
+\texttt{xcms}~\cite{Smith:2006} R package, that completely builds on the on-disk infrastructure
+of \texttt{MSnbase}, is described in [TODO: add citation once published.]
 

### 2020-04-28 msnbase2.tex

@@
 2.13.9. The main features presented here were available since
-version 2.0.
+version 2.0. The code to reproduce the analyses and figures in this
+article is available at
+\url{https://github.com/lgatto/2020-msnbase-v2/}.
 

### 2020-04-28 msnbase2.Rnw
Remove redundant reference to Nothias et al

@@
 
-This new on-disk infrastructure enables analyses also of very large scale
-experiments using \texttt{MSnbase}). An example analysis of an intermediate
-scale metabolomics experiment (~2000 raw data files) using the
-\texttt{xcms}~\cite{Smith:2006} R package, that completely builds on the on-disk infrastructure
-of \texttt{MSnbase}, is described in [TODO: add citation once published.]
-
 \subsection{Prototyping}

### 2020-04-29 msnbase2.Rnw
typos and some rephrasing

@@
 Mass spectrometry is a powerful technology to assay chemical and
-biological samples. It is used routinely with well characterised
+biological samples. It is used in routine with well characterised
 protocol, as well a development platform, to improve on existing
 methods and devise new ones. The complexity and diversity of mass
-spectrometry yields data that is itself complex and often times of
-considerable size, that requires non trivial processing before
+spectrometry yield data that is itself complex and often times of
+considerable size, that require non trivial processing before
 producing interpretable results. This is particularly relevant, and
-can constitute a significant challenge for method developers: in
+can constitute a significant challenge for method development: in
 addition to the development of sample processing and mass spectrometry
 methods, there is a need to process, analyse, interpret and assess
-these new data to demonstrate the improvement in their technical and
-analytical workflows.
+these new data to demonstrate the improvement in the technical,
+analytical and computational workflows.
 
-There exists a very diverse catalogue of software tools to explore, process and
-interpret mass spectrometry data. These range from low level software libraries
-that are aimed at programmers to enable development of new applications, to more
-user-oriented applications with graphical user interfaces. Example of software
-libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}, that are used as application programming
-interfaces. ProteomeDiscoverer (Thermo Scientific), MaxQuant~\cite{Cox:2008} and
-PeptideShaker~\cite{Vaudel:2015} are among the most widely used user-centric
-applications. They generally provide a limited and fixed set of functionality
-featuring often advanced graphical user interfaces.
+Practitioners have a diverse catalogue of software tools to explore,
+process and interpret mass spectrometry data at their disposal. These
+range from low level software libraries that are aimed at programmers
+to enable development of new applications, to more user-oriented
+applications with graphical user interfaces which provide a more
+limited set of functionalities to address a defined scope. Examples of
+software libraries include Java-based jmzML~\cite{Cote:2010} or
+C/C++-based ProteoWizard~\cite{Chambers:2012}. ProteomeDiscoverer
+(Thermo Scientific), MaxQuant~\cite{Cox:2008} and
+PeptideShaker~\cite{Vaudel:2015} are among the most widely used
+user-centric applications.
 
 In this software note, we present version 2 of the \texttt{MSnbase}
-\cite{Gatto:2012} software package, available from the
+\cite{Gatto:2012} software, available from the
 Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
-software packages such as Python-based {pyOpenMS}~\cite{Rost:2014},
+software such as Python-based {pyOpenMS}~\cite{Rost:2014},
 spectrum\_utils~\cite{Bittremieux:2020} or
 Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
-low level libraries and end-use software. \texttt{MSnbase} provides a
+low level libraries and end-user software. \texttt{MSnbase} provides a
 flexible R~\cite{R} command line environment for metabolomics and
@@
 step-by-step processing, analysis and exploration of the data and
-development of novel computational mass spectrometry methods~\cite{Stanstrup:2019}.
+development of novel computational mass spectrometry
+methods~\cite{Stanstrup:2019}.
 
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects, tailored to
-manage mass spectrometry experiments, storing data either in-memory or
-on-disk.
+demonstrates how to create two \texttt{MSnExp} objects storing data
+either in-memory or on-disk.
 
@@
   only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interations on in-memory and on-disk data
+  benchmark assessed over 10 interactions on in-memory and on-disk data
   containing 6103 MS2 spectra.  (d) Access time to spectra for the
@@
 into memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only loads metadata into
-memory and accesses the spectra data (i.e. m/z and intensity values) in the original files when
-needed. There are two direct benefits using the on-disk backend,
-namely faster reading and reduced memory footprint. Figure
-\ref{fig:bench} shows 5-fold faster reading times (a) and over a
-10-fold reduction in memory usage (b).
+was complemented by the on-disk backend. The on-disk backend loads the
+metadata into memory and accesses the spectra data (i.e. m/z and
+intensity values) in the original files only when needed. There are
+two direct benefits using the on-disk backend, namely faster reading
+and reduced memory footprint. Figure \ref{fig:bench} shows 5-fold
+faster reading times (a) and over a 10-fold reduction in memory usage
+(b).
 
@@
 \item Identify and filter the groups of spectra that represent
-  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)).. This can be
-  done using the `filterString` metadata variable that identifies
-  BoxCar spectra by their adjacent m/z segments with the
+  adjacent BoxCar acquisitions (Figure~\ref{fig:bc}~(b)). This can be
+  done using the \textit{filterString} metadata variable that
+  identifies BoxCar spectra by their adjacent m/z segments with the
   \texttt{bc\_groups()} function and filtering relevant spectra with
@@
 
-\item Remove any signal outside the BoxCar segnments using the
+\item Remove any signal outside the BoxCar segments using the
   \texttt{bc\_zero\_out\_box()} function from \texttt{MSnbaseBoxCar}
@@
 
-The R environment is well known for the quality of its visualiation
+The R environment is well known for the quality of its visualisation
 capacity. This also holds true for mass
@@
 the raw MS profile data for the serine and the same data after
-smoothing, centroiding and m/z refinement. Detailed execution and
-description of these operations can be found in the \emph{MSnbase:
-  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.
+smoothing, centroiding and m/z refinement, as illustrated in the code
+chunk below. Detailed execution and description of these operations
+can be found in the \emph{MSnbase: centroiding of profile-mode MS
+  data} \texttt{MSnbase} vignette.
 
@@
 
-We have presented here some some important functionality of the
-version 2 of the \texttt{MSnbase} package. The new on-disk
-infrastructure enables large scale data analyses~\cite{Nothias:2020}
-using \texttt{MSnbase} or other packages that rely on it such as
+We have presented here some some important functionality of
+\texttt{MSnbase} version 2. The new on-disk infrastructure enables
+large scale data analyses~\cite{Nothias:2020}, either using
+\texttt{MSnbase} directly or through packages that rely on it, such as
 \texttt{xcms}. We have also illustrated how \texttt{MSnbase} can be
@@
 particularly significant ones such as the on-disk backend described
-above. According to the package's Bioconductor page, there are 36
-Bioconductor packages that depend, import or suggest it. Among these
-are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
+herein. According to \texttt{MSnbase}'s Bioconductor page, there are
+36 Bioconductor packages that depend, import or suggest it. Among
+these are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
 spectrometry-based spatial proteomics data,
@@
 suitability monitoring and quality control of targeted proteomic
-experiments and the widely used \texttt{xcms}~\cite{Smith:2006} packge
+experiments and the widely used \texttt{xcms}~\cite{Smith:2006} package
 for the processing and analysis of metabolomics data. \texttt{MSnbase}
@@
 \textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
-raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantiative
+raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantitative
 data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under

### 2020-04-29 msnbase2.tex
typos and some rephrasing

@@
 
-Mass spectrometry is a powerful technology to assays chemical and
-biological samples. It is used routinely with well characterised
+Mass spectrometry is a powerful technology to assay chemical and
+biological samples. It is used in routine with well characterised
 protocol, as well a development platform, to improve on existing
 methods and devise new ones. The complexity and diversity of mass
-spectrometry yields data that is itself complex and often times of
-considerable size, that requires non trivial processing before
+spectrometry yield data that is itself complex and often times of
+considerable size, that require non trivial processing before
 producing interpretable results. This is particularly relevant, and
-can constitute a significant challenge for method developers: in
+can constitute a significant challenge for method development: in
 addition to the development of sample processing and mass spectrometry
 methods, there is a need to process, analyse, interpret and assess
-these new data to demonstrate the improvement in their technical and
-analytical workflows.
+these new data to demonstrate the improvement in the technical,
+analytical and computational workflows.
 
-There exists a very diverse catalogue of software tools to explore,
-process and interpret mass spectrometry data. These range from low
-level software libraries that are aimed at programmers to develop new
-applications, to user-oriented applications. Example of software
-libraries include Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}, that are used as application
-programming interfaces. ProteomeDiscoverer (Thermo Scientific),
-MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
-among the most widely user-centric application. They generally provide
-a limited and fixed set of functionality featuring often advanced
-graphical user interfaces.
+Practitioners have a diverse catalogue of software tools to explore,
+process and interpret mass spectrometry data at their disposal. These
+range from low level software libraries that are aimed at programmers
+to enable development of new applications, to more user-oriented
+applications with graphical user interfaces which provide a more
+limited set of functionalities to address a defined scope. Examples of
+software libraries include Java-based jmzML~\cite{Cote:2010} or
+C/C++-based ProteoWizard~\cite{Chambers:2012}. ProteomeDiscoverer
+(Thermo Scientific), MaxQuant~\cite{Cox:2008} and
+PeptideShaker~\cite{Vaudel:2015} are among the most widely used
+user-centric applications.
 
 In this software note, we present version 2 of the \texttt{MSnbase}
-\cite{Gatto:2012} software package, available from the
+\cite{Gatto:2012} software, available from the
 Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
-Python-based software packages such as {pyOpenMS}~\cite{Rost:2014},
+software such as Python-based {pyOpenMS}~\cite{Rost:2014},
 spectrum\_utils~\cite{Bittremieux:2020} or
 Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
-low level libraries and end-use software. \texttt{MSnbase} provides a
+low level libraries and end-user software. \texttt{MSnbase} provides a
 flexible R~\cite{R} command line environment for metabolomics and
-proteomics mass spectrometry-based application. It allows a detailed
+proteomics mass spectrometry-based applications. It allows a detailed
 step-by-step processing, analysis and exploration of the data and
-development of novel computational mass spectrometry methods.
+development of novel computational mass spectrometry
+methods~\cite{Stanstrup:2019}.
 
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects, tailored to
-manage mass spectrometry experiments, storing data in-memory or
-on-disk.
+demonstrates how to create two \texttt{MSnExp} objects storing data
+either in-memory or on-disk.
 
@@
 version of the package, that loads all the raw data and the metadata
-in memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend, that only loads metadata into
-memory and accesses the spectra in the original files when
-needed. There are two direct benefits using the on-disk backend,
-namely faster reading and reduced memory footpring. Figure
-\ref{fig:bench} shows 5-fold faster reading times (a) and over a
-10-fold reduction in memory usage (b).
+into memory. This solution doesn't scale for modern large dataset, and
+was complemented by the on-disk backend. The on-disk backend loads the
+metadata into memory and accesses the spectra data (i.e. m/z and
+intensity values) in the original files only when needed. There are
+two direct benefits using the on-disk backend, namely faster reading
+and reduced memory footprint. Figure \ref{fig:bench} shows 5-fold
+faster reading times (a) and over a 10-fold reduction in memory usage
+(b).
 
@@
 becomes more efficient, as illustrated on Figure~\ref{fig:bench}
-(d). This gain is maximal when the whole dataset is the be accessed
-(i.e. all spectra are already in memory) and negligeable when large
-fractions of the data need to be subset.
+(d). This gain is maximal when the whole dataset is accessed
+(i.e. all spectra are already in memory) and negligible when only subsets of the
+data are requested.
 

### 2020-04-29 msnbase2.Rnw
last updates before submission

@@
 \title[MSnbase version 2]
-  {\texttt{MSnbase}, efficient R-based processing and visualisation of raw mass spectrometry data}
+  {\texttt{MSnbase}, efficient and elegant R-based processing and
+    visualisation of raw mass spectrometry data}
 
@@
   allows the handling of large raw mass spectrometry experiment on
-  commodity hardware and illustrate how the package is used for data
-  processing, method development and visualisation.
+  commodity hardware and illustrate how the package is used for
+  elegant data processing, method development and visualisation.
 \end{abstract}

### 2020-04-29 msnbase2.tex
last updates before submission

@@
 The version of \texttt{MSnbase} used in this manuscript is
-2.13.9. The main features presented here were available since
+2.14.0. The main features presented here were available since
 version 2.0. The code to reproduce the analyses and figures in this

### 2020-04-30 msnbase2.Rnw
add keywords
mention mzR - closes #6

@@
 \abbreviations{}
-\keywords{Bioconductor, mass spectrometry, software, reproducible research} %% up to 10 keywords
+\keywords{Bioconductor, mass spectrometry, software, metabolomics, proteomics, visualisation, reproducible research} %% up to 10 keywords
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 demonstrates how to create two \texttt{MSnExp} objects storing data
-either in-memory or on-disk.
+either in-memory or on-disk. Both modes rely on the
+\texttt{mzR}~\cite{Chambers:2012} package to access the data in the
+raw data files.
 

### 2020-04-30 msnbase2.Rnw
mention testing - closes #7

@@
 dedicated manual page. In addition, the package offers 5 vignettes,
-including one aimed at developers. Questions from users and developers
-are answered on the Bioconductor support forum as well as on the
-package GitHub page. The package provides several sample and
-benchmarking datasets, and relies on other dedicated
-\textit{experiment packages} such as \texttt{msdata}~\cite{msdata} for
-raw data or \texttt{pRolocdata}~\cite{Gatto:2014a} for quantitative
+including one aimed at developers. The package is checked nightly on
+the Bioconductor servers: it implements unit tests covering 72\% of
+the code base and, through its vignettes, also provides integration
+testing. Questions from users and developers are answered on the
+Bioconductor support forum as well as on the package GitHub page. The
+package provides several sample and benchmarking datasets, and relies
+on other dedicated \textit{experiment packages} such as
+\texttt{msdata}~\cite{msdata} for raw data or
+\texttt{pRolocdata}~\cite{Gatto:2014a} for quantitative
 data. \texttt{MSnbase} is available on Windows, Mac OS and Linux under

### 2020-05-07 msnbase2.Rnw
add specs - closes #5
fix: number of processors
fix typo in code

@@
   only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interactions on in-memory and on-disk data
-  containing 6103 MS2 spectra.  (d) Access time to spectra for the
-  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
-  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
-  red. }
+  benchmark assessed over 10 interactions on in-memory and on-disk
+  data containing 6103 MS2 spectra.  (d) Access time to spectra for
+  the in-memory (left) and on-disk (right) backends for 1, 10, 100
+  1000, 5000 and all 6103 spectra. On-disk backend: blue. In-memory
+  backend: red. Benchmarks were performed on a Dell XPS laptop with an
+  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads) and 7.5 GiB RAM
+  running Ubuntu 18.04.4 LTS 64-bit. }
 \label{fig:bench}
@@
       combineSpectra(fcol = "bc_groups", ## reconstruct full spectrum
-                     method = boxcarCompbine)
+                     method = boxcarCombine)
 @

### 2020-05-07 msnbase2.Rnw

@@
   backend: red. Benchmarks were performed on a Dell XPS laptop with an
-  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads) and 7.5 GiB RAM
+  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads) and 7.5 GB RAM
   running Ubuntu 18.04.4 LTS 64-bit. }

### 2020-05-07 msnbase2.Rnw
fix serine example code

@@
     pickPeaks(refineMz = "descendPeak") %>%
-    filterRt(c(serine_mz - 0.01, serine_mz + 0.01)) %>%
-    filterMz(c(175, 187))
+    filterMz(c(serine_mz - 0.01, serine_mz + 0.01)) %>%
+    filterRt(c(175, 187))
 @

### 2020-05-09 msnbase2.Rnw
use [s] - close #13

@@
     scale_x_continuous(breaks=c(1, 5, 10)) + 
-    ylab("Reading time (s)") + 
+    ylab("Reading time [s]") + 
     xlab("Number of files") +

### 2020-05-09 msnbase2.Rnw
use [MB] - close #13

@@
     scale_x_continuous(breaks=c(1, 5, 10)) + 
-    ylab("Object size (MB)") +
+    ylab("Object size [MB]") +
     xlab("Number of files") +

### 2020-05-09 msnbase2.Rnw
explain a bit more mzR usage for backends - close #9

--- 2020-05-09 msnbase2.Rnw
@@
 demonstrates how to create two \texttt{MSnExp} objects storing data
-either in-memory or on-disk. Both modes rely on the
-\texttt{mzR}~\cite{Chambers:2012} package to access the data in the
-raw data files.
+either in-memory or on-disk.
 
@@
 
-The former is the legacy storage mode, implemented in the first
+Both modes rely on the \texttt{mzR}~\cite{Chambers:2012} package to
+access the spectra (using the \texttt{mzR::peaks()} function) and the
+metadata (using the \texttt{mzR::header()} function) in the data
+files. The former is the legacy storage mode, implemented in the first
 version of the package, that loads all the raw data and the metadata
-into memory. This solution doesn't scale for modern large dataset, and
-was complemented by the on-disk backend. The on-disk backend loads the
-metadata into memory and accesses the spectra data (i.e. m/z and
-intensity values) in the original files only when needed. There are
-two direct benefits using the on-disk backend, namely faster reading
-and reduced memory footprint. Figure \ref{fig:bench} shows 5-fold
-faster reading times (a) and over a 10-fold reduction in memory usage
-(b).
+into memory upon creation of the in-memory \texttt{MSnExp}
+object. This solution doesn't scale for modern large dataset, and was
+complemented by the on-disk backend. The on-disk backend only loads
+the metadata into memory when the on-disk \texttt{MSnExp} is created
+and accesses the spectra data (i.e. m/z and intensity values) in the
+original files on disk only when needed (see below and Figure
+\ref{fig:bench} (d)), such as for example for plotting. There are two
+direct benefits using the on-disk backend, namely faster reading and
+reduced memory footprint. Figure \ref{fig:bench} shows 5-fold faster
+reading times (a) and over a 10-fold reduction in memory usage (b).
 
 The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing}. Operations on the raw data are stored in a
-processing queue and only effectively applied when the user accesses m/z or
-intensity values. As an example, the following short analysis
-pipeline, that can equally be applied to in-memory or on-disk data,
-retains MS2 spectra acquired between 1000 and 3000 seconds, extracts
-the m/z range corresponding to the TMT 6-plex range and focuses on the
-MS2 spectra with a precursor intensity greater than $11 \times 10^6$
-(the median precursor intensity).
+\textit{lazy processing} in the object itself. Operations on the raw
+data are stored in a processing queue and only effectively applied
+when the user accesses m/z or intensity values. As an example, the
+following short analysis pipeline, that can equally be applied to
+in-memory or on-disk data, retains MS2 spectra acquired between 1000
+and 3000 seconds, extracts the m/z range corresponding to the TMT
+6-plex range and focuses on the MS2 spectra with a precursor intensity
+greater than $11 \times 10^6$ (the median precursor intensity).
 

### 2020-05-09 msnbase2.tex
add specs - closes #5
rebuild pdf
fix typo in code
build with fix serine example code
explain a bit more mzR usage for backends - close #9

@@
   only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interactions on in-memory and on-disk data
-  containing 6103 MS2 spectra.  (d) Access time to spectra for the
-  in-memory (left) and on-disk (right) backends for 1, 10, 100 1000,
-  5000 and all 6103 spectra. On-disk backend: blue. In-memory backend:
-  red. }
+  benchmark assessed over 10 interactions on in-memory and on-disk
+  data containing 6103 MS2 spectra.  (d) Access time to spectra for
+  the in-memory (left) and on-disk (right) backends for 1, 10, 100
+  1000, 5000 and all 6103 spectra. On-disk backend: blue. In-memory
+  backend: red. Benchmarks were performed on a Dell XPS laptop with an
+  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads) and 7.5 GB RAM
+  running Ubuntu 18.04.4 LTS 64-bit. }
 \label{fig:bench}
@@
       \hlkwd{combineSpectra}\hlstd{(}\hlkwc{fcol} \hlstd{=} \hlstr{"bc_groups"}\hlstd{,} \hlcom{## reconstruct full spectrum}
-                     \hlkwc{method} \hlstd{= boxcarCompbine)}
+                     \hlkwc{method} \hlstd{= boxcarCombine)}
 \end{alltt}
@@
     \hlkwd{pickPeaks}\hlstd{(}\hlkwc{refineMz} \hlstd{=} \hlstr{"descendPeak"}\hlstd{)} \hlopt{%>%}
-    \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(serine_mz} \hlopt{-} \hlnum{0.01}\hlstd{, serine_mz} \hlopt{+} \hlnum{0.01}\hlstd{))} \hlopt{%>%}
-    \hlkwd{filterMz}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{175}\hlstd{,} \hlnum{187}\hlstd{))}
+    \hlkwd{filterMz}\hlstd{(}\hlkwd{c}\hlstd{(serine_mz} \hlopt{-} \hlnum{0.01}\hlstd{, serine_mz} \hlopt{+} \hlnum{0.01}\hlstd{))} \hlopt{%>%}
+    \hlkwd{filterRt}\hlstd{(}\hlkwd{c}\hlstd{(}\hlnum{175}\hlstd{,} \hlnum{187}\hlstd{))}
 \end{alltt}

### 2020-05-09 msnbase2.Rnw
mention what to do after boxcar
provide more general info and clarifications about the package

@@
 
-In this software note, we present version 2 of the \texttt{MSnbase}
-\cite{Gatto:2012} software, available from the
-Bioconductor~\cite{Huber:2015} project. \texttt{MSnbase}, like other
+In this software note, we present version 2 of the
+\texttt{MSnbase}~\cite{Gatto:2012} software, available from the
+Bioconductor~\cite{Huber:2015} project. The package, like other
 software such as Python-based {pyOpenMS}~\cite{Rost:2014},
@@
 low level libraries and end-user software. \texttt{MSnbase} provides a
-flexible R~\cite{R} command line environment for metabolomics and
-proteomics mass spectrometry-based applications. It allows a detailed
-step-by-step processing, analysis and exploration of the data and
-development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}.
+flexible R~\cite{R} command-line environment for metabolomics and
+proteomics mass spectrometry-based applications. It lays out a sound
+infrastructure to work with raw mass spectrometry, quantitative and
+proteomics identification data. The package enables manipulation of
+these data (for example subsetting, filtering, or accessing specific
+parts thereof), a detailed step-by-step processing (for example
+smoothing and centroiding of raw data, or normalisation and imputation
+of quantitative data), analysis and visualisation of these data and
+the development of novel computational mass spectrometry
+methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
+pertaining to raw mass spectrometry data handling and processing.
 
-\section{Software functionality}
+\section{Infrastructure for raw data}
 
@@
 
-In the following sections, we describe \texttt{MSnbase}'s ability to efficiently
-manage very large mass spectrometry data and experiments and how it can be used
-for data processing and visualisation. We will also illustrate how it makes use
-of the forward-pipe operator (\texttt{\%>\%}) defined in the \texttt{magrittr}
-package. This operator has proved useful to develop non-trivial analyses by
+In the following sections, we describe \texttt{MSnbase}'s ability to
+efficiently handle very large mass spectrometry data and experiments
+and how it can be used for data processing and visualisation. We will
+also illustrate how it makes use of the forward-pipe operator
+(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
+operator has proved useful to develop non-trivial analyses by
 combining individual functions into easily readable pipelines.
@@
 
+After processing of the BoxCar data, the final object can either be
+further analysed in R or written back to disk as an \texttt{mzML} file
+using \texttt{writeMSData()} for processing with other tools.
+
 \begin{figure}[p]

### 2020-05-09 msnbase2.tex
mention what to do after boxcar
provide more general info and clarifications about the package

@@
 \end{knitrout}
+
+After processing of the BoxCar data, the final object can either be
+further analysed in R or written back to disk as an \texttt{mzML} file
+using \texttt{writeMSData()} for processing with other tools.
 

### 2020-05-09 msnbase2.Rnw
some last rephrasing

@@
 infrastructure to work with raw mass spectrometry, quantitative and
-proteomics identification data. The package enables manipulation of
-these data (for example subsetting, filtering, or accessing specific
-parts thereof), a detailed step-by-step processing (for example
-smoothing and centroiding of raw data, or normalisation and imputation
-of quantitative data), analysis and visualisation of these data and
-the development of novel computational mass spectrometry
+proteomics identification data. The package enables manipulation (for
+example subsetting, filtering, or accessing specific parts thereof),
+detailed step-by-step processing (for example smoothing and
+centroiding of raw data, or normalisation and imputation of
+quantitative data), analysis and visualisation of these data and the
+development of novel computational mass spectrometry
 methods~\cite{Stanstrup:2019}. Here, we focus on the new developments

### 2020-06-23 msnbase2.Rnw
Add description of raw data
This should address point 1 of the first reviewer and fourth reviewer.

@@
 proteomics mass spectrometry-based applications. It lays out a sound
-infrastructure to work with raw mass spectrometry, quantitative and
+infrastructure to work with raw mass spectrometry \color{red}{data from MS files
+  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
 proteomics identification data. The package enables manipulation (for
@@
 detailed step-by-step processing (for example smoothing and
-centroiding of raw data, or normalisation and imputation of
+centroiding of \sout{raw} \color{red}{profile-mode MS} data, or normalisation and imputation of
 quantitative data), analysis and visualisation of these data and the
@@
 level, acquisition number, retention time, precursor m/z and intensity
-(for MS level 2 and above), and many more. \texttt{MSnbase} provides a
+(for MS level 2 and above), and many more. \texttt{MSnbase}\color{red}{
+  ,\textit{via} the \texttt{mzR} package~\cite{Chambers:2012}, allows import of raw mass spectrometry
+  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and} provides a
 rich interface to manipulate such objects. The code chunk below
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects storing data
+demonstrates how to create two \texttt{MSnExp} objects \color{red}{importing
+  data from an mzML file and} storing \sout{data}\color{red}{it} 
 either in-memory or on-disk.
@@
 The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing} in the object itself. Operations on the raw
+\textit{lazy processing} in the object itself. Operations on the \sout{raw}\color{red}{MS}
 data are stored in a processing queue and only effectively applied
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data for the serine and the same data after
+the raw MS profile data \color{red}{imported from an mzML file} for \sout{the} serine and the same data after
 smoothing, centroiding and m/z refinement, as illustrated in the code

### 2020-06-24 msnbase2.Rnw
Expand explanation of lazy processing queue

@@
 
-The on-disk backend also offers efficient data manipulation by way of
+\color{red}{Because the on-disk backend does not hold all the data in memory,
+  direct data manipulations are not possible. We thus implemented a \textit{lazy
+  processing} mechanism in this backend that caches any data manipulation
+operations on m/z and intensity values}
+\sout{The on-disk backend also offers efficient data manipulation by way of
 \textit{lazy processing} in the object itself. Operations on the \sout{raw}\color{red}{MS}
-data are stored in a processing queue and only effectively applied
-when the user accesses m/z or intensity values. As an example, the
+data are stored} in a processing queue and only effectively
+\sout{applied}\color{red}{applies them}
+when the user accesses m/z or intensity values. \color{red}{As an additional
+  advantage, operations on subsets of the data become much faster since data
+  manipulations are applied only to data subsets instead of the full data set
+  at once. Also, data access in on-disk backends is parallelized by data file
+  ensuring a higher performance of this backend over conventional in-memory data
+  representations.} 
+As an example, the
 following short analysis pipeline, that can equally be applied to

### 2020-06-24 msnbase2.Rnw
fix: name of PD

--- 2020-06-23 msnbase2.Rnw
@@
 software libraries include Java-based jmzML~\cite{Cote:2010} or
-C/C++-based ProteoWizard~\cite{Chambers:2012}. ProteomeDiscoverer
-(Thermo Scientific), MaxQuant~\cite{Cox:2008} and
-PeptideShaker~\cite{Vaudel:2015} are among the most widely used
-user-centric applications.
+C/C++-based ProteoWizard~\cite{Chambers:2012}.
+\sout{ProteomeDiscoverer (Thermo Scientific)}
+\red{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
+MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
+most widely used user-centric applications.
 

### 2020-06-24 msnbase2.Rnw
fix: remove duplicated some
Amend according to @sgibb's suggestion

--- 2020-06-24 msnbase2.Rnw
@@
   processing} mechanism in this backend that caches any data manipulation
-operations on m/z and intensity values}
+operation in a processing queue in the object itself. Only when the user
+accesses m/z or intensity values these operations are applied}.
 \sout{The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing} in the object itself. Operations on the \sout{raw}\color{red}{MS}
-data are stored} in a processing queue and only effectively
-\sout{applied}\color{red}{applies them}
+\textit{lazy processing} in the object itself. Operations on the raw
+data are stored in a processing queue and only effectively applied
 when the user accesses m/z or intensity values. \color{red}{As an additional

### 2020-06-25 large_experiment_processing.Rmd
feat: add supplementary data showcasing MSnbase on a large data set

---
title: "Supplementary data for: \\textit{`MSnbase`, efficient and elegant R-based processing and visualisation of raw mass spectrometry data}"
author: "Laurent Gatto, Sebastian Gibb and Johannes Rainer"
output:
    BiocStyle::pdf_document
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document describes handling of mass spectrometry data from large
experiments using the `MSnbase` package and more specifically its *on-disk*
backend. For demonstration purposes, the
[MassIVE](https://massive.ucsd.edu/ProteoSAFe/static/massive.jsp) data set
[MSV000080030](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=5e7034cc98c54a47b803b144bff6a296)
is used. This consists of over 1,000 mzXML files from swab-samples collected
from hands and various personal objects of 80 volunteers.

# Data handling and analysis with `MSnbase`

In this section we demonstrate data handling and access by `MSnbase` on a large
experiment consisting of more than 1,000 data files.

To reproduce the analysis in this document, download the *MSV000080030* folder
from
[ftp://massive.ucsd.edu/MSV000080030/](ftp://massive.ucsd.edu/MSV000080030/) and
place it into the same folder than this document.

Below we load the required libraries and define the files to be analyzed.

```{r}
library(MSnbase)
library(magrittr)
library(pryr)

fls <- dir("MSV000080030/ccms_peak/Forensic_study_80_volunteers/",
           pattern = "mzXML", full.names = TRUE, recursive = TRUE)
```

The data set consists of in total `r length(fls)` mzXML files. We next load the
data using the two different `MSnbase` backends `"inMemory`" and `"onDisk"`. For
the in-memory backend, due to the larger memory requirements, we import the data
only from a subset of the files.

```{r import-inmem, eval = !file.exists("ms_mem_hand.RData")}
ms_mem <- readMSData(fls[grep("Hand", fls)], mode = "inMemory")
```

```{r import-inmem-save, echo = FALSE}
if (!file.exists("ms_mem_hand.RData")) {
    save(ms_mem, file = "ms_mem_hand.RData")
} else {
    load("ms_mem_hand.RData")
}
```

Next we load data from all mzXML files as an on-disk `MSnExp` object.

```{r import-ondisk, eval = !file.exists("ms_dsk_all.RData")}
ms_dsk <- readMSData(fls, mode = "onDisk")
```

```{r import-ondisk-save, echo = FALSE}
if (!file.exists("ms_dsk_all.RData")) {
    save(ms_dsk, file = "ms_dsk_all.RData")
} else {
    load("ms_dsk_all.RData")
}
```

Below we count the number of spectra per MS level of the whole experiment.

```{r ondisk-spectra-table}
table(msLevel(ms_dsk))
```

Note that the in-memory `MSnExp` object contains only MS2 spectra (in total 
`r length(ms_mem)`) from a subset of data files, still, data import was much 
slower (over ~ 12 hours for the in-memory backend while creating the on-disk 
object from the full data data set took ~ 3 hours).

Next we subset the on-disk object to contain the same set of spectra than the
in-memory `MSnExp` and compare their memory footprint.

```{r memory-footprint}
ms_dsk_hands <- ms_dsk %>%
    filterFile(grep("Hand", fls)) %>%
    filterMsLevel(2L)

object_size(ms_mem)
object_size(ms_dsk_hands)
```

Since the on-disk object stores only spectra metadata in memory it occupies also
much less system memory. As a comparison, the on-disk `MSnExp` for the full
experiment was still much smaller than the in-memory object:

```{r memory-footprint-all}
object_size(ms_dsk)
```

## Performance of the *on-disk* backend on large scale data sets

To demonstrate `MSnbase`'s efficiency in processing large scale experiments we
perform some standard subsetting, data access and manipulation operations.

```{r parallel-setup, echo = FALSE}
library(BiocParallel)
register(SerialParam())
```

We first compare the performance of the on-disk and in-memory backend on
accessing m/z values with the `mz` function on a set of 100 randomly selected
spectra. The performance is assessed with the `microbenchmark` function.

```{r random-mz}
set.seed(123)
idx <- sample(seq_along(ms_mem), 100)

library(microbenchmark)
microbenchmark(mz(ms_mem[idx]),
               mz(ms_dsk_hands[idx]),
               times = 5)
```

Thus, for this combined subsetting and data access operation the on-disk backend
performed better than the in-memory `MSnExp`, while even requiring much less
memory.

Next we extract all MS2 spectra with a retention time between 50 and 60 seconds
and a precursor m/z of 108.5362 (+/- 5ppm). This subsetting operation is
performed on the on-disk `MSnExp` object representing the full experiment with
the `r length(fileNames(ms_dsk))` data files/samples. To assess the performance
of the following operations we use `system.time` calls that record elapsed time
in seconds.

```{r precursor-selection}
system.time(
    ms_sub <- ms_dsk %>%
        filterMsLevel(2L) %>%
        filterRt(c(50, 60)) %>%
        filterPrecursorMz(mz = 108.5362, ppm = 5)
)["elapsed"]
```

In total `length(ms_sub)` spectra were selected from in total 
`r length(unique(fromFile(ms_sub)))` data files/samples. The plot below shows 
the data for the first spectrum.

```{r precursor-selection-plot-1}
system.time(
    plot(ms_sub[[1]])
)["elapsed"]
```

Since there seems to be quite some background noise in the MS2 spectrum we next
remove peaks with an intensity below 50 by first replacing their intensities
with 0 (with the `removePeaks` call) and subsequently removing all 0-intensity
peaks from each spectrum with the `clean` call. In addition we *normalize* each
spectrum by dividing the maximum intensity per spectrum from the spectrum's
intensities.

```{r normalize}
system.time(
    ms_sub <- ms_sub %>%
        removePeaks(t = 50) %>%
        clean(all = TRUE) %>%
        normalize(method = "max")
)["elapsed"]
```

The result on the first spectrum is shown below.

```{r precursor-selection-plot-normalized}
system.time(
    plot(ms_sub[[1]])
)["elapsed"]
```

Note that any of the data manipulations above are not directly applied to the
data but *cached* in the object's internal *lazy processing queue* (explaining
the very short running time of the normalization call). The operations are only
effectively applied to the data when m/z or intensity values are extracted from
the object as it is the case in the `plot` call above.

For additional workflows employing `MSnbase` see also
[metabolomics2018](https://jorainer.github.io/metabolomics2018/xcms-preprocessing.html)[^1]
that explains filtering, plotting and centroiding of profile-mode MS data with
`MSnbase` and subsequent pre-processing of the (label free/untargeted) LC-MS
data with the `xcms` package (that builds upon `MSnbase` for MS data
representation and access).

## System information

The present analysis was run on a MacBook Pro 16,1 with 2.3 GHz 8-Core Intel
Core i9 CPU and 64 GB 2667 MHz DDR4 memory running macOS version 10.15.5. The R
version and the version of the used packages are listed below.

```{r session-info}
sessionInfo()
```

[^1]: https://github.com/jorainer/metabolomics2018

### 2020-06-26 msnbase2.Rnw
Mention and refer to the Supplementary Information in the main text

--- 2020-06-24 msnbase2.Rnw
@@
 
-In the following sections, we describe \texttt{MSnbase}'s ability to
-efficiently handle very large mass spectrometry data and experiments
-and how it can be used for data processing and visualisation. We will
-also illustrate how it makes use of the forward-pipe operator
-(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by
-combining individual functions into easily readable pipelines.
+In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
+  efficiently handle very large mass spectrometry data and experiments and how
+  it} \code{red}{how \texttt{MSnbase}} can be used for data processing and
+visualisation. We will also illustrate how it makes use of the forward-pipe
+operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
+operator has proved useful to develop non-trivial analyses by combining
+individual functions into easily readable pipelines. \code{red}{An example for
+  \texttt{MSnbase}'s ability to efficiently handle also very large mass
+  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
+  files) is provided as Supplementary Information.}
 
@@
 
-We have presented here some some important functionality of
+We have presented here some important functionality of
 \texttt{MSnbase} version 2. The new on-disk infrastructure enables

### 2020-06-26 msnbase2.Rnw
Add reference to documentation and workflows

--- 2020-06-24 msnbase2.Rnw
@@
 development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
+methods~\cite{Stanstrup:2019}. \color{red}{Extensive documentation and use cases 
+are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
+workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
 pertaining to raw mass spectrometry data handling and processing.

### 2020-06-26 large_experiment_processing.Rmd
Add section with general usage

@@
 
+## Basic MS data access functionality
+
+Before evaluating the `MSnbase` performance on the large data set we provide
+first some general description of the `MSnbase` data classes in `MSnbase` basic
+data access operations. MS data from raw data files in mzML, mzXML, mzData or
+netCDF format is represented by the `MSnExp` object which organizes the spectra
+from the original files in a one-dimensional list. Functions like `rtime` and
+`msLevel` allow to extract the retention time and MS level, respectively, for
+each spectrum and return a `numeric` (or `integer`) vector with the same length
+than the number of spectra in the `MSnExp`. In the example below we use the
+`rtime` function to extract the retention times for each spectrum.
+
+```{r rtime}
+rts <- rtime(ms_dsk)
+length(rts)
+head(rts)
+```
+
+The `fromFile` function can be used to determine from which file (sample) a
+specific spectrum in the `MSnExp` object is. This function returns an `integer`
+vector, same length than spectra in the experiment, with the file index; the
+file names can be accessed with the `fileNames` method. A `MSnExp` object can be
+subsetted with `[` and e.g. the index of the spectra that should be retained. In
+the code block below we subset our `ms_dsk` object to contain only spectra from
+the 3rd file.
+
+```{r subset}
+one_file <- ms_dsk[fromFile(ms_dsk) == 3]
+length(one_file)
+```
+
+Note however that there are also dedicated *filter* functions to subset an
+`MSnExp` object such as `filterFile`, `filterMsLevel`, `filterRt`, `filterMz`,
+`filterPrecursorMz` or `filterIsolationWindow`. In the example below we use the
+`filterRt` function to further subset our data to contain only spectra acquired
+within a certain time range.
+
+```{r filterrt}
+one_file <- filterRt(one_file, rt = c(40, 300))
+length(one_file)
+```
+
+As mentioned above, the `MSnExp` object is comparable with a list of
+spectra. Thus, to extract a single spectrum from it we can use `[[`. This will
+return an object of type `Spectrum` which encapsules/represents all information
+of the measured spectrum (i.e. m/z and intensity values as well as metadata
+information). In the example below we extract the 15th spectrum from our data
+subset and access its m/z values with the `mz` function.
+
+```{r}
+sp <- one_file[[15]]
+mz(sp)
+```
+
+This particular spectrum has thus only 3 peaks.
+
+Note that m/z or intensity values can also be directly extracted from the
+`MSnExp` object as shown in the example below. The result will be a `list` of
+`numeric` vectors, each element representing the m/z values for each spectrum in
+the object.
+
+```{r}
+mzs <- mz(one_file)
+class(mzs)
+length(mzs)
+```
+
+In addition, it is also possible to extract all m/z and intensity values from an
+`MSnExp` object as a `data.frame` as shown in the code block below, but this is
+not suggested, since it loads all the data into memory and all MS spectrum
+metadata such as MS level or precursor m/z get lost.
+
+```{r data.frame}
+df <- as(one_file, "data.frame")
+head(df)
+nrow(df)
+```
+
+Note that for all these operations it is irrelevant whether an in-memory or
+on-disk backend was used. In general it is advisable to use the on-disk backend
+especially for experiments with more than ~ 50 files.
+
 ## Performance of the *on-disk* backend on large scale data sets

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 
-To reproduce the analysis in this document, download the *MSV000080030* folder
+To reproduce the analysis described in this document, download the *MSV000080030* folder
 from

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

--- 2020-06-26 large_experiment_processing.Rmd
@@
 
-To reproduce the analysis in this document, download the *MSV000080030* folder
+To reproduce the analysis described in this document, download the *MSV000080030* folder
 from
 [ftp://massive.ucsd.edu/MSV000080030/](ftp://massive.ucsd.edu/MSV000080030/) and
-place it into the same folder than this document.
+place it into the same folder as this document.
 

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 
-The data set consists of in total `r length(fls)` mzXML files. We next load the
+The data set consists of `r length(fls)` mzXML files. We next load the
 data using the two different `MSnbase` backends `"inMemory`" and `"onDisk"`. For
@@
 Note that the in-memory `MSnExp` object contains only MS2 spectra (in total 
-`r length(ms_mem)`) from a subset of data files, still, data import was much 
+`r length(ms_mem)`) from a subset of data files. However, the data import was much 
 slower (over ~ 12 hours for the in-memory backend while creating the on-disk 

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 
-Next we subset the on-disk object to contain the same set of spectra than the
+Next we subset the on-disk object to contain the same set of spectra as the
 in-memory `MSnExp` and compare their memory footprint.
@@
 Before evaluating the `MSnbase` performance on the large data set we provide
-first some general description of the `MSnbase` data classes in `MSnbase` basic
+some general description of the `MSnbase` data classes and basic
 data access operations. MS data from raw data files in mzML, mzXML, mzData or

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 netCDF format is represented by the `MSnExp` object which organizes the spectra
-from the original files in a one-dimensional list. Functions like `rtime` and
+from the original files in an one-dimensional list. Functions like `rtime` and
 `msLevel` allow to extract the retention time and MS level, respectively, for

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 from the original files in an one-dimensional list. Functions like `rtime` and
-`msLevel` allow to extract the retention time and MS level, respectively, for
+`msLevel` allow to extract the retention time and MS level, respectively.
 each spectrum and return a `numeric` (or `integer`) vector with the same length

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 `msLevel` allow to extract the retention time and MS level, respectively.
-each spectrum and return a `numeric` (or `integer`) vector with the same length
+They return a `numeric` (or `integer`) vector with the same length
 than the number of spectra in the `MSnExp`. In the example below we use the
@@
 
-This particular spectrum has thus only 3 peaks.
+This particular spectrum has only 3 peaks.
 

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 They return a `numeric` (or `integer`) vector with the same length
-than the number of spectra in the `MSnExp`. In the example below we use the
+as the number of spectra in the `MSnExp`. In the example below we use the
 `rtime` function to extract the retention times for each spectrum.
@@
 `MSnExp` object as a `data.frame` as shown in the code block below, but this is
-not suggested, since it loads all the data into memory and all MS spectrum
+not suggested, since it loads all the data into memory but all MS spectrum
 metadata such as MS level or precursor m/z get lost.

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 
-Thus, for this combined subsetting and data access operation the on-disk backend
+For this combined subsetting and data access operation the on-disk backend
 performed better than the in-memory `MSnExp`, while even requiring much less

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 
-The `fromFile` function can be used to determine from which file (sample) a
+The `fromFile` function can be used to determine the source file (sample)
 specific spectrum in the `MSnExp` object is. This function returns an `integer`
@@
 effectively applied to the data when m/z or intensity values are extracted from
-the object as it is the case in the `plot` call above.
+the object, e.g. in the `plot` call above.
 

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 The `fromFile` function can be used to determine the source file (sample)
-specific spectrum in the `MSnExp` object is. This function returns an `integer`
-vector, same length than spectra in the experiment, with the file index; the
+of a specific spectrum in the `MSnExp` object. This function returns an `integer`
+vector, of the same length as spectra in the experiment, with the file index. The
 file names can be accessed with the `fileNames` method. A `MSnExp` object can be

### 2020-07-02 large_experiment_processing.Rmd
Update large_experiment_processing.Rmd
Co-authored-by: Sebastian Gibb <mail@sebastiangibb.de>

@@
 vector, of the same length as spectra in the experiment, with the file index. The
-file names can be accessed with the `fileNames` method. A `MSnExp` object can be
+file names can be accessed with the `fileNames` method. An `MSnExp` object can be
 subsetted with `[` and e.g. the index of the spectra that should be retained. In
-the code block below we subset our `ms_dsk` object to contain only spectra from
+the code block below we subset our `ms_dsk` object to keep only spectra from
 the 3rd file.
@@
 
-Note however that there are also dedicated *filter* functions to subset an
+Note that there are also dedicated *filter* functions to subset an
 `MSnExp` object such as `filterFile`, `filterMsLevel`, `filterRt`, `filterMz`,

### 2020-07-15 msnbase2.Rnw
merge PRs, compile with red text

--- 2020-07-15 msnbase2.Rnw
@@
-\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[journal=jacsat,manuscript=article]{achemso}
+
+\usepackage[]{graphicx}
+\usepackage[]{color}
+
+\usepackage[]{xcolor}
+\usepackage[normalem]{ulem}
+
 %% maxwidth is the original width if it is less than linewidth
@@
 \sout{ProteomeDiscoverer (Thermo Scientific)}
-\red{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
+\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
 MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
@@
 proteomics mass spectrometry-based applications. It lays out a sound
-infrastructure to work with raw mass spectrometry \color{red}{data from MS files
+infrastructure to work with raw mass spectrometry \textcolor{red}{data from MS files
   in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
@@
 detailed step-by-step processing (for example smoothing and
-centroiding of \sout{raw} \color{red}{profile-mode MS} data, or normalisation and imputation of
+centroiding of \sout{raw} \textcolor{red}{profile-mode MS} data, or normalisation and imputation of
 quantitative data), analysis and visualisation of these data and the
 development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
+methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation and use cases 
+are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
+workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
 pertaining to raw mass spectrometry data handling and processing.
@@
 level, acquisition number, retention time, precursor m/z and intensity
-(for MS level 2 and above), and many more. \texttt{MSnbase}\color{red}{
-  ,\textit{via} the \texttt{mzR} package~\cite{Chambers:2012}, allows import of raw mass spectrometry
-  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and} provides a
-rich interface to manipulate such objects. The code chunk below
-illustrates such an object as displayed in the R console and an
+(for MS level 2 and above), and many
+more. \texttt{MSnbase}\textcolor{red}{, \textit{via} the \texttt{mzR}
+  package~\cite{Chambers:2012}, allows import of raw mass spectrometry
+  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and}
+provides a rich interface to manipulate such objects. The code chunk
+below illustrates such an object as displayed in the R console and an
 enumeration of the metadata fields.
@@
   efficiently handle very large mass spectrometry data and experiments and how
-  it} \code{red}{how \texttt{MSnbase}} can be used for data processing and
+  it} \textcolor{red}{how \texttt{MSnbase}} can be used for data processing and
 visualisation. We will also illustrate how it makes use of the forward-pipe
@@
 operator has proved useful to develop non-trivial analyses by combining
-individual functions into easily readable pipelines. \code{red}{An example for
+individual functions into easily readable pipelines. \textcolor{red}{An example for
   \texttt{MSnbase}'s ability to efficiently handle also very large mass
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects \color{red}{importing
-  data from an mzML file and} storing \sout{data}\color{red}{it} 
+demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
+  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
 either in-memory or on-disk.
@@
 
-\color{red}{Because the on-disk backend does not hold all the data in memory,
+\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
   direct data manipulations are not possible. We thus implemented a \textit{lazy
@@
 \sout{The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing} in the object itself. Operations on the raw
+\textit{lazy processing} in the object itself.} Operations on the raw
 data are stored in a processing queue and only effectively applied
-when the user accesses m/z or intensity values. \color{red}{As an additional
+when the user accesses m/z or intensity values. \textcolor{red}{As an additional
   advantage, operations on subsets of the data become much faster since data
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data \color{red}{imported from an mzML file} for \sout{the} serine and the same data after
+the raw MS profile data \textcolor{red}{imported from an mzML file} for \sout{the} serine and the same data after
 smoothing, centroiding and m/z refinement, as illustrated in the code

### 2020-07-15 msnbase2.tex
merge PRs, compile with red text

@@
 
-\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
+\usepackage{alltt}
+
+\usepackage[]{graphicx}
+\usepackage[]{color}
+
+\usepackage[]{xcolor}
+\usepackage[normalem]{ulem}
+
 %% maxwidth is the original width if it is less than linewidth
@@
 software libraries include Java-based jmzML~\cite{Cote:2010} or
-C/C++-based ProteoWizard~\cite{Chambers:2012}. ProteomeDiscoverer
-(Thermo Scientific), MaxQuant~\cite{Cox:2008} and
-PeptideShaker~\cite{Vaudel:2015} are among the most widely used
-user-centric applications.
+C/C++-based ProteoWizard~\cite{Chambers:2012}.
+\sout{ProteomeDiscoverer (Thermo Scientific)}
+\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
+MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
+most widely used user-centric applications.
 
@@
 proteomics mass spectrometry-based applications. It lays out a sound
-infrastructure to work with raw mass spectrometry, quantitative and
+infrastructure to work with raw mass spectrometry \textcolor{red}{data from MS files
+  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
 proteomics identification data. The package enables manipulation (for
@@
 detailed step-by-step processing (for example smoothing and
-centroiding of raw data, or normalisation and imputation of
+centroiding of \sout{raw} \textcolor{red}{profile-mode MS} data, or normalisation and imputation of
 quantitative data), analysis and visualisation of these data and the
 development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
+methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation and use cases 
+are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
+workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
 pertaining to raw mass spectrometry data handling and processing.
@@
 level, acquisition number, retention time, precursor m/z and intensity
-(for MS level 2 and above), and many more. \texttt{MSnbase} provides a
-rich interface to manipulate such objects. The code chunk below
-illustrates such an object as displayed in the R console and an
+(for MS level 2 and above), and many
+more. \texttt{MSnbase}\textcolor{red}{, \textit{via} the \texttt{mzR}
+  package~\cite{Chambers:2012}, allows import of raw mass spectrometry
+  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and}
+provides a rich interface to manipulate such objects. The code chunk
+below illustrates such an object as displayed in the R console and an
 enumeration of the metadata fields.
@@
 
-In the following sections, we describe \texttt{MSnbase}'s ability to
-efficiently handle very large mass spectrometry data and experiments
-and how it can be used for data processing and visualisation. We will
-also illustrate how it makes use of the forward-pipe operator
-(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by
-combining individual functions into easily readable pipelines.
+In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
+  efficiently handle very large mass spectrometry data and experiments and how
+  it} \textcolor{red}{how \texttt{MSnbase}} can be used for data processing and
+visualisation. We will also illustrate how it makes use of the forward-pipe
+operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
+operator has proved useful to develop non-trivial analyses by combining
+individual functions into easily readable pipelines. \textcolor{red}{An example for
+  \texttt{MSnbase}'s ability to efficiently handle also very large mass
+  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
+  files) is provided as Supplementary Information.}
 
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects storing data
+demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
+  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
 either in-memory or on-disk.
@@
 
-The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing} in the object itself. Operations on the raw
+\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
+  direct data manipulations are not possible. We thus implemented a \textit{lazy
+  processing} mechanism in this backend that caches any data manipulation
+operation in a processing queue in the object itself. Only when the user
+accesses m/z or intensity values these operations are applied}.
+\sout{The on-disk backend also offers efficient data manipulation by way of
+\textit{lazy processing} in the object itself.} Operations on the raw
 data are stored in a processing queue and only effectively applied
-when the user accesses m/z or intensity values. As an example, the
+when the user accesses m/z or intensity values. \textcolor{red}{As an additional
+  advantage, operations on subsets of the data become much faster since data
+  manipulations are applied only to data subsets instead of the full data set
+  at once. Also, data access in on-disk backends is parallelized by data file
+  ensuring a higher performance of this backend over conventional in-memory data
+  representations.} 
+As an example, the
 following short analysis pipeline, that can equally be applied to
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data for the serine and the same data after
+the raw MS profile data \textcolor{red}{imported from an mzML file} for \sout{the} serine and the same data after
 smoothing, centroiding and m/z refinement, as illustrated in the code
@@
 The version of \texttt{MSnbase} used in this manuscript is
-2.14.0. The main features presented here were available since
+2.14.2. The main features presented here were available since
 version 2.0. The code to reproduce the analyses and figures in this

### 2020-07-15 msnbase2.Rnw
addressing more reviewer comments

@@
 proteomics mass spectrometry-based applications. It lays out a sound
-infrastructure to work with raw mass spectrometry \textcolor{red}{data from MS files
-  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
-proteomics identification data. The package enables manipulation (for
-example subsetting, filtering, or accessing specific parts thereof),
-detailed step-by-step processing (for example smoothing and
-centroiding of \sout{raw} \textcolor{red}{profile-mode MS} data, or normalisation and imputation of
-quantitative data), analysis and visualisation of these data and the
-development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation and use cases 
-are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
-workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
-pertaining to raw mass spectrometry data handling and processing.
+infrastructure to work with raw mass spectrometry \textcolor{red}{data
+  from MS files in mzML, mzXML, mzData or ANDI-MS/netCDF format as
+  well as} quantitative and proteomics identification data. The
+package enables manipulation (for example subsetting, filtering, or
+accessing specific parts thereof), detailed step-by-step processing
+(for example smoothing and centroiding of \sout{raw}
+\textcolor{red}{profile-mode MS} data, or normalisation and imputation
+of quantitative data), analysis and visualisation of these data and
+the development of novel computational mass spectrometry
+methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation
+  and use cases are provided in \textit{package
+    vignettes}~\cite{MSnbaseVignettes} and
+  workflows~\cite{xcmsWorkflow}.} Here, we focus on the new
+developments pertaining to raw mass spectrometry data handling and
+processing.
 
@@
 level, acquisition number, retention time, precursor m/z and intensity
-(for MS level 2 and above), and many
-more. \texttt{MSnbase}\textcolor{red}{, \textit{via} the \texttt{mzR}
-  package~\cite{Chambers:2012}, allows import of raw mass spectrometry
-  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and}
-provides a rich interface to manipulate such objects. The code chunk
-below illustrates such an object as displayed in the R console and an
-enumeration of the metadata fields.
+(for MS level 2 and above), and many more. \texttt{MSnbase}
+\textcolor{red}{relies on the \texttt{mzR}
+  package~\cite{Chambers:2012} to import raw mass spectrometry data
+  from one of the many community-maintained open standards formats
+  (mzML, mzXML, mzdata or ANDI-MS/netCDF) and} provides a rich
+interface to manipulate such objects. The code chunk below illustrates
+such an object as displayed in the R console and an enumeration of the
+metadata fields.
 
@@
 
-In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
-  efficiently handle very large mass spectrometry data and experiments and how
-  it} \textcolor{red}{how \texttt{MSnbase}} can be used for data processing and
-visualisation. We will also illustrate how it makes use of the forward-pipe
-operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by combining
-individual functions into easily readable pipelines. \textcolor{red}{An example for
-  \texttt{MSnbase}'s ability to efficiently handle also very large mass
-  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
-  files) is provided as Supplementary Information.}
+In the following sections, we describe \sout{\texttt{MSnbase}'s
+  ability to efficiently handle very large mass spectrometry data and
+  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
+used for data processing and visualisation. \textcolor{red}{An example
+  of its ability to efficiently handle also very large mass
+  spectrometry experiments (in this case with 5,773,464 spectra in
+  1,182 mzXML files) is provided as supplementary information.}  We
+will also illustrate how it makes use of the forward-pipe operator
+(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
+operator has proved useful to develop non-trivial analyses by
+combining individual functions into easily readable pipelines.
 
@@
     xlab("Number of files") + 
-    theme(legend.position = "none")
+    theme(legend.position = c(0.35, 0.86),
+          legend.title = element_blank(),
+          legend.text=element_text(size = 6))
 
@@
   the in-memory (left) and on-disk (right) backends for 1, 10, 100
-  1000, 5000 and all 6103 spectra. On-disk backend: blue. In-memory
-  backend: red. Benchmarks were performed on a Dell XPS laptop with an
-  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads) and 7.5 GB RAM
-  running Ubuntu 18.04.4 LTS 64-bit. }
+  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
+  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
+  threads) and 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit. }
 \label{fig:bench}
@@
   of the segments were removed. The reconstructed full scan is shown
-  on panel (e).}
+  on panel (e). Spectra visualisation, as shown here, rely on the
+  \texttt{ggplot2} \cite{ggplot2} package. }
 \label{fig:bc}

### 2020-07-15 msnbase2.Rnw
addressing more reviewer comments

@@
   commodity hardware and illustrate how the package is used for
-  elegant data processing, method development and visualisation.
+  elegant data processing, method development, and visualisation.
 \end{abstract}
@@
 spectra-level metadata. Each spectrum is composed of m/z values and
-associated intensities. The metadata are represented by a table with
-variables along the columns and each row associated to a
-spectrum. Among the metadata available for each spectrum, there are MS
-level, acquisition number, retention time, precursor m/z and intensity
-(for MS level 2 and above), and many more. \texttt{MSnbase}
-\textcolor{red}{relies on the \texttt{mzR}
+associated intensities. The metadata are represented by a
+\textcolor{red}{single} table with variables along the columns and
+each row associated to a spectrum. Among the metadata available for
+each spectrum, there are MS level, acquisition number, retention time,
+precursor m/z and intensity (for MS level 2 and above), and many
+more. \texttt{MSnbase} \textcolor{red}{relies on the \texttt{mzR}
   package~\cite{Chambers:2012} to import raw mass spectrometry data
   from one of the many community-maintained open standards formats
-  (mzML, mzXML, mzdata or ANDI-MS/netCDF) and} provides a rich
+  (mzML, mzXML, mzData or ANDI-MS/netCDF) and} provides a rich
 interface to manipulate such objects. The code chunk below illustrates
@@
 demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
-  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
+  data from an mzML file and} storing \sout{data}\textcolor{red}{them} 
 either in-memory or on-disk.
@@
     ggplot2::scale_y_log10() +
-    ylab("Filtering time [millisec]") + 
+    ylab("Filtering time [ms]") + 
     xlab("Backend") +    
@@
     ggplot2::scale_y_log10() +
-    ylab("Raw data access [millisec]") +
+    ylab("Raw data access [ms]") +
     xlab("Number of spectra (out of 6103)") +
@@
 
-\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
-  direct data manipulations are not possible. We thus implemented a \textit{lazy
-  processing} mechanism in this backend that caches any data manipulation
-operation in a processing queue in the object itself. Only when the user
-accesses m/z or intensity values these operations are applied}.
-\sout{The on-disk backend also offers efficient data manipulation by way of
-\textit{lazy processing} in the object itself.} Operations on the raw
-data are stored in a processing queue and only effectively applied
-when the user accesses m/z or intensity values. \textcolor{red}{As an additional
-  advantage, operations on subsets of the data become much faster since data
-  manipulations are applied only to data subsets instead of the full data set
-  at once. Also, data access in on-disk backends is parallelized by data file
-  ensuring a higher performance of this backend over conventional in-memory data
-  representations.} 
-As an example, the
-following short analysis pipeline, that can equally be applied to
-in-memory or on-disk data, retains MS2 spectra acquired between 1000
-and 3000 seconds, extracts the m/z range corresponding to the TMT
-6-plex range and focuses on the MS2 spectra with a precursor intensity
-greater than $11 \times 10^6$ (the median precursor intensity).
+\textcolor{red}{Because the on-disk backend does not hold all the data
+  in memory, direct data manipulations are not possible. We thus
+  implemented a \textit{lazy processing} mechanism for this backend
+  that caches any data manipulation operations in a processing queue
+  in the object itself. Only when the user accesses m/z or intensity
+  values are these operations are applied}.  \sout{The on-disk backend
+  also offers efficient data manipulation by way of \textit{lazy
+    processing} in the object itself.} Operations on the raw data are
+stored in a processing queue and only effectively applied when the
+user accesses m/z or intensity values. \textcolor{red}{As an
+  additional advantage, operations on subsets of the data become much
+  faster since data manipulations are applied only to data subsets
+  instead of the full data set at once. Also, data access in on-disk
+  backends is parallelized by data file ensuring a higher performance
+  of this backend over conventional in-memory data representations.}
+As an example, the following short analysis pipeline, that can equally
+be applied to in-memory or on-disk data, retains MS2 spectra acquired
+between 1000 and 3000 seconds, extracts the m/z range corresponding to
+the TMT 6-plex range and focuses on the MS2 spectra with a precursor
+intensity greater than $11 \times 10^6$ (the median precursor
+intensity).
 
@@
 
-The \texttt{MSnExp} data structure and its interface constitute a
+The \texttt{MSnExp} data structure and its interface constitute an
 efficient prototyping environment for computational method
@@
 bc <- readMSData("boxcar.mzML", mode = "onDisk") %>%
-      bc_groups() %>%       ## identify BoxCar groups (bc_groups)
+      bc_groups() %>%       ## identify BoxCar groups (creates 'bc_groups')
       filterBoxCar() %>%    ## keep only BoxCar spectra
@@
 load("boxcar.rda")
+p4 <- p4 + xlab("m/z")
 p1 + p2 + p3 + p4 + 
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data \textcolor{red}{imported from an mzML file} for \sout{the} serine and the same data after
-smoothing, centroiding and m/z refinement, as illustrated in the code
-chunk below. Detailed execution and description of these operations
-can be found in the \emph{MSnbase: centroiding of profile-mode MS
-  data} \texttt{MSnbase} vignette.
+the raw MS profile data \textcolor{red}{imported from an mzML file}
+for \sout{the} serine and the same data after smoothing, centroiding
+and m/z refinement, as illustrated in the code chunk below. Detailed
+execution and description of these operations can be found in the
+\emph{MSnbase: centroiding of profile-mode MS data} \texttt{MSnbase}
+vignette.
 
@@
 \texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
-statistical analysis for quantitative proteomics data,
-\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
-MS files and build MassBank records,
+statistical analysis \sout{for} \textcolor{red}{of} quantitative
+proteomics data, \texttt{RMassBank}~\cite{Stravs:2013} to process
+metabolomics tandem MS files and build MassBank records,
 \texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system
 suitability monitoring and quality control of targeted proteomic
-experiments and the widely used \texttt{xcms}~\cite{Smith:2006} package
-for the processing and analysis of metabolomics data. \texttt{MSnbase}
-is also used in non-R/Bioconductor software, such as for example
-IsoProt~\cite{Griss:2019}, that provides a reproducible workflow for
-iTRAQ/TMT experiments. \texttt{MSnbase} currently ranks 101 out of
-1823 packages based on the monthly downloads from unique IP addresses,
-tallying over 1000 downloads from unique IP addresses each months.
+experiments and the widely used \texttt{xcms}~\cite{Smith:2006}
+package for the processing and analysis of metabolomics
+data. \texttt{MSnbase} is also used in non-R/Bioconductor software,
+such as for example IsoProt~\cite{Griss:2019}, that provides a
+reproducible workflow for iTRAQ/TMT experiments. \texttt{MSnbase}
+currently ranks 101 out of 1823 packages based on the monthly
+downloads from unique IP addresses, tallying over 1000 downloads from
+unique IP addresses each months.
 

### 2020-07-15 msnbase2.tex
addressing more reviewer comments

@@
 \hlstd{bc} \hlkwb{<-} \hlkwd{readMSData}\hlstd{(}\hlstr{"boxcar.mzML"}\hlstd{,} \hlkwc{mode} \hlstd{=} \hlstr{"onDisk"}\hlstd{)} \hlopt{%>%}
-      \hlkwd{bc_groups}\hlstd{()} \hlopt{%>%}       \hlcom{## identify BoxCar groups (bc_groups)}
+      \hlkwd{bc_groups}\hlstd{()} \hlopt{%>%}       \hlcom{## identify BoxCar groups (creates 'bc_groups')}
       \hlkwd{filterBoxCar}\hlstd{()} \hlopt{%>%}    \hlcom{## keep only BoxCar spectra}

### 2020-07-22 msnbase2.Rnw
address comments ...

@@
 data, i.e. a collection of individual spectra, as well as
-spectra-level metadata. Each spectrum is composed of m/z values and
-associated intensities. The metadata are represented by a
-\textcolor{red}{single} table with variables along the columns and
-each row associated to a spectrum. Among the metadata available for
-each spectrum, there are MS level, acquisition number, retention time,
-precursor m/z and intensity (for MS level 2 and above), and many
-more. \texttt{MSnbase} \textcolor{red}{relies on the \texttt{mzR}
-  package~\cite{Chambers:2012} to import raw mass spectrometry data
-  from one of the many community-maintained open standards formats
-  (mzML, mzXML, mzData or ANDI-MS/netCDF) and} provides a rich
-interface to manipulate such objects. The code chunk below illustrates
-such an object as displayed in the R console and an enumeration of the
-metadata fields.
+spectra-level metadata (Figure \ref{fig:raw}). Each spectrum is
+composed of m/z values and associated intensities. The metadata are
+represented by a \textcolor{red}{single} table with variables along
+the columns and each row associated to a spectrum. Among the metadata
+available for each spectrum, there are MS level, acquisition number,
+retention time, precursor m/z and intensity (for MS level 2 and
+above), and many more. \texttt{MSnbase} \textcolor{red}{relies on the
+  \texttt{mzR} package~\cite{Chambers:2012} to import raw mass
+  spectrometry data from one of the many community-maintained open
+  standards formats (mzML, mzXML, mzData or ANDI-MS/netCDF) and}
+provides a rich interface to manipulate such objects. The code chunk
+below illustrates such an object as displayed in the R console and an
+enumeration of the metadata fields.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{./figure/raw.png}
+  \caption{Schematic representation of what is referred to by
+    \textit{raw data}: a collection of mass spectra and a table
+    containing spectrum-level annotations along the lines. Raw data is
+    contained in one of the many community-maintained open standards
+    formats (mzML, mzXML, mzData or ANDI-MS/netCDF). }
+  \label{fig:raw}
+\end{figure}
 
@@
 After processing of the BoxCar data, the final object can either be
-further analysed in R or written back to disk as an \texttt{mzML} file
-using \texttt{writeMSData()} for processing with other tools.
+further analysed \sout{in R} \textcolor{red}{using \texttt{MSnbase}}
+or written back to disk as an \texttt{mzML} file using
+\texttt{writeMSData()} for processing with other tools.
 

### 2020-07-22 msnbase2.Rnw
address text comments from reviewer 4

@@
 Mass spectrometry is a powerful technology to assay chemical and
-biological samples. It is used in routine with well characterised
-protocol, as well a development platform, to improve on existing
-methods and devise new ones. The complexity and diversity of mass
-spectrometry yield data that is itself complex and often times of
-considerable size, that require non trivial processing before
-producing interpretable results. This is particularly relevant, and
-can constitute a significant challenge for method development: in
-addition to the development of sample processing and mass spectrometry
-methods, there is a need to process, analyse, interpret and assess
-these new data to demonstrate the improvement in the technical,
+biological samples. It is used in routine applications with well
+characterised \textcolor{red}{protocols} \textcolor{red}{such as
+  diagnostics}, as well as a development platform,
+\textcolor{red}{with the aim} to improve on existing \sout{methods}
+\textcolor{red}{protocols} and devise new ones. The complexity and
+diversity of mass spectrometry yield data that is itself complex and
+often times of considerable size, that require non trivial processing
+before producing interpretable results. \sout{This is particularly
+  relevant, and can} \textcolor{red}{The complexity and size of these
+  data} constitute a significant challenge for \sout{method}
+\textcolor{red}{protocol} development: in addition to the development
+of sample processing and mass spectrometry methods
+\textcolor{red}{that yield the raw data}, \textcolor{red}{it is
+  essential} \sout{there is a need} to process, analyse, interpret and
+assess these new data to demonstrate the improvement in the technical,
 analytical and computational workflows.
@@
   XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
-  threads) and 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit. }
+  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
+  \textcolor{red}{and SSD drive}. }
 \label{fig:bench}

### 2020-07-22 msnbase2.Rnw
new Package management and governance section

@@
 
-\section{Discussion}
-
-We have presented here some important functionality of
-\texttt{MSnbase} version 2. The new on-disk infrastructure enables
-large scale data analyses~\cite{Nothias:2020}, either using
-\texttt{MSnbase} directly or through packages that rely on it, such as
-\texttt{xcms}. We have also illustrated how \texttt{MSnbase} can be
-used for standard data analysis and visualisation, and how it can be
-used for method development and prototyping.
+\subsection{\textcolor{red}{Package management and governance}}
 
@@
 particularly significant ones such as the on-disk backend described
-herein. According to \texttt{MSnbase}'s Bioconductor page, there are
-36 Bioconductor packages that depend, import or suggest it. Among
-these are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
+herein. \textcolor{red}{Contributions to the package are explicitly
+  encouraged, rewarded by an official contributor status and coverned
+  by a code of conduct.}
+
+According to \texttt{MSnbase}'s Bioconductor page, there are 36
+Bioconductor packages that depend, import or suggest it. Among these
+are \texttt{pRoloc}~\cite{Gatto:2014a} to analyse mass
 spectrometry-based spatial proteomics data,
@@
 standard installation procedures.
+
+\textcolor{red}{The growth of \texttt{MSnbase} and the user support
+  provided over the years attest to the core maintainers commitment to
+  long-term development, and the quality and maintainability of the
+  code base.}
+
+\section{Discussion}
+
+We have presented here some important functionality of
+\texttt{MSnbase} version 2. The new on-disk infrastructure enables
+large scale data analyses~\cite{Nothias:2020}, either using
+\texttt{MSnbase} directly or through packages that rely on it, such as
+\texttt{xcms}. We have also illustrated how \texttt{MSnbase} can be
+used for standard data analysis and visualisation, and how it can be
+used for method development and prototyping.
 

### 2020-07-22 msnbase2.tex
new Package management and governance section

@@
 
+\textcolor{red}{The growth of \texttt{MSnbase} and the user support
+  provided over the years attest to the core maintainers commitment to
+  long-term development, and the quality and maintainability of the
+  code base.}
+
+\section{Discussion}
+
+We have presented here some important functionality of
+\texttt{MSnbase} version 2. The new on-disk infrastructure enables
+large scale data analyses~\cite{Nothias:2020}, either using
+\texttt{MSnbase} directly or through packages that rely on it, such as
+\texttt{xcms}. We have also illustrated how \texttt{MSnbase} can be
+used for standard data analysis and visualisation, and how it can be
+used for method development and prototyping.
+
 The version of \texttt{MSnbase} used in this manuscript is

### 2020-07-23 msnbase2.Rnw
rename section Package maintenance and governance
define elegant
explain subsetting overhead - closes #16

@@
 operator has proved useful to develop non-trivial analyses by
-combining individual functions into easily readable pipelines.
+combining individual functions into easily readable
+\textcolor{red}{and elegant} pipelines.
 
@@
   threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
-  \textcolor{red}{and SSD drive}. }
+  \textcolor{red}{and an SSD drive}. }
 \label{fig:bench}
@@
 advantageous reading and execution times and memory footprint of the
-on-disk backend are possible by retrieving only spectra data from the selected
-subset hence avoiding access to the full raw
-data. Once access to the spectra m/z and intensity values become
-mandatory (for example for plotting), then the in-memory backend
-becomes more efficient, as illustrated on Figure~\ref{fig:bench}
-(d). This gain is maximal when the whole dataset is accessed
-(i.e. all spectra are already in memory) and negligible when only subsets of the
-data are requested.
+on-disk backend are possible by retrieving only spectra data from the
+selected subset hence avoiding access to the full raw data. Once
+access to the spectra m/z and intensity values become mandatory (for
+example for plotting), then the in-memory backend becomes more
+efficient, as illustrated on Figure~\ref{fig:bench} (d).
+\textcolor{red}{The benefit of accessing data in memory is however
+  reduced by underlying copies that are performed during the
+  subsetting operation. When subsetting an in-memory \texttt{MSnExp}
+  into a new, smaller in-memory \texttt{MSnExp} instance, the matrices
+  that contain the spectra for the new object are copied, thus leading
+  to increased execution time and (transient, if the original data is
+  replaced) memory usage. Figure~\ref{fig:bench} (d) shows that the
+  larger the subset, that smaller the benefits of an in-memory
+  backend. The example with the 6103 spectra, corresponding to the
+  full data (i.e. all spectra are already in memory and there is no
+  memory management overhead) is representative of memory access only
+  and constitutes the best case scenario. } The gain of an in-memory
+backend is maximal when the whole dataset is accessed \sout{(i.e. all
+  spectra are already in memory)} and becomes negligible when large
+subsets of the data are requested.
 
@@
 
-\subsection{\textcolor{red}{Package management and governance}}
+\subsection{\textcolor{red}{Package maintenance and governance}}
 

### 2020-07-23 S04-ssd.R
ssd benchmarking

@@ -1,0 +1,79 @@
+## This script assumes that the same files are available in on an SSD drive
+## (ssd_dir) and hard drive (hd_dir).
+
+ssd_dir <- "./data"
+hd_dir <- "~/disk/tmp/data/"
+
+
+library("MSnbase")
+
+###############################
+## Compare object creation
+
+res <- data.frame(n = c(1, 10, 50, 100))
+res$sata <- res$ssd <- rep(NA, 4)
+
+for (i in seq_len(nrow(res))) {
+    n <- res[i, "n"]
+    fls_ssd <- dir(ssd_dir, full.names = TRUE)[1:n]
+    fls_hd <- dir(hd_dir, full.names = TRUE)[1:n]
+    res[i, "ssd"] <- system.time(readMSData(fls_ssd, mode = "onDisk"))[["elapsed"]]
+    res[i, "sata"] <- system.time(readMSData(fls_hd, mode = "onDisk"))[["elapsed"]]
+}
+
+res$ratio <- res$ssd / res$sata
+
+saveRDS(res, file = "bench_ssh.rds")
+
+res
+##     n     sata     ssd     ratio
+## 1   1    7.917   7.936 1.0023999
+## 2  10   87.955  79.196 0.9004150
+## 3  50  482.165 396.744 0.8228387
+## 4 100 1001.587 807.840 0.8065600
+
+
+###############################
+## Compare access to data
+
+library(microbenchmark)
+
+#########################################
+## Access spectra from a single file
+
+x1_ssd <- readMSData(dir(ssd_dir, full.names = TRUE)[1], mode = "onDisk")
+x1_hd <- readMSData(dir(hd_dir, full.names = TRUE)[1], mode = "onDisk")
+
+b <- microbenchmark(spectra(x1_ssd[1:100]),
+                    spectra(x1_hd[1:100]),
+                    times = 10L)
+
+## b - no noticeable difference when accessing 100 spectra in a single file
+## > Unit: seconds
+##                expr      min       lq     mean   median       uq      max    neval
+## spectra(x1_ssd[1:100]) 1.858352 1.864124 1.882133 1.871416 1.902859 1.923378    10
+##  spectra(x1_hd[1:100]) 1.855695 1.860831 1.896199 1.881418 1.889071 2.084152    10
+
+#########################################
+## Access spectra from different files
+
+x_ssd <- readMSData(dir(ssd_dir, full.names = TRUE)[1:50], mode = "onDisk")
[…]

### 2020-07-23 S04-ssd.R
light clean of s04

@@ -48,7 +48,8 @@ b <- microbenchmark(spectra(x1_ssd[1:100]),
                     spectra(x1_hd[1:100]),
                     times = 10L)
 
-## b - no noticeable difference when accessing 100 spectra in a single file
+## Results
+## No noticeable difference when accessing 100 spectra from a single file
 ## > Unit: seconds
 ##                expr      min       lq     mean   median       uq      max    neval
 ## spectra(x1_ssd[1:100]) 1.858352 1.864124 1.882133 1.871416 1.902859 1.923378    10
@@ -77,3 +78,7 @@ b <- microbenchmark(spectra(x_ssd[i]),
 ##               expr      min       lq     mean   median       uq      max neval
 ##  spectra(x_ssd[i])  67.1182  67.1182  67.1182  67.1182  67.1182  67.1182     1
 ##   spectra(x_hd[i]) 105.2481 105.2481 105.2481 105.2481 105.2481 105.2481     1
+
+b <- microbenchmark(spectra(x_ssd[i]),
+                    spectra(x_hd[i]),
+                    times = 5L)

### 2020-07-23 msnbase2.Rnw
ssd benchmarking
describe test data

@@
   threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
-  \textcolor{red}{and an SSD drive}. }
+  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
+    the benchmarking is a TMT 4-plex experiment acquired on a LTQ
+    Orbitrap Velos ((Thermo Fisher Scientific) available in the
+    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
 \label{fig:bench}
@@
 such as for example IsoProt~\cite{Griss:2019}, that provides a
-reproducible workflow for iTRAQ/TMT experiments. \texttt{MSnbase}
-currently ranks 101 out of 1823 packages based on the monthly
-downloads from unique IP addresses, tallying over 1000 downloads from
-unique IP addresses each months.
+reproducible workflow for iTRAQ/TMT experiments. \textcolor{red}{The
+  BioContainers~\cite{da_Veiga_Leprevost:2017} project offers a
+  dedicated container for the \texttt{MSnbase} package, this
+  facilitating the reuse of the package in third-party pipelines.}
+\texttt{MSnbase} currently ranks 101 out of 1823 packages based on the
+monthly downloads from unique IP addresses, tallying over 1000
+downloads from unique IP addresses each months.
 

### 2020-07-24 msnbase2.Rnw
further updates/cleaning of text

@@
   data. We focus on the new \textit{on-disk} infrastructure, that
-  allows the handling of large raw mass spectrometry experiment on
-  commodity hardware and illustrate how the package is used for
-  elegant data processing, method development, and visualisation.
+  allows the handling of large raw mass spectrometry
+  experiment\textcolor{red}{s} on commodity hardware and illustrate
+  how the package is used for elegant data processing, method
+  development, and visualisation.
 \end{abstract}
@@
 biological samples. It is used in routine applications with well
-characterised \textcolor{red}{protocols} \textcolor{red}{such as
-  diagnostics}, as well as a development platform,
+characterised \textcolor{red}{protocols} \textcolor{red}{such as in
+  clinical settings}, as well as a development platform,
 \textcolor{red}{with the aim} to improve on existing \sout{methods}
 \textcolor{red}{protocols} and devise new ones. The complexity and
-diversity of mass spectrometry yield data that is itself complex and
-often times of considerable size, that require non trivial processing
-before producing interpretable results. \sout{This is particularly
-  relevant, and can} \textcolor{red}{The complexity and size of these
-  data} constitute a significant challenge for \sout{method}
-\textcolor{red}{protocol} development: in addition to the development
-of sample processing and mass spectrometry methods
-\textcolor{red}{that yield the raw data}, \textcolor{red}{it is
-  essential} \sout{there is a need} to process, analyse, interpret and
-assess these new data to demonstrate the improvement in the technical,
-analytical and computational workflows.
+diversity of mass spectrometry yield \sout{data that is itself}
+complex \textcolor{red}{data} \sout{and often times} of considerable
+size, that require non trivial processing before producing
+interpretable results. \sout{This is particularly relevant, and can}
+\textcolor{red}{The complexity and size of these data} constitute a
+significant challenge for \sout{method} \textcolor{red}{protocol}
+development: in addition to the development of sample processing and
+mass spectrometry methods \textcolor{red}{that yield the raw data},
+\textcolor{red}{it is essential} \sout{there is a need} to process,
+analyse, interpret and assess these new data to demonstrate the
+improvement in the technical, analytical and computational workflows.
 
-Practitioners have a diverse catalogue of software tools to explore,
-process and interpret mass spectrometry data at their disposal. These
-range from low level software libraries that are aimed at programmers
-to enable development of new applications, to more user-oriented
-applications with graphical user interfaces which provide a more
-limited set of functionalities to address a defined scope. Examples of
-software libraries include Java-based jmzML~\cite{Cote:2010} or
-C/C++-based ProteoWizard~\cite{Chambers:2012}.
-\sout{ProteomeDiscoverer (Thermo Scientific)}
-\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
-MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
-most widely used user-centric applications.
+Practitioners have a diverse catalogue of software tools \sout{to
+  explore, process and interpret mass spectrometry data} at their
+disposal. These range from low level software libraries that are aimed
+at programmers to enable \textcolor{red}{the} development of new
+applications, to more user-oriented applications with graphical user
+interfaces which provide a more limited set of functionalities to
+address a defined scope. Examples of software libraries include
+Java-based jmzML~\cite{Cote:2010} or C/C++-based
+ProteoWizard~\cite{Chambers:2012}.  \sout{ProteomeDiscoverer (Thermo
+  Scientific)} \textcolor{red}{Thermo Scientific Proteome Discoverer
+  (Thermo Fisher Scientific)}, MaxQuant~\cite{Cox:2008} and
+PeptideShaker~\cite{Vaudel:2015} are among the most widely used
+user-centric applications.
 
@@
   standards formats (mzML, mzXML, mzData or ANDI-MS/netCDF) and}
-provides a rich interface to manipulate such objects. The code chunk
-below illustrates such an object as displayed in the R console and an
-enumeration of the metadata fields.
+provides a rich \textcolor{red}{and principled} interface to
+manipulate such objects. The code chunk below illustrates such an
+object as displayed in the R console and an enumeration of the
+metadata fields.
 
@@
   \includegraphics[width=0.8\linewidth]{./figure/raw.png}
-  \caption{Schematic representation of what is referred to by
-    \textit{raw data}: a collection of mass spectra and a table
-    containing spectrum-level annotations along the lines. Raw data is
-    contained in one of the many community-maintained open standards
-    formats (mzML, mzXML, mzData or ANDI-MS/netCDF). }
+  \caption{\textcolor{red}{Schematic representation of what is
+      referred to by \textit{raw data}: a collection of mass spectra
+      and a table containing spectrum-level annotations along the
+      lines. Raw data are imported from one of the many
+      community-maintained open standards formats (mzML, mzXML, mzData
+      or ANDI-MS/netCDF).} }
   \label{fig:raw}
@@
 used for data processing and visualisation. \textcolor{red}{An example
-  of its ability to efficiently handle also very large mass
+  of its ability to also efficiently handle very large mass
   spectrometry experiments (in this case with 5,773,464 spectra in
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
-  data from an mzML file and} storing \sout{data}\textcolor{red}{them}
-either in-memory or on-disk.
+demonstrates how to \textcolor{red}{import data from an mzML file} to
+create two \texttt{MSnExp} objects that store the data either
+in memory or on disk.
 
@@
   \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
-    the benchmarking is a TMT 4-plex experiment acquired on a LTQ
+    the benchmarking are a TMT 4-plex experiment acquired on a LTQ
     Orbitrap Velos ((Thermo Fisher Scientific) available in the
@@
 
-\textcolor{red}{Because the on-disk backend does not hold all the data
-  in memory, direct data manipulations are not possible. We thus
-  implemented a \textit{lazy processing} mechanism for this backend
-  that caches any data manipulation operations in a processing queue
-  in the object itself. Only when the user accesses m/z or intensity
-  values are these operations are applied}.  \sout{The on-disk backend
-  also offers efficient data manipulation by way of \textit{lazy
-    processing} in the object itself.} Operations on the raw data are
-stored in a processing queue and only effectively applied when the
-user accesses m/z or intensity values. \textcolor{red}{As an
-  additional advantage, operations on subsets of the data become much
-  faster since data manipulations are applied only to data subsets
-  instead of the full data set at once. Also, data access in on-disk
-  backends is parallelized by data file ensuring a higher performance
-  of this backend over conventional in-memory data representations.}
-As an example, the following short analysis pipeline, that can equally
-be applied to in-memory or on-disk data, retains MS2 spectra acquired
-between 1000 and 3000 seconds, extracts the m/z range corresponding to
-the TMT 6-plex range and focuses on the MS2 spectra with a precursor
-intensity greater than $11 \times 10^6$ (the median precursor
-intensity).
+\textcolor{red}{Because the on-disk backend does not hold all the
+  spectra data in memory, direct manipulations of these data are not
+  possible. We thus implemented a \textit{lazy processing} mechanism
+  for this backend that caches any data manipulation operations in a
+  processing queue in the object itself. These operations are then
+  applied only when the user accesses m/z or intensity values. }
+\sout{The on-disk backend also offers efficient data manipulation by
+  way of \textit{lazy processing} in the object itself. Operations on
+  the raw data are stored in a processing queue and only effectively
+  applied when the user accesses m/z or intensity values.}
+\textcolor{red}{As an additional advantage, operations on subsets of
+  the data become much faster since data manipulations are applied
+  only to data subsets instead of the full data set at once. Also,
+  on-disk data access is parallelized by data file ensuring a higher
+  performance of this backend over conventional in-memory data
+  representations.}  As an example, the following short analysis
+pipeline, that can equally be applied to in-memory or on-disk data,
+retains MS2 spectra acquired between 1000 and 3000 seconds, extracts
+the m/z range corresponding to the TMT 6-plex range and focuses on the
+MS2 spectra with a precursor intensity greater than $11 \times 10^6$
+(the median precursor intensity).
 
@@
   that contain the spectra for the new object are copied, thus leading
-  to increased execution time and (transient, if the original data is
+  to increased execution time and (transient, if the original data are
   replaced) memory usage. Figure~\ref{fig:bench} (d) shows that the
-  larger the subset, that smaller the benefits of an in-memory
-  backend. The example with the 6103 spectra, corresponding to the
-  full data (i.e. all spectra are already in memory and there is no
-  memory management overhead) is representative of memory access only
-  and constitutes the best case scenario. } The gain of an in-memory
-backend is maximal when the whole dataset is accessed \sout{(i.e. all
-  spectra are already in memory)} and becomes negligible when large
-subsets of the data are requested.
+  larger the subset, the smaller the benefits of an in-memory backend
+  become. The example with the 6103 spectra, corresponding to the full
+  data (i.e. all spectra are already in memory and there is no memory
+  management overhead) is representative of memory access only and
+  constitutes the best case scenario. } \sout{The gain of an in-memory
+  backend is maximal when the whole dataset is accessed (i.e. all
+  spectra are already in memory) and becomes negligible when large
+  subsets of the data are requested.}
 
@@
 \texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
-that demonstrates this. The simple method is composed of three steps
-is described below and illustrated with code from
-\texttt{MSnbaseBoxCar} in the following code chunk.
+that demonstrates this. The simple \sout{method}
+\textcolor{red}{pipeline} is composed of three steps,\sout{is}
+described below, and illustrated with code from \texttt{MSnbaseBoxCar}
+in the following code chunk.
 

### 2020-07-24 S01-ondisk.R

@@ -36,6 +36,22 @@ time_sz <- lapply(n, function(i) {
 
 save(time_sz, file = "bench_time_sz.rda")
 
+## Repeat reading time in triplicates
+
+time_2 <- data.frame(
+    n = rep(n, each = 3),
+    time_mem = rep(NA, 9),
+    time_dsk = rep(NA, 9))
+
+for (i in seq_len(nrow(time_2))) {
+    n <- time_2$n[i]
+    f <- fls[seq_len(n)]
+    time_2[i, "time_mem"] <- system.time(x_mem <- readMSData(f, mode = "inMemory"))[["elapsed"]]
+    time_2[i, "time_dsk"] <- system.time(x_dsk <- readMSData(f, mode = "onDisk"))[["elapsed"]]
+    print(time_2)
+}
+
+
 ## ----------------------------------------------------------------------
 ## BENCHMARK: filtering
 ## ----------------------------------------------------------------------

### 2020-07-24 msnbase2.Rnw
triplicated reading timings

@@
 
-<<time_sz>>=
+<<time_sz, message=FALSE>>=
 load("bench_time_sz.rda")
-time <- sapply(time_sz, function(x) x[, "time"])
+load("bench_time_2.rda")
+
 sz <- sapply(time_sz, function(x) x[, "sz"])
-colnames(sz) <- colnames(time) <- c(1, 5, 10)
-rownames(sz) <- rownames(time) <- c("in-memory", "on-disk")
+colnames(sz) <- c(1, 5, 10)
+rownames(sz) <- c("in-memory", "on-disk")
 sz <- round(sz/(1014^2), 2)
 
+set.seed(1L) ## set jittering
 p_read <- 
-    data.frame(time) %>%
-    rownames_to_column() %>%
-    pivot_longer(names_to = "files",
+    time_2 %>%
+    magrittr::set_colnames(c("n", "in-memory", "on-disk")) %>%
+    pivot_longer(names_to = "backend",
                  values_to = "time", 
-                 -rowname) %>%
-    rename(backend = rowname) %>%
-    mutate(files = as.numeric(sub("X", "", files))) %>%
-    ggplot(aes(x = files, y = time, colour = backend)) +
-    geom_point() +
-    geom_line() +
-    scale_x_continuous(breaks=c(1, 5, 10)) +
+                 cols = -n) %>%
+    ggplot(aes(x = factor(n), y = time, colour = backend)) + 
+    geom_jitter(width = 0.05, size = 2, alpha = 0.5) + 
+    geom_smooth(aes(group = backend),
+                method = "loess", 
+                se = FALSE,
+                size = 0.8) + 
     ylab("Reading time [s]") +
@@
     ggplot(aes(x = files, y = size, colour = backend)) +
-    geom_point() +
-    geom_line() +
+    geom_point(size = 2, alpha = 0.5) +
+    geom_line(size = 0.8) +
     scale_x_continuous(breaks=c(1, 5, 10)) +
@@
   \centering
-<<plot_bench>>=
+<<plot_bench, warning=FALSE, message=FALSE>>=
 (p_read + p_sz + p_filt) / p_access +
@@
 @
-\caption{(a) Reading time (in seconds) and (b) data size in memory (in
-  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
-  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interactions on in-memory and on-disk
-  data containing 6103 MS2 spectra.  (d) Access time to spectra for
-  the in-memory (left) and on-disk (right) backends for 1, 10, 100
-  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
-  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
-  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
+\caption{(a) Reading time (triplicates, in seconds) and (b) data size
+  in memory (in MB) to read/store 1, 5 and 10 files containing 1431
+  MS1 (on-disk only) and 6103 MS2 (on-disk and in-memory) spectra. (c)
+  Filtering benchmark assessed over 10 interactions on in-memory and
+  on-disk data containing 6103 MS2 spectra.  (d) Access time to
+  spectra for the in-memory (left) and on-disk (right) backends for 1,
+  10, 100 1000, 5000 and all 6103 spectra. Benchmarks were performed
+  on a Dell XPS laptop with an Intel i5-8250U processor 1.60 GHz (4
+  cores, 8 threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
   \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for

### 2020-07-24 msnbase2.tex
triplicated reading timings

@@
 \end{knitrout}
-\caption{(a) Reading time (in seconds) and (b) data size in memory (in
-  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
-  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interactions on in-memory and on-disk
-  data containing 6103 MS2 spectra.  (d) Access time to spectra for
-  the in-memory (left) and on-disk (right) backends for 1, 10, 100
-  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
-  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
-  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
+\caption{(a) Reading time (triplicates, in seconds) and (b) data size
+  in memory (in MB) to read/store 1, 5 and 10 files containing 1431
+  MS1 (on-disk only) and 6103 MS2 (on-disk and in-memory) spectra. (c)
+  Filtering benchmark assessed over 10 interactions on in-memory and
+  on-disk data containing 6103 MS2 spectra.  (d) Access time to
+  spectra for the in-memory (left) and on-disk (right) backends for 1,
+  10, 100 1000, 5000 and all 6103 spectra. Benchmarks were performed
+  on a Dell XPS laptop with an Intel i5-8250U processor 1.60 GHz (4
+  cores, 8 threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
   \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for

### 2020-07-24 S01-ondisk.R

@@ -51,6 +51,7 @@ for (i in seq_len(nrow(time_2))) {
     print(time_2)
 }
 
+save(time_2, file = "bench_time_2.rda")
 
 ## ----------------------------------------------------------------------
 ## BENCHMARK: filtering

### 2020-07-24 msnbase2.Rnw
mention triplicates - re-compute average improvement

@@
 @
-\caption{(a) Reading time (triplicates, in seconds) and (b) data size
-  in memory (in MB) to read/store 1, 5 and 10 files containing 1431
-  MS1 (on-disk only) and 6103 MS2 (on-disk and in-memory) spectra. (c)
-  Filtering benchmark assessed over 10 interactions on in-memory and
-  on-disk data containing 6103 MS2 spectra.  (d) Access time to
-  spectra for the in-memory (left) and on-disk (right) backends for 1,
-  10, 100 1000, 5000 and all 6103 spectra. Benchmarks were performed
-  on a Dell XPS laptop with an Intel i5-8250U processor 1.60 GHz (4
-  cores, 8 threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
-  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
-    the benchmarking are a TMT 4-plex experiment acquired on a LTQ
-    Orbitrap Velos ((Thermo Fisher Scientific) available in the
-    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
+\caption{(a) Reading time (\textcolor{red}{triplicates,} in seconds)
+  and (b) data size in memory (in MB) to read/store 1, 5 and 10 files
+  containing 1431 MS1 (on-disk only) and 6103 MS2 (on-disk and
+  in-memory) spectra. (c) Filtering benchmark assessed over 10
+  interactions on in-memory and on-disk data containing 6103 MS2
+  spectra.  (d) Access time to spectra for the in-memory (left) and
+  on-disk (right) backends for 1, 10, 100 1000, 5000 and all 6103
+  spectra. Benchmarks were performed on a Dell XPS laptop with an
+  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads), 7.5 GB RAM
+  running Ubuntu 18.04.4 LTS 64-bit \textcolor{red}{and an SSD
+    drive}. \textcolor{red}{The data used for the benchmarking are a
+    TMT 4-plex experiment acquired on a LTQ Orbitrap Velos ((Thermo
+    Fisher Scientific) available in the \texttt{msdata} package and
+    described in \citep{Gatto:2014}.}}
 \label{fig:bench}

### 2020-07-24 msnbase2.tex
mention triplicates - re-compute average improvement

--- 2020-07-24 msnbase2.tex
@@
 \end{knitrout}
-\caption{(a) Reading time (in seconds) and (b) data size in memory (in
-  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
-  only) and 6103 MS2 (on-disk and in-memory) spectra. (c) Filtering
-  benchmark assessed over 10 interactions on in-memory and on-disk
-  data containing 6103 MS2 spectra.  (d) Access time to spectra for
-  the in-memory (left) and on-disk (right) backends for 1, 10, 100
-  1000, 5000 and all 6103 spectra. Benchmarks were performed on a Dell
-  XPS laptop with an Intel i5-8250U processor 1.60 GHz (4 cores, 8
-  threads), 7.5 GB RAM running Ubuntu 18.04.4 LTS 64-bit
-  \textcolor{red}{and an SSD drive}. \textcolor{red}{The data used for
-    the benchmarking are a TMT 4-plex experiment acquired on a LTQ
-    Orbitrap Velos ((Thermo Fisher Scientific) available in the
-    \texttt{msdata} package and described in \citep{Gatto:2014}.}}
+\caption{(a) Reading time (\textcolor{red}{triplicates,} in seconds)
+  and (b) data size in memory (in MB) to read/store 1, 5 and 10 files
+  containing 1431 MS1 (on-disk only) and 6103 MS2 (on-disk and
+  in-memory) spectra. (c) Filtering benchmark assessed over 10
+  interactions on in-memory and on-disk data containing 6103 MS2
+  spectra.  (d) Access time to spectra for the in-memory (left) and
+  on-disk (right) backends for 1, 10, 100 1000, 5000 and all 6103
+  spectra. Benchmarks were performed on a Dell XPS laptop with an
+  Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads), 7.5 GB RAM
+  running Ubuntu 18.04.4 LTS 64-bit \textcolor{red}{and an SSD
+    drive}. \textcolor{red}{The data used for the benchmarking are a
+    TMT 4-plex experiment acquired on a LTQ Orbitrap Velos ((Thermo
+    Fisher Scientific) available in the \texttt{msdata} package and
+    described in \citep{Gatto:2014}.}}
 \label{fig:bench}

### 2020-07-24 S01-ondisk.R

@@ -53,6 +53,13 @@ for (i in seq_len(nrow(time_2))) {
 
 save(time_2, file = "bench_time_2.rda")
 
+
+## time_2 %>%
+##     mutate(ratio = time_mem/time_dsk) %>%
+##     pull(ratio) %>%
+##     mean
+## 5.367308
+
 ## ----------------------------------------------------------------------
 ## BENCHMARK: filtering
 ## ----------------------------------------------------------------------

### 2020-08-04 large_experiment_processing.Rmd

@@
 output:
-    BiocStyle::pdf_document
+    BiocStyle::pdf_document:
+        includes:
+            in_header: "header.sty"
 ---
@@
 `filterPrecursorMz` or `filterIsolationWindow`. In the example below we use the
-`filterRt` function to further subset our data to contain only spectra acquired
+`filterRt` function to further subset our data to keep only spectra acquired
 within a certain time range.
@@
 
-```{r precursor-selection-plot-1}
+```{r precursor-selection-plot-1, fig.cap = "Example spectrum of the data set."}
 system.time(
@@
 
-```{r precursor-selection-plot-normalized}
+```{r precursor-selection-plot-normalized, fig.cap = "Example spectrum after cleaning."}
 system.time(

### 2020-08-28 msnbase2.Rnw
add keywords and assoc content section
udpate keywords
fix typos identified by rev 4

@@
 \end{abstract}
+
+Keywords: R, Bioconductor, mass spectrometry, software, metabolomics,
+proteomics, visualisation, reproducible research
 
@@
     drive}. \textcolor{red}{The data used for the benchmarking are a
-    TMT 4-plex experiment acquired on a LTQ Orbitrap Velos ((Thermo
+    TMT 4-plex experiment acquired on a LTQ Orbitrap Velos (Thermo
     Fisher Scientific) available in the \texttt{msdata} package and
@@
 herein. \textcolor{red}{Contributions to the package are explicitly
-  encouraged, rewarded by an official contributor status and coverned
+  encouraged, rewarded by an official contributor status and governed
   by a code of conduct.}
@@
 
+\section{Associated Content}
+
+Supplementary file 1: script documenting the processing of 1182 mzXML
+files (5773464 spectra) using \texttt{MSnbase}.
+
 \begin{acknowledgement}

### 2020-08-29 msnbase2.Rnw
add sentence about on-disk or im-mem (rev 3 suggestion)

@@
 
+\textcolor{blue}{The on-disk backend has become the preferred backend
+  for large data manipulation, and the only viable alternative for
+  very large data (i.e. larger than the available RAM) and/or when
+  several MS levels are to loaded and handled simultaneously. The
+  on-disk backend can still prove useful in cases when small MS2-only
+  data are to be analysed and will remain available in future versions
+  of \texttt{MSnbase}.}
+
 \subsection{Prototyping}

### 2020-08-29 msnbase2.Rnw
remove all \sout{}

@@
   clinical settings}, as well as a development platform,
-\textcolor{red}{with the aim} to improve on existing \sout{methods}
+\textcolor{red}{with the aim} to improve on existing
 \textcolor{red}{protocols} and devise new ones. The complexity and
-diversity of mass spectrometry yield \sout{data that is itself}
-complex \textcolor{red}{data} \sout{and often times} of considerable
-size, that require non trivial processing before producing
-interpretable results. \sout{This is particularly relevant, and can}
-\textcolor{red}{The complexity and size of these data} constitute a
-significant challenge for \sout{method} \textcolor{red}{protocol}
-development: in addition to the development of sample processing and
-mass spectrometry methods \textcolor{red}{that yield the raw data},
-\textcolor{red}{it is essential} \sout{there is a need} to process,
-analyse, interpret and assess these new data to demonstrate the
-improvement in the technical, analytical and computational workflows.
+diversity of mass spectrometry yield complex \textcolor{red}{data} of
+considerable size, that require non trivial processing before
+producing interpretable results.  \textcolor{red}{The complexity and
+  size of these data} constitute a significant challenge for
+\textcolor{red}{protocol} development: in addition to the development
+of sample processing and mass spectrometry methods
+\textcolor{red}{that yield the raw data}, \textcolor{red}{it is
+  essential} to process, analyse, interpret and assess these new data
+to demonstrate the improvement in the technical, analytical and
+computational workflows.
 
-Practitioners have a diverse catalogue of software tools \sout{to
-  explore, process and interpret mass spectrometry data} at their
+Practitioners have a diverse catalogue of software tools at their
 disposal. These range from low level software libraries that are aimed
@@
 Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}.  \sout{ProteomeDiscoverer (Thermo
-  Scientific)} \textcolor{red}{Thermo Scientific Proteome Discoverer
-  (Thermo Fisher Scientific)}, MaxQuant~\cite{Cox:2008} and
-PeptideShaker~\cite{Vaudel:2015} are among the most widely used
-user-centric applications.
+ProteoWizard~\cite{Chambers:2012}.  \textcolor{red}{Thermo Scientific
+  Proteome Discoverer (Thermo Fisher Scientific)},
+MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
+among the most widely used user-centric applications.
 
@@
 accessing specific parts thereof), detailed step-by-step processing
-(for example smoothing and centroiding of \sout{raw}
-\textcolor{red}{profile-mode MS} data, or normalisation and imputation
-of quantitative data), analysis and visualisation of these data and
-the development of novel computational mass spectrometry
+(for example smoothing and centroiding of \textcolor{red}{profile-mode
+  MS} data, or normalisation and imputation of quantitative data),
+analysis and visualisation of these data and the development of novel
+computational mass spectrometry
 methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation
@@
 
-In the following sections, we describe \sout{\texttt{MSnbase}'s
-  ability to efficiently handle very large mass spectrometry data and
-  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
-used for data processing and visualisation. \textcolor{red}{An example
-  of its ability to also efficiently handle very large mass
-  spectrometry experiments (in this case with 5,773,464 spectra in
-  1,182 mzXML files) is provided as supplementary information.}  We
-will also illustrate how it makes use of the forward-pipe operator
-(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by
-combining individual functions into easily readable
-\textcolor{red}{and elegant} pipelines.
+In the following sections, we describe \textcolor{red}{how
+  \texttt{MSnbase}} can be used for data processing and
+visualisation. \textcolor{red}{An example of its ability to also
+  efficiently handle very large mass spectrometry experiments (in this
+  case with 5,773,464 spectra in 1,182 mzXML files) is provided as
+  supplementary information.}  We will also illustrate how it makes
+use of the forward-pipe operator (\texttt{\%>\%}) defined in the
+\texttt{magrittr} package. This operator has proved useful to develop
+non-trivial analyses by combining individual functions into easily
+readable \textcolor{red}{and elegant} pipelines.
 
@@
   applied only when the user accesses m/z or intensity values. }
-\sout{The on-disk backend also offers efficient data manipulation by
-  way of \textit{lazy processing} in the object itself. Operations on
-  the raw data are stored in a processing queue and only effectively
-  applied when the user accesses m/z or intensity values.}
 \textcolor{red}{As an additional advantage, operations on subsets of
@@
   management overhead) is representative of memory access only and
-  constitutes the best case scenario. } \sout{The gain of an in-memory
-  backend is maximal when the whole dataset is accessed (i.e. all
-  spectra are already in memory) and becomes negligible when large
-  subsets of the data are requested.}
+  constitutes the best case scenario. }
 
@@
 \texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
-that demonstrates this. The simple \sout{method}
-\textcolor{red}{pipeline} is composed of three steps,\sout{is}
-described below, and illustrated with code from \texttt{MSnbaseBoxCar}
-in the following code chunk.
+that demonstrates this. The simple \textcolor{red}{pipeline} is
+composed of three steps, described below, and illustrated with code
+from \texttt{MSnbaseBoxCar} in the following code chunk.
 
@@
 After processing of the BoxCar data, the final object can either be
-further analysed \sout{in R} \textcolor{red}{using \texttt{MSnbase}}
-or written back to disk as an \texttt{mzML} file using
-\texttt{writeMSData()} for processing with other tools.
+further analysed \textcolor{red}{using \texttt{MSnbase}} or written
+back to disk as an \texttt{mzML} file using \texttt{writeMSData()} for
+processing with other tools.
 
@@
 the raw MS profile data \textcolor{red}{imported from an mzML file}
-for \sout{the} serine and the same data after smoothing, centroiding
-and m/z refinement, as illustrated in the code chunk below. Detailed
-execution and description of these operations can be found in the
-\emph{MSnbase: centroiding of profile-mode MS data} \texttt{MSnbase}
-vignette.
+for serine and the same data after smoothing, centroiding and m/z
+refinement, as illustrated in the code chunk below. Detailed execution
+and description of these operations can be found in the \emph{MSnbase:
+  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.
 
@@
 \texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
-statistical analysis \sout{for} \textcolor{red}{of} quantitative
-proteomics data, \texttt{RMassBank}~\cite{Stravs:2013} to process
-metabolomics tandem MS files and build MassBank records,
+statistical analysis \textcolor{red}{of} quantitative proteomics data,
+\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
+MS files and build MassBank records,
 \texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system

### 2020-08-29 msnbase2.Rnw
recompile pdf

@@
   allows the handling of large raw mass spectrometry
-  experiment\textcolor{red}{s} on commodity hardware and illustrate
+  experiment\textcolor{black}{s} on commodity hardware and illustrate
   how the package is used for elegant data processing, method
@@
 biological samples. It is used in routine applications with well
-characterised \textcolor{red}{protocols} \textcolor{red}{such as in
+characterised \textcolor{black}{protocols} \textcolor{black}{such as in
   clinical settings}, as well as a development platform,
-\textcolor{red}{with the aim} to improve on existing
-\textcolor{red}{protocols} and devise new ones. The complexity and
-diversity of mass spectrometry yield complex \textcolor{red}{data} of
+\textcolor{black}{with the aim} to improve on existing
+\textcolor{black}{protocols} and devise new ones. The complexity and
+diversity of mass spectrometry yield complex \textcolor{black}{data} of
 considerable size, that require non trivial processing before
-producing interpretable results.  \textcolor{red}{The complexity and
+producing interpretable results.  \textcolor{black}{The complexity and
   size of these data} constitute a significant challenge for
-\textcolor{red}{protocol} development: in addition to the development
+\textcolor{black}{protocol} development: in addition to the development
 of sample processing and mass spectrometry methods
-\textcolor{red}{that yield the raw data}, \textcolor{red}{it is
+\textcolor{black}{that yield the raw data}, \textcolor{black}{it is
   essential} to process, analyse, interpret and assess these new data
@@
 disposal. These range from low level software libraries that are aimed
-at programmers to enable \textcolor{red}{the} development of new
+at programmers to enable \textcolor{black}{the} development of new
 applications, to more user-oriented applications with graphical user
@@
 Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}.  \textcolor{red}{Thermo Scientific
+ProteoWizard~\cite{Chambers:2012}.  \textcolor{black}{Thermo Scientific
   Proteome Discoverer (Thermo Fisher Scientific)},
@@
 proteomics mass spectrometry-based applications. It lays out a sound
-infrastructure to work with raw mass spectrometry \textcolor{red}{data
+infrastructure to work with raw mass spectrometry \textcolor{black}{data
   from MS files in mzML, mzXML, mzData or ANDI-MS/netCDF format as
@@
 accessing specific parts thereof), detailed step-by-step processing
-(for example smoothing and centroiding of \textcolor{red}{profile-mode
+(for example smoothing and centroiding of \textcolor{black}{profile-mode
   MS} data, or normalisation and imputation of quantitative data),
@@
 computational mass spectrometry
-methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation
+methods~\cite{Stanstrup:2019}. \textcolor{black}{Extensive documentation
   and use cases are provided in \textit{package
@@
 composed of m/z values and associated intensities. The metadata are
-represented by a \textcolor{red}{single} table with variables along
+represented by a \textcolor{black}{single} table with variables along
 the columns and each row associated to a spectrum. Among the metadata
@@
 retention time, precursor m/z and intensity (for MS level 2 and
-above), and many more. \texttt{MSnbase} \textcolor{red}{relies on the
+above), and many more. \texttt{MSnbase} \textcolor{black}{relies on the
   \texttt{mzR} package~\cite{Chambers:2012} to import raw mass
@@
   standards formats (mzML, mzXML, mzData or ANDI-MS/netCDF) and}
-provides a rich \textcolor{red}{and principled} interface to
+provides a rich \textcolor{black}{and principled} interface to
 manipulate such objects. The code chunk below illustrates such an
@@
   \includegraphics[width=0.8\linewidth]{./figure/raw.png}
-  \caption{\textcolor{red}{Schematic representation of what is
+  \caption{\textcolor{black}{Schematic representation of what is
       referred to by \textit{raw data}: a collection of mass spectra
@@
 
-In the following sections, we describe \textcolor{red}{how
+In the following sections, we describe \textcolor{black}{how
   \texttt{MSnbase}} can be used for data processing and
-visualisation. \textcolor{red}{An example of its ability to also
+visualisation. \textcolor{black}{An example of its ability to also
   efficiently handle very large mass spectrometry experiments (in this
@@
 non-trivial analyses by combining individual functions into easily
-readable \textcolor{red}{and elegant} pipelines.
+readable \textcolor{black}{and elegant} pipelines.
 
@@
 \textit{in-memory} and \textit{on-disk}. The following code chunk
-demonstrates how to \textcolor{red}{import data from an mzML file} to
+demonstrates how to \textcolor{black}{import data from an mzML file} to
 create two \texttt{MSnExp} objects that store the data either
@@
 @
-\caption{(a) Reading time (\textcolor{red}{triplicates,} in seconds)
+\caption{(a) Reading time (\textcolor{black}{triplicates,} in seconds)
   and (b) data size in memory (in MB) to read/store 1, 5 and 10 files
@@
   Intel i5-8250U processor 1.60 GHz (4 cores, 8 threads), 7.5 GB RAM
-  running Ubuntu 18.04.4 LTS 64-bit \textcolor{red}{and an SSD
-    drive}. \textcolor{red}{The data used for the benchmarking are a
+  running Ubuntu 18.04.4 LTS 64-bit \textcolor{black}{and an SSD
+    drive}. \textcolor{black}{The data used for the benchmarking are a
     TMT 4-plex experiment acquired on a LTQ Orbitrap Velos (Thermo
@@
 
-\textcolor{red}{Because the on-disk backend does not hold all the
+\textcolor{black}{Because the on-disk backend does not hold all the
   spectra data in memory, direct manipulations of these data are not
@@
   applied only when the user accesses m/z or intensity values. }
-\textcolor{red}{As an additional advantage, operations on subsets of
+\textcolor{black}{As an additional advantage, operations on subsets of
   the data become much faster since data manipulations are applied
@@
 efficient, as illustrated on Figure~\ref{fig:bench} (d).
-\textcolor{red}{The benefit of accessing data in memory is however
+\textcolor{black}{The benefit of accessing data in memory is however
   reduced by underlying copies that are performed during the
@@
 
-\textcolor{blue}{The on-disk backend has become the preferred backend
-  for large data manipulation, and the only viable alternative for
-  very large data (i.e. larger than the available RAM) and/or when
-  several MS levels are to loaded and handled simultaneously. The
-  on-disk backend can still prove useful in cases when small MS2-only
-  data are to be analysed and will remain available in future versions
-  of \texttt{MSnbase}.}
+\textcolor{black}{The on-disk backend has become the preferred backend
+  for large data, and the only viable alternative when the size of the
+  data exceeds the available RAM and/or when several MS levels are to
+  be loaded and handled simultaneously. The on-disk backend can still
+  prove useful in cases when small MS2-only data are to be analysed
+  and will remain available in future versions of \texttt{MSnbase}.}
 
@@
 \texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
-that demonstrates this. The simple \textcolor{red}{pipeline} is
+that demonstrates this. The simple \textcolor{black}{pipeline} is
 composed of three steps, described below, and illustrated with code
@@
 After processing of the BoxCar data, the final object can either be
-further analysed \textcolor{red}{using \texttt{MSnbase}} or written
+further analysed \textcolor{black}{using \texttt{MSnbase}} or written
 back to disk as an \texttt{mzML} file using \texttt{writeMSData()} for
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data \textcolor{red}{imported from an mzML file}
+the raw MS profile data \textcolor{black}{imported from an mzML file}
 for serine and the same data after smoothing, centroiding and m/z
@@
 
-\subsection{\textcolor{red}{Package maintenance and governance}}
+\subsection{\textcolor{black}{Package maintenance and governance}}
 
@@
 particularly significant ones such as the on-disk backend described
-herein. \textcolor{red}{Contributions to the package are explicitly
+herein. \textcolor{black}{Contributions to the package are explicitly
   encouraged, rewarded by an official contributor status and governed
@@
 \texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
-statistical analysis \textcolor{red}{of} quantitative proteomics data,
+statistical analysis \textcolor{black}{of} quantitative proteomics data,
 \texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
@@
 such as for example IsoProt~\cite{Griss:2019}, that provides a
-reproducible workflow for iTRAQ/TMT experiments. \textcolor{red}{The
+reproducible workflow for iTRAQ/TMT experiments. \textcolor{black}{The
   BioContainers~\cite{da_Veiga_Leprevost:2017} project offers a
@@
 
-\textcolor{red}{The growth of \texttt{MSnbase} and the user support
+\textcolor{black}{The growth of \texttt{MSnbase} and the user support
   provided over the years attest to the core maintainers commitment to

### 2020-08-29 msnbase2.tex
recompile pdf

--- 2020-08-28 msnbase2.tex
@@
 biological samples. It is used in routine applications with well
-characterised \textcolor{red}{protocols} \textcolor{red}{such as in
+characterised \textcolor{black}{protocols} \textcolor{black}{such as in
   clinical settings}, as well as a development platform,
-\textcolor{red}{with the aim} to improve on existing \sout{methods}
-\textcolor{red}{protocols} and devise new ones. The complexity and
-diversity of mass spectrometry yield \sout{data that is itself}
-complex \textcolor{red}{data} \sout{and often times} of considerable
-size, that require non trivial processing before producing
-interpretable results. \sout{This is particularly relevant, and can}
-\textcolor{red}{The complexity and size of these data} constitute a
-significant challenge for \sout{method} \textcolor{red}{protocol}
-development: in addition to the development of sample processing and
-mass spectrometry methods \textcolor{red}{that yield the raw data},
-\textcolor{red}{it is essential} \sout{there is a need} to process,
-analyse, interpret and assess these new data to demonstrate the
-improvement in the technical, analytical and computational workflows.
+\textcolor{black}{with the aim} to improve on existing
+\textcolor{black}{protocols} and devise new ones. The complexity and
+diversity of mass spectrometry yield complex \textcolor{black}{data} of
+considerable size, that require non trivial processing before
+producing interpretable results.  \textcolor{black}{The complexity and
+  size of these data} constitute a significant challenge for
+\textcolor{black}{protocol} development: in addition to the development
+of sample processing and mass spectrometry methods
+\textcolor{black}{that yield the raw data}, \textcolor{black}{it is
+  essential} to process, analyse, interpret and assess these new data
+to demonstrate the improvement in the technical, analytical and
+computational workflows.
 
-Practitioners have a diverse catalogue of software tools \sout{to
-  explore, process and interpret mass spectrometry data} at their
+Practitioners have a diverse catalogue of software tools at their
 disposal. These range from low level software libraries that are aimed
-at programmers to enable \textcolor{red}{the} development of new
+at programmers to enable \textcolor{black}{the} development of new
 applications, to more user-oriented applications with graphical user
@@
 Java-based jmzML~\cite{Cote:2010} or C/C++-based
-ProteoWizard~\cite{Chambers:2012}.  \sout{ProteomeDiscoverer (Thermo
-  Scientific)} \textcolor{red}{Thermo Scientific Proteome Discoverer
-  (Thermo Fisher Scientific)}, MaxQuant~\cite{Cox:2008} and
-PeptideShaker~\cite{Vaudel:2015} are among the most widely used
-user-centric applications.
+ProteoWizard~\cite{Chambers:2012}.  \textcolor{black}{Thermo Scientific
+  Proteome Discoverer (Thermo Fisher Scientific)},
+MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are
+among the most widely used user-centric applications.
 
@@
 accessing specific parts thereof), detailed step-by-step processing
-(for example smoothing and centroiding of \sout{raw}
-\textcolor{red}{profile-mode MS} data, or normalisation and imputation
-of quantitative data), analysis and visualisation of these data and
-the development of novel computational mass spectrometry
-methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation
+(for example smoothing and centroiding of \textcolor{black}{profile-mode
+  MS} data, or normalisation and imputation of quantitative data),
+analysis and visualisation of these data and the development of novel
+computational mass spectrometry
+methods~\cite{Stanstrup:2019}. \textcolor{black}{Extensive documentation
   and use cases are provided in \textit{package
@@
 
-In the following sections, we describe \sout{\texttt{MSnbase}'s
-  ability to efficiently handle very large mass spectrometry data and
-  experiments and how it} \textcolor{red}{how \texttt{MSnbase}} can be
-used for data processing and visualisation. \textcolor{red}{An example
-  of its ability to also efficiently handle very large mass
-  spectrometry experiments (in this case with 5,773,464 spectra in
-  1,182 mzXML files) is provided as supplementary information.}  We
-will also illustrate how it makes use of the forward-pipe operator
-(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
-operator has proved useful to develop non-trivial analyses by
-combining individual functions into easily readable
-\textcolor{red}{and elegant} pipelines.
+In the following sections, we describe \textcolor{black}{how
+  \texttt{MSnbase}} can be used for data processing and
+visualisation. \textcolor{black}{An example of its ability to also
+  efficiently handle very large mass spectrometry experiments (in this
+  case with 5,773,464 spectra in 1,182 mzXML files) is provided as
+  supplementary information.}  We will also illustrate how it makes
+use of the forward-pipe operator (\texttt{\%>\%}) defined in the
+\texttt{magrittr} package. This operator has proved useful to develop
+non-trivial analyses by combining individual functions into easily
+readable \textcolor{black}{and elegant} pipelines.
 
@@
 \end{knitrout}
-\caption{(a) Reading time (\textcolor{red}{triplicates,} in seconds)
+\caption{(a) Reading time (\textcolor{black}{triplicates,} in seconds)
   and (b) data size in memory (in MB) to read/store 1, 5 and 10 files
@@
   applied only when the user accesses m/z or intensity values. }
-\sout{The on-disk backend also offers efficient data manipulation by
-  way of \textit{lazy processing} in the object itself. Operations on
-  the raw data are stored in a processing queue and only effectively
-  applied when the user accesses m/z or intensity values.}
-\textcolor{red}{As an additional advantage, operations on subsets of
+\textcolor{black}{As an additional advantage, operations on subsets of
   the data become much faster since data manipulations are applied
@@
   management overhead) is representative of memory access only and
-  constitutes the best case scenario. } \sout{The gain of an in-memory
-  backend is maximal when the whole dataset is accessed (i.e. all
-  spectra are already in memory) and becomes negligible when large
-  subsets of the data are requested.}
+  constitutes the best case scenario. }
+
+\textcolor{black}{The on-disk backend has become the preferred backend
+  for large data, and the only viable alternative when the size of the
+  data exceeds the available RAM and/or when several MS levels are to
+  be loaded and handled simultaneously. The on-disk backend can still
+  prove useful in cases when small MS2-only data are to be analysed
+  and will remain available in future versions of \texttt{MSnbase}.}
 
@@
 \texttt{MSnbaseBoxCar} package\cite{MSnbaseBoxCar} is a small package
-that demonstrates this. The simple \sout{method}
-\textcolor{red}{pipeline} is composed of three steps,\sout{is}
-described below, and illustrated with code from \texttt{MSnbaseBoxCar}
-in the following code chunk.
+that demonstrates this. The simple \textcolor{black}{pipeline} is
+composed of three steps, described below, and illustrated with code
+from \texttt{MSnbaseBoxCar} in the following code chunk.
 
@@
 After processing of the BoxCar data, the final object can either be
-further analysed \sout{in R} \textcolor{red}{using \texttt{MSnbase}}
-or written back to disk as an \texttt{mzML} file using
-\texttt{writeMSData()} for processing with other tools.
+further analysed \textcolor{black}{using \texttt{MSnbase}} or written
+back to disk as an \texttt{mzML} file using \texttt{writeMSData()} for
+processing with other tools.
 
@@
 the efficiency of raw data processing. Figure \ref{fig:cent} compares
-the raw MS profile data \textcolor{red}{imported from an mzML file}
-for \sout{the} serine and the same data after smoothing, centroiding
-and m/z refinement, as illustrated in the code chunk below. Detailed
-execution and description of these operations can be found in the
-\emph{MSnbase: centroiding of profile-mode MS data} \texttt{MSnbase}
-vignette.
+the raw MS profile data \textcolor{black}{imported from an mzML file}
+for serine and the same data after smoothing, centroiding and m/z
+refinement, as illustrated in the code chunk below. Detailed execution
+and description of these operations can be found in the \emph{MSnbase:
+  centroiding of profile-mode MS data} \texttt{MSnbase} vignette.
 
@@
 \texttt{DAPAR} and \texttt{ProStaR}~\cite{Wieczorek:2017} for the
-statistical analysis \sout{for} \textcolor{red}{of} quantitative
-proteomics data, \texttt{RMassBank}~\cite{Stravs:2013} to process
-metabolomics tandem MS files and build MassBank records,
+statistical analysis \textcolor{black}{of} quantitative proteomics data,
+\texttt{RMassBank}~\cite{Stravs:2013} to process metabolomics tandem
+MS files and build MassBank records,
 \texttt{MSstatsQC}~\cite{Dogu:2017} for longitudinal system

### 2020-08-29 msnbase2.Rnw
fix typo

@@
   data exceeds the available RAM and/or when several MS levels are to
-  be loaded and handled simultaneously. The on-disk backend can still
-  prove useful in cases when small MS2-only data are to be analysed
-  and will remain available in future versions of \texttt{MSnbase}.}
+  be loaded and handled simultaneously. The in-memory backend can
+  still prove useful in cases when small MS2-only data are to be
+  analysed, and will remain available in future versions of
+  \texttt{MSnbase}.}
 

### 2020-09-07 large_experiment_processing.tex
Add manually fixed tex for supplement

\documentclass[journal=jacsat,manuscript=suppinfo]{achemso}

\usepackage[]{graphicx}
\usepackage[]{color}

\usepackage[]{xcolor}
\usepackage[normalem]{ulem}
\usepackage[T1]{fontenc} % Use modern font encodings

\usepackage[version=3]{mhchem}
\usepackage{amsmath}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

\author{Laurent Gatto}
\email{laurent.gatto@uclouvain.be}
\affiliation[UCLouvain]{Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium}
\author{Sebastian Gibb}
\affiliation[University of Greifswald]{Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany}
\author{Johannes Rainer}
\affiliation[Eurac Research]{Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of L\"ubeck, Bolzano, Italy}

\title[MSnbase version 2]
  {\texttt{MSnbase}, efficient and elegant R-based processing and
    visualisation of raw mass spectrometry data}

\makeatletter
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Supplementary data for: },
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

% pandoc header

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\begin{document}

\maketitle
\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This document describes handling of mass spectrometry data from large
experiments using the \texttt{MSnbase} package and more specifically its
\emph{on-disk} backend. For demonstration purposes, the
\href{https://massive.ucsd.edu/ProteoSAFe/static/massive.jsp}{MassIVE}
data set
\href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=5e7034cc98c54a47b803b144bff6a296}{MSV000080030}
is used. This consists of over 1,000 mzXML files from swab-samples
collected from hands and various personal objects of 80 volunteers.

\hypertarget{data-handling-and-analysis-with-msnbase}{%
\section{\texorpdfstring{Data handling and analysis with
\texttt{MSnbase}}{Data handling and analysis with MSnbase}}\label{data-handling-and-analysis-with-msnbase}}

In this section we demonstrate data handling and access by
\texttt{MSnbase} on a large experiment consisting of more than 1,000
data files.

To reproduce the analysis described in this document, download the
\emph{MSV000080030} folder from
\url{ftp://massive.ucsd.edu/MSV000080030/} and place it into the same
folder as this document.

Below we load the required libraries and define the files to be
analyzed.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(MSnbase)}
\KeywordTok{library}\NormalTok{(magrittr)}
\KeywordTok{library}\NormalTok{(pryr)}

\NormalTok{fls \textless{}{-}}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"MSV000080030/ccms\_peak/Forensic\_study\_80\_volunteers/"}\NormalTok{,}
           \DataTypeTok{pattern =} \StringTok{"mzXML"}\NormalTok{, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{recursive =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The data set consists of 1182 mzXML files. We next load the data using
the two different \texttt{MSnbase} backends \texttt{"inMemory}" and
\texttt{"onDisk"}. For the in-memory backend, due to the larger memory
requirements, we import the data only from a subset of the files.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ms\_mem \textless{}{-}}\StringTok{ }\KeywordTok{readMSData}\NormalTok{(fls[}\KeywordTok{grep}\NormalTok{(}\StringTok{"Hand"}\NormalTok{, fls)], }\DataTypeTok{mode =} \StringTok{"inMemory"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Next we load data from all mzXML files as an on-disk \texttt{MSnExp}
object.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ms\_dsk \textless{}{-}}\StringTok{ }\KeywordTok{readMSData}\NormalTok{(fls, }\DataTypeTok{mode =} \StringTok{"onDisk"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Below we count the number of spectra per MS level of the whole
experiment.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{table}\NormalTok{(}\KeywordTok{msLevel}\NormalTok{(ms\_dsk))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
##       1       2 
## 1173678 4599786
\end{verbatim}

Note that the in-memory \texttt{MSnExp} object contains only MS2 spectra
(in total 2140520) from a subset of data files. However, the data import
was much slower (over \textasciitilde{} 12 hours for the in-memory
backend while creating the on-disk object from the full data data set
took \textasciitilde{} 3 hours).

Next we subset the on-disk object to contain the same set of spectra as
the in-memory \texttt{MSnExp} and compare their memory footprint.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ms\_dsk\_hands \textless{}{-}}\StringTok{ }\NormalTok{ms\_dsk }\OperatorTok{\%\textgreater{}\%}
\StringTok{    }\KeywordTok{filterFile}\NormalTok{(}\KeywordTok{grep}\NormalTok{(}\StringTok{"Hand"}\NormalTok{, fls)) }\OperatorTok{\%\textgreater{}\%}
\StringTok{    }\KeywordTok{filterMsLevel}\NormalTok{(2L)}

\KeywordTok{object\_size}\NormalTok{(ms\_mem)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 21.8 GB
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{object\_size}\NormalTok{(ms\_dsk\_hands)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 617 MB
\end{verbatim}

Since the on-disk object stores only spectra metadata in memory it
occupies also much less system memory. As a comparison, the on-disk
\texttt{MSnExp} for the full experiment was still much smaller than the
in-memory object:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{object\_size}\NormalTok{(ms\_dsk)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 1.66 GB
\end{verbatim}

\hypertarget{basic-ms-data-access-functionality}{%
\subsection{Basic MS data access
functionality}\label{basic-ms-data-access-functionality}}

Before evaluating the \texttt{MSnbase} performance on the large data set
we provide some general description of the \texttt{MSnbase} data classes
and basic data access operations. MS data from raw data files in mzML,
mzXML, mzData or netCDF format is represented by the \texttt{MSnExp}
object which organizes the spectra from the original files in an
one-dimensional list. Functions like \texttt{rtime} and \texttt{msLevel}
allow to extract the retention time and MS level, respectively. They
return a \texttt{numeric} (or \texttt{integer}) vector with the same
length as the number of spectra in the \texttt{MSnExp}. In the example
below we use the \texttt{rtime} function to extract the retention times
for each spectrum.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rts \textless{}{-}}\StringTok{ }\KeywordTok{rtime}\NormalTok{(ms\_dsk)}
\KeywordTok{length}\NormalTok{(rts)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 5773464
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(rts)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## F0001.S0001 F0001.S0002 F0001.S0003 F0001.S0004 F0001.S0005 F0001.S0006 
##       0.470       0.803       1.136       1.468       1.801       2.134
\end{verbatim}

The \texttt{fromFile} function can be used to determine the source file
(sample) of a specific spectrum in the \texttt{MSnExp} object. This
function returns an \texttt{integer} vector, of the same length as
spectra in the experiment, with the file index. The file names can be
accessed with the \texttt{fileNames} method. An \texttt{MSnExp} object
can be subsetted with \texttt{{[}} and e.g.~the index of the spectra
that should be retained. In the code block below we subset our
\texttt{ms\_dsk} object to keep only spectra from the 3rd file.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{one\_file \textless{}{-}}\StringTok{ }\NormalTok{ms\_dsk[}\KeywordTok{fromFile}\NormalTok{(ms\_dsk) }\OperatorTok{==}\StringTok{ }\DecValTok{3}\NormalTok{]}
\KeywordTok{length}\NormalTok{(one\_file)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 4911
\end{verbatim}

Note that there are also dedicated \emph{filter} functions to subset an
\texttt{MSnExp} object such as \texttt{filterFile},
\texttt{filterMsLevel}, \texttt{filterRt}, \texttt{filterMz},
\texttt{filterPrecursorMz} or \texttt{filterIsolationWindow}. In the
example below we use the \texttt{filterRt} function to further subset
our data to keep only spectra acquired within a certain time range.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{one\_file \textless{}{-}}\StringTok{ }\KeywordTok{filterRt}\NormalTok{(one\_file, }\DataTypeTok{rt =} \KeywordTok{c}\NormalTok{(}\DecValTok{40}\NormalTok{, }\DecValTok{300}\NormalTok{))}
\KeywordTok{length}\NormalTok{(one\_file)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 1996
\end{verbatim}

As mentioned above, the \texttt{MSnExp} object is comparable with a list
of spectra. Thus, to extract a single spectrum from it we can use
\texttt{{[}{[}}. This will return an object of type \texttt{Spectrum}
which encapsules/represents all information of the measured spectrum
(i.e.~m/z and intensity values as well as metadata information). In the
example below we extract the 15th spectrum from our data subset and
access its m/z values with the \texttt{mz} function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sp \textless{}{-}}\StringTok{ }\NormalTok{one\_file[[}\DecValTok{15}\NormalTok{]]}
\KeywordTok{mz}\NormalTok{(sp)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1]  400.4412  431.2400 1617.8282
\end{verbatim}

This particular spectrum has only 3 peaks.

Note that m/z or intensity values can also be directly extracted from
the \texttt{MSnExp} object as shown in the example below. The result
will be a \texttt{list} of \texttt{numeric} vectors, each element
representing the m/z values for each spectrum in the object.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mzs \textless{}{-}}\StringTok{ }\KeywordTok{mz}\NormalTok{(one\_file)}
\KeywordTok{class}\NormalTok{(mzs)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "list"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{length}\NormalTok{(mzs)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 1996
\end{verbatim}

In addition, it is also possible to extract all m/z and intensity values
from an \texttt{MSnExp} object as a \texttt{data.frame} as shown in the
code block below, but this is not suggested, since it loads all the data
into memory but all MS spectrum metadata such as MS level or precursor
m/z get lost.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df \textless{}{-}}\StringTok{ }\KeywordTok{as}\NormalTok{(one\_file, }\StringTok{"data.frame"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   file     rt       mz   i
## 1    1 40.118 387.2650  88
## 2    1 40.118 389.2627 192
## 3    1 40.118 474.2964 164
## 4    1 40.450 387.2416 212
## 5    1 40.450 389.2666 184
## 6    1 40.450 445.2680 132
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{nrow}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 2854657
\end{verbatim}

Note that for all these operations it is irrelevant whether an in-memory
or on-disk backend was used. In general it is advisable to use the
on-disk backend especially for experiments with more than
\textasciitilde{} 50 files.

\hypertarget{performance-of-the-on-disk-backend-on-large-scale-data-sets}{%
\subsection{\texorpdfstring{Performance of the \emph{on-disk} backend on
large scale data
sets}{Performance of the on-disk backend on large scale data sets}}\label{performance-of-the-on-disk-backend-on-large-scale-data-sets}}

To demonstrate \texttt{MSnbase}'s efficiency in processing large scale
experiments we perform some standard subsetting, data access and
manipulation operations.

We first compare the performance of the on-disk and in-memory backend on
accessing m/z values with the \texttt{mz} function on a set of 100
randomly selected spectra. The performance is assessed with the
\texttt{microbenchmark} function.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{idx \textless{}{-}}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\KeywordTok{seq\_along}\NormalTok{(ms\_mem), }\DecValTok{100}\NormalTok{)}

\KeywordTok{library}\NormalTok{(microbenchmark)}
\KeywordTok{microbenchmark}\NormalTok{(}\KeywordTok{mz}\NormalTok{(ms\_mem[idx]),}
               \KeywordTok{mz}\NormalTok{(ms\_dsk\_hands[idx]),}
               \DataTypeTok{times =} \DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Unit: seconds
##                   expr       min        lq     mean    median       uq      max
##        mz(ms_mem[idx]) 51.109825 52.054915 61.29039 63.480004 64.92025 74.88694
##  mz(ms_dsk_hands[idx])  3.638812  3.644038 13.97493  3.970938 28.53509 30.08579
##  neval
##      5
##      5
\end{verbatim}

For this combined subsetting and data access operation the on-disk
backend performed better than the in-memory \texttt{MSnExp}, while even
requiring much less memory.

Next we extract all MS2 spectra with a retention time between 50 and 60
seconds and a precursor m/z of 108.5362 (+/- 5ppm). This subsetting
operation is performed on the on-disk \texttt{MSnExp} object
representing the full experiment with the 1182 data files/samples. To
assess the performance of the following operations we use
\texttt{system.time} calls that record elapsed time in seconds.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{system.time}\NormalTok{(}
\NormalTok{    ms\_sub \textless{}{-}}\StringTok{ }\NormalTok{ms\_dsk }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{filterMsLevel}\NormalTok{(2L) }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{filterRt}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{50}\NormalTok{, }\DecValTok{60}\NormalTok{)) }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{filterPrecursorMz}\NormalTok{(}\DataTypeTok{mz =} \FloatTok{108.5362}\NormalTok{, }\DataTypeTok{ppm =} \DecValTok{5}\NormalTok{)}
\NormalTok{)[}\StringTok{"elapsed"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## elapsed 
##   6.529
\end{verbatim}

In total \texttt{length(ms\_sub)} spectra were selected from in total
928 data files/samples. The plot below shows the data for the first
spectrum.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{system.time}\NormalTok{(}
    \KeywordTok{plot}\NormalTok{(ms\_sub[[}\DecValTok{1}\NormalTok{]])}
\NormalTok{)[}\StringTok{"elapsed"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{large_experiment_processing_files/figure-latex/precursor-selection-plot-1-1.pdf}
\caption{Example spectrum of the data set.}
\end{figure}

\begin{verbatim}
## elapsed 
##   0.398
\end{verbatim}

Since there seems to be quite some background noise in the MS2 spectrum
we next remove peaks with an intensity below 50 by first replacing their
intensities with 0 (with the \texttt{removePeaks} call) and subsequently
removing all 0-intensity peaks from each spectrum with the
\texttt{clean} call. In addition we \emph{normalize} each spectrum by
dividing the maximum intensity per spectrum from the spectrum's
intensities.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{system.time}\NormalTok{(}
\NormalTok{    ms\_sub \textless{}{-}}\StringTok{ }\NormalTok{ms\_sub }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{removePeaks}\NormalTok{(}\DataTypeTok{t =} \DecValTok{50}\NormalTok{) }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{clean}\NormalTok{(}\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{\%\textgreater{}\%}
\StringTok{        }\KeywordTok{normalize}\NormalTok{(}\DataTypeTok{method =} \StringTok{"max"}\NormalTok{)}
\NormalTok{)[}\StringTok{"elapsed"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## elapsed 
##   0.006
\end{verbatim}

The result on the first spectrum is shown below.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{system.time}\NormalTok{(}
    \KeywordTok{plot}\NormalTok{(ms\_sub[[}\DecValTok{1}\NormalTok{]])}
\NormalTok{)[}\StringTok{"elapsed"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{figure}[h!]
\centering
\includegraphics{large_experiment_processing_files/figure-latex/precursor-selection-plot-normalized-1.pdf}
\caption{Example spectrum after cleaning.}
\end{figure}

\begin{verbatim}
## elapsed 
##   0.547
\end{verbatim}

\pagebreak

Note that any of the data manipulations above are not directly applied
to the data but \emph{cached} in the object's internal \emph{lazy
processing queue} (explaining the very short running time of the
normalization call). The operations are only effectively applied to the
data when m/z or intensity values are extracted from the object, e.g.~in
the \texttt{plot} call above.

For additional workflows employing \texttt{MSnbase} see also
\href{https://jorainer.github.io/metabolomics2018/xcms-preprocessing.html}{metabolomics2018}\footnote{https://github.com/jorainer/metabolomics2018}
that explains filtering, plotting and centroiding of profile-mode MS
data with \texttt{MSnbase} and subsequent pre-processing of the (label
free/untargeted) LC-MS data with the \texttt{xcms} package (that builds
upon \texttt{MSnbase} for MS data representation and access).

\hypertarget{system-information}{%
\subsection{System information}\label{system-information}}

The present analysis was run on a MacBook Pro 16,1 with 2.3 GHz 8-Core
Intel Core i9 CPU and 64 GB 2667 MHz DDR4 memory running macOS version
10.15.5. The R version and the version of the used packages are listed
below.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{sessionInfo}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## R version 4.0.2 (2020-06-22)
## Platform: x86_64-apple-darwin17.0 (64-bit)
## Running under: macOS Catalina 10.15.6
## 
## Matrix products: default
## BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
## LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
## 
## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
## 
## attached base packages:
## [1] stats4    parallel  stats     graphics  grDevices utils     datasets 
## [8] methods   base     
## 
## other attached packages:
##  [1] microbenchmark_1.4-7 BiocParallel_1.22.0  pryr_0.1.4          
##  [4] magrittr_1.5         MSnbase_2.14.2       ProtGenerics_1.20.0 
##  [7] S4Vectors_0.26.1     mzR_2.22.0           Rcpp_1.0.5          
## [10] Biobase_2.48.0       BiocGenerics_0.34.0  BiocStyle_2.16.0    
## [13] rmarkdown_2.3       
## 
## loaded via a namespace (and not attached):
##  [1] lattice_0.20-41       digest_0.6.25         foreach_1.5.0        
##  [4] R6_2.4.1              plyr_1.8.6            mzID_1.26.0          
##  [7] evaluate_0.14         ggplot2_3.3.2         highr_0.8            
## [10] pillar_1.4.6          zlibbioc_1.34.0       rlang_0.4.7          
## [13] rticles_0.15          preprocessCore_1.50.0 labeling_0.3         
## [16] stringr_1.4.0         munsell_0.5.0         tinytex_0.25         
## [19] compiler_4.0.2        xfun_0.16             pkgconfig_2.0.3      
## [22] pcaMethods_1.80.0     htmltools_0.5.0       tidyselect_1.1.0     
## [25] tibble_3.0.3          bookdown_0.20         IRanges_2.22.2       
## [28] codetools_0.2-16      XML_3.99-0.5          crayon_1.3.4         
## [31] dplyr_1.0.2           MASS_7.3-52           grid_4.0.2           
## [34] gtable_0.3.0          lifecycle_0.2.0       affy_1.66.0          
## [37] scales_1.1.1          ncdf4_1.17            stringi_1.4.6        
## [40] impute_1.62.0         farver_2.0.3          affyio_1.58.0        
## [43] doParallel_1.0.15     limma_3.44.3          ellipsis_0.3.1       
## [46] generics_0.0.2        vctrs_0.3.4           iterators_1.0.12     
## [49] tools_4.0.2           glue_1.4.2            purrr_0.3.4          
## [52] yaml_2.2.1            colorspace_1.4-1      BiocManager_1.30.10  
## [55] vsn_3.56.0            MALDIquant_1.19.3     knitr_1.29
\end{verbatim}
\end{document}

### 2020-09-07 large_experiment_processing.Rmd
Update Rmd

@@
 ---
-title: "Supplementary data for: \\textit{`MSnbase`, efficient and elegant R-based processing and visualisation of raw mass spectrometry data}"
-author: "Laurent Gatto, Sebastian Gibb and Johannes Rainer"
+title: "MSnbase, efficient and elegant R-based processing and visualisation of raw mass spectrometry data"
+author:
+  - name: Laurent Gatto
+    affiliation: Computational Biology Unit, de Duve Institute, Universit\'e catholique de Louvain, Brussels, Belgium
+  - name: Sebastian Gibb
+    affiliation: Department of Anaesthesiology and Intensive Care of the University Medicine Greifswald, Germany
+  - name: Johannes Rainer
+    affiliation: Institute for Biomedicine, Eurac Research, Affiliated Institueof the University of L\"ubeck, Bolzano, Italy
 output: 
-    BiocStyle::pdf_document:
-        includes:
-            in_header: "header.sty"
+  rticles::acs_article:
 ---
